---
title: "Coder Upgrade 2025: Cell Cell interaction analysis"
author: "Matt Cannon and Xin Wang"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 5
        number_sections: false
        code-fold: true
        code-link: true
        df-print: kable
        embed-resources: true
        self_contained: true
knitr:
    opts_chunk:
        cache.lazy: false
        warning: false
        message: false
---

Load libraries
```{r libraries, cache=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)
library(Seurat)
library(CellChat)
library(qs2)
```

--------------------------------------------------------------------------------

# CellChat Overview

https://www.nature.com/articles/s41467-021-21246-9

Essentially, CellChat quantifies the expression of ligands and receptors in a cell type, and then uses that to infer which cell types are likely to be communicating with each other. It uses a hand-curated database of ligand-receptor interactions and signaling pathways. It quantifies the number of interactions and the strength of those interactions and has nice tools for visualizing the results.

https://htmlpreview.github.io/?https://github.com/jinworks/CellChat/blob/master/tutorial/CellChat-vignette.html

## Load in the data

```{r}
kidney_data <-
  qs_read("~/workshop_data/rdata/merged_sobj_harmony_sample_batch_theta.qs2") %>%
  subset(batch == "GSE252772") %>%
  process_seurat()

kidney_data$age <-
  kidney_data$orig.ident %>%
  str_extract("NMK(.)[MF]$", group = TRUE) %>%
  str_c(" wk")

DimPlot(
  kidney_data,
  group.by = "predicted.id",
  label = TRUE,
  label.box = TRUE,
  label.size = 2.5,
  repel = TRUE
) +
  NoLegend()

kidney_data$cell_sample <-
    paste(
        kidney_data$anchor_cell_types,
        kidney_data$orig.ident,
        sep = "_"
    )

# CellChat requires your metadata to have this column
kidney_data$samples <- as.factor(kidney_data$orig.ident)
```

## Filter down to just prominent cell types
```{r}
min_cells <- 50

cell_type_counts <-
    table(kidney_data$anchor_cell_types) %>%
    as.list()

keep_cell_types <-
    cell_type_counts[cell_type_counts > min_cells] %>%
    names()

kidney_data <-
    subset(kidney_data, cell_type %in% keep_cell_types)

Idents(kidney_data) <-
    kidney_data$cell_sample
```

## Downsample to reduce memory usage and speed things up
This keeps 50 cells per cell type per sample because our Idents() are cell_sample
```{r}
kidney_data <-
    subset(kidney_data, downsample = 50)
gc() # reduce ram usage from old object
```

## Make CellChat object
```{r}
cellchat_obj <-
    createCellChat(
        object = kidney_data,
        group.by = "anchor_cell_types",
        assay = "RNA"
    )

# Calculate the number of cells in each group (cell_type)
groupSize <- as.numeric(table(cellchat_obj@idents))
groupSize
```

## Load in ligand-receptor interactions
We keep only interactions within the "Secreted Signaling" group
```{r}
cellchat_db <- CellChatDB.mouse

showDatabaseCategory(cellchat_db)

cellchat_obj@DB <-
    subsetDB(
        cellchat_db,
        search = "Secreted Signaling",
        key = "annotation"
    )
```

## Preprocess the data
subsetData() subsets the data down to just genes in the database. This adds data to the data.signaling slot.

identifyOverExpressedGenes() identifies genes that are overexpressed in each cell type. This is used to identify which cell types are likely to be communicating with each other. This adds data into the var.features slot which is a list.

identifyOverExpressedInteractions() identifies interactions that are overexpressed in each cell type. This adds data into the LRsig slot as a data frame.
```{r}
cellchat_obj <-
    subsetData(cellchat_obj) %>%
    identifyOverExpressedGenes() %>%
    identifyOverExpressedInteractions()
```

## Compute the communication network
Populates the cellchat_obj@net slot with interaction probability and pvalues.

This takes about two minutes
```{r}
cellchat_obj <- computeCommunProb(object = cellchat_obj, type = "triMean")
```

## Lets look at the interaction probability

This is a 3d array, where the third dimension is each ligand-receptor pair
```{r}
cellchat_obj@net$prob[, , 1:2]

dimnames(cellchat_obj@net$prob)[[3]] %>%
    head(n = 20)

third_d_names <- dimnames(cellchat_obj@net$prob)[[3]]
third_d_names

matts_cell_type <- "kidney proximal convoluted tubule epithelial cell"

apply(
  cellchat_obj@net$prob,
  MARGIN = 3,
  function(x) {
    x %>%
      as.data.frame() %>%
      .[matts_cell_type] %>%
      max()
  }
) %>%
  sort() %>%
  tail()

cellchat_obj@net$prob[, , "EGF_EGFR"] %>%
    pheatmap::pheatmap(
        display_numbers = TRUE,
        main = "EGF_EGFR"
    )
```

## Infer at signaling pathway level
Populates the cellchat_obj@netP slot with interaction probability and pvalues, operating at the pathway level.

`prob` is the calculated probability and is a 3d array. The first two dimensions are the cell types and the third dimension is the pathway. The values in the array are the probabilities of interaction between the cell types for each pathway.

```{r}
cellchat_obj <- computeCommunProbPathway(cellchat_obj)

cellchat_obj@netP$pathways

third_d_names <- dimnames(cellchat_obj@netP$prob)[[3]]

apply(
  cellchat_obj@netP$prob,
  MARGIN = 3,
  function(x) {
    x %>%
      as.data.frame() %>%
      .[matts_cell_type] %>%
      max()
  }
) %>%
  sort() %>%
  tail()

cellchat_obj@netP$prob[, , "EGF"]

# Look at "MK" pathway
cellchat_obj@netP$prob[, , "EGF"] %>%
    pheatmap::pheatmap(display_numbers = TRUE)
```

## Calculate aggregated communication network
Populates "weight" and "count" inside the cellchat_obj@net slot. This is the aggregated communication network. The weight is the sum of the probabilities of interaction between the cell types for each ligand-receptor interaction The count is the number of interactions between the cell types for each pathway.
```{r}
cellchat_obj <- aggregateNet(cellchat_obj)

cellchat_obj@net$weight

cellchat_obj@net$weight %>%
    pheatmap::pheatmap(
        display_numbers = TRUE,
        main = "Weight of interactions"
    )

cellchat_obj@net$count %>%
    pheatmap::pheatmap(
        display_numbers = TRUE,
        main = "Number of interactions",
        number_format = "%.0f",
        fontsize_number = 15
    )
```

## Pull results out of the object
This gets the data in a more friendly format (long format)
```{r}
cc_results <-
  subsetCommunication(cellchat_obj) %>%
  as_tibble()
cc_results

cc_results_pathway <-
  subsetCommunication(cellchat_obj, slot.name = "netP") %>%
  as_tibble()
cc_results_pathway

cc_results_pathway %>%
  filter(prob >= 0.1) %>%
  filter(target == matts_cell_type) %>%
  arrange(desc(prob))
```

## Visualize the results
CellChat makes very pretty plots
```{r}
png(
    filename = "output/ccc/figures/cc_num_interactions_plot.png",
    width = 3000,
    height = 3000,
    res = 300,
    type = "cairo"
)
netVisual_circle(
    cellchat_obj@net$count,
    vertex.weight = groupSize,
    weight.scale = TRUE,
    label.edge = FALSE,
    title.name = "Number of interactions"
)
dev.off()

png(
    filename = "output/ccc/figures/cc_weights_plot.png",
    width = 3000,
    height = 3000,
    res = 300,
    type = "cairo"
)
netVisual_circle(
    cellchat_obj@net$weight,
    vertex.weight = groupSize,
    weight.scale = TRUE,
    label.edge = FALSE,
    title.name = "Number of interactions"
)
dev.off()
```

## Look at the composition of pathways
```{r}
cellchat_db$interaction %>%
            filter(pathway_name == "EGF") %>%
    head()

cellchat_db$interaction %>%
    filter(pathway_name == "EGF") %>%
    select(ligand.symbol, receptor.symbol)
```
