---
title: "Coder Upgrade 2023: Seurat Under the Hood"
author: "Matt Cannon"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 5
        number_sections: false
        code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      cache.lazy = FALSE)
```

Load libraries
```{r libraries, cache=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(rrrSingleCellUtils)
library(tidyverse)
library(Seurat)
```

--------------------------------------------------------------------------------

# Section 1: Objects in R and Seurat objects

R has many variable types
```{r}
my_vector <- c(1, 2, 3, 4, 5)
my_matrix <- matrix(1:9, nrow = 3)
my_data_frame <- data.frame(x = 1:5, y = 6:10)
my_list <- list(a = 1, b = 2, c = "stuff")
```

R is mostly a "functional language"

But it also has some "object oriented" features

- Objects are a way to store data and functions together
- Objects are intended to be "black boxes" that you can interact with
- Methods or functions are used to interact with objects
- Conceptually different than normal variables
    - Variables are meant to be directly manipulated
    - Objects are meant to be interacted with through methods

Three main types of objects in R

- S3
    - The original object type in R
    - Less strictly defined
- S4
    - More strictly defined
    - More complicated to create
    - Less prone to errors
- R6
    - More traditional OO behavior
    - Not super common

## Seurat objects are S4 objects
```{r}
isS4(pbmc_small)
class(pbmc_small)
```

This is the code that defines a Seurat object
```{r}
Seurat <- setClass(
  Class = 'Seurat',
  slots = c(
    assays = 'list',
    meta.data = 'data.frame',
    active.assay = 'character',
    active.ident = 'factor',
    graphs = 'list',
    neighbors = 'list',
    reductions = 'list',
    images = 'list',
    project.name = 'character',
    misc = 'list',
    version = 'package_version',
    commands = 'list',
    tools = 'list'
  )
)
```

Then methods/functions are created to work with Seurat objects

Note that the function has a function name (merge), then ".", then the object type (Seurat)

This tells R how to use the merge function on a Seurat object
```{r}
merge.Seurat <- function(
  x = NULL,
  y = NULL,
  add.cell.ids = NULL,
  merge.data = TRUE,
  merge.dr = NULL,
  project = "SeuratProject",
  ...
) {
  CheckDots(...)
  objects <- c(x, y)
  if (!is.null(x = add.cell.ids)) {
    if (length(x = add.cell.ids) != length(x = objects)) {
      stop("Please provide a cell identifier for each object provided to merge")
    }
    for (i in 1:length(x = objects)) {
      objects[[i]] <- RenameCells(object = objects[[i]], add.cell.id = add.cell.ids[i])
    }
  }
  # ensure unique cell names
  objects <- CheckDuplicateCellNames(object.list = objects)
  assays <- lapply(
    X = objects,
    FUN = FilterObjects,
    classes.keep = 'Assay'
  )
#
#more stuff down here, but you get the idea
#
}
```

## Why do it this way?

- The structure of an object is defined and static
- Can write functions that are "guaranteed" to work with objects of a certain type
- Can extend base R functions to work with objects of a certain type

```{r}
str(pbmc_small)
```

> Show ppt slide with Seurat object structure

## Seurat object slots:

There are only a handful of slots in a Seurat object

But, each slot can contain objects, which then have their own slots

```{r}
str(pbmc_small)
```

Ends up looking really complex if you don't break it down by level

- assays = 'list'
    - list of assay objects containing the gene expression data
- meta.data = 'data.frame'
    - Metadata for each cell
- active.assay = 'character'
    - Which assay is currently active
- active.ident = 'factor'
    - Vector of default "idents" for each cell
- graphs = 'list'
    - Filled in by `FindNeighbors()`
- neighbors = 'list'
    - Filled in by `FindNeighbors()`
- reductions = 'list'
    - Dimensionality reduction results
- images = 'list'
    - Not really applicable to our type of data
- project.name = 'character'
    - Name of the project
- misc = 'list'
- version = 'package_version'
- commands = 'list'
    - **List of commands used to create the object**
- tools = 'list'

--------------------------------------------------------------------------------

# Activity 1:
Explore the pbmc_small Seurat object

Type "pbmc_small@" in the console/terminal, then hit tab a few times to see the available slots
```{r}

```

Type "pbmc_small$" in the console/terminal, then hit tab a few times to see the metadata column names
```{r}

```

What commands were used to create the object?
What resolution was used in `FindClusters()`?
```{r}

```

What are the active assay and active identity of this object?
```{r}

```

Pull the metadata out of the object and save it as a new variable
```{r}

```

What information is encoded by the rownames of the metadata?

--------------------------------------------------------------------------------

# Section 2: Assay object structure

The "@assay" slot of a Seurat object contains the actual gene expression data

Each element of the list is an assay object

> Show ppt of Assay object structure

## Assay object slots:

- counts = 'matrix' - usually a sparse matrix (dgCMatrix)
    - Raw counts
- data = 'matrix' - usually a sparse matrix (dgCMatrix)
    - Normalized counts
- scale.data = 'matrix'
    - Scaled counts
- key = 'character'
    - Not used externally
- assay.orig = 'character'
    - Label for the assay
- var.features = 'vector'
    - Vector of variable genes
- meta.features = 'data.frame'
- misc = 'list'

```{r}
class(pbmc_small@assays[[1]])
```

## Navigating a Seurat object

Key concepts:

-  Start with object name
-  Use "@" to access the slot you want
-  If the slot contains a list, reference it by name
-  If you are referencing an internal object, use "@" to access the slot you want
-  Repeat all the way down to the bit you need

> Show ppt of navigating Seurat object

```{verbatim}
obj            slot            list element     slot of "DimReduc" object
pbmc_small     @reductions     $pca             @cell.embeddings
```

```{r}
pca_stuff <- pbmc_small@reductions$pca@cell.embeddings
pca_stuff[1:10, 1:3]
```

## Fun with @commands

This list is structured to make pulling out specific commands easier

-  Each element of the list is named by the specific command run
-  This in turn is a named list of arguments

```{r}
pbmc_small@commands$FindVariableFeatures.RNA$nfeatures
```

```{verbatim}
obj          slot       command                     argument
pbmc_small   @commands  $FindVariableFeatures.RNA   $nfeatures
```

## Other random manual bits
Getting cell names and gene names

```{r}
genes <- rownames(pbmc_small)
head(genes)

cells <- colnames(pbmc_small)
head(cells)
```

Subsetting a Seurat object by row/column numbers

            genes        cells
object_name[row_numbers, column_numbers]
```{r}
subset_data <- pbmc_small[1:50, 1:20]
subset_data
```

--------------------------------------------------------------------------------

# Activity 2:

Pull the "RNA" assay out of the Seurat object and save it as a new variable named `rna_assay`
```{r}

```

Pull the raw counts out of `rna_assay`, save it as a new variable named "rna_counts" and look at the first 6 rows
```{r}

```

Make a vector that contains the names of the cells in the object
```{r}

```

What is the length of this vector? How many cells are in the object?
```{r}

```

Use the square bracket notation to randomly sub-sample your seurat object down to 50 cells and 100 genes
```{r}

```

--------------------------------------------------------------------------------

# Section 3: Interacting with Seurat objects
## The "right" way
Use Seurat functions to interact with the object instead of diving into the object

Key functions
https://satijalab.org/seurat/reference/index.html

## Set active Idents

The active idents is essentially the default metadata your cells are grouped by

Can specify either a metadata column name, or provide your own vector of values
```{r}
Idents(pbmc_small) <- "letter.idents"
table(Idents(pbmc_small))

Idents(pbmc_small) <- "this_is_not_a_real_column_in_the_data"
table(Idents(pbmc_small))

Idents(pbmc_small) <- pbmc_small$letter.idents
table(Idents(pbmc_small))

DimPlot(pbmc_small)
```

## Get list of cells
```{r}
cells <- Cells(pbmc_small)
head(cells)
```

## Get expression data

- counts - raw counts
- data - normalized counts  *<- you most likely want this*
- scale.data - scaled counts

Returns sparse matrix
Can convert to a matrix with `as.matrix()`
Rownames are the gene names
```{r}
my_data <-
    GetAssayData(pbmc_small,
                 assay = "RNA",
                 slot = "counts")

GetAssayData(pbmc_small, slot = "counts")[1:5, 1:5] %>%
    as.matrix()

GetAssayData(pbmc_small, slot = "data")[1:5, 1:5] %>%
    as.matrix()

GetAssayData(pbmc_small, slot = "scale.data")[1:5, 1:5] %>%
    as.matrix()
```

## Drop new data into an object
```{r}
raw_counts <-
    GetAssayData(pbmc_small, slot = "counts") %>%
    as.matrix()

bad_normalization <- raw_counts / colSums(raw_counts)
pbmc_small_borked <-
    SetAssayData(pbmc_small,
                 slot = "data",
                 new.data = bad_normalization)
```

## Subset a Seurat object
Can subset by cells, idents, features, gene expression, etc...
```{r}
# Note that there is only one equal sign here
just_a <- subset(pbmc_small, idents = "A")

# But two equal signs here
just_a <- subset(pbmc_small, letter.idents == "A")

a_cells <- WhichCells(pbmc_small, idents = "A")

just_a <- subset(pbmc_small, cells = a_cells)

high_cd19 <- subset(pbmc_small, subset = CD19 > 1)
```

## Calculate average expression per group
```{r}
ave_expr <- AverageExpression(pbmc_small,
                              assays = "RNA",
                              group.by = "letter.idents")$RNA
ave_expr[1:5, ]
```

## Add metadata
If adding a dataframe of metadata, need rownames of cell names
When adding with "$", or passing a vector to `AddMetaData`, it doesn't check if cells match
```{r}
pbmc_small$nonsense <-
    sample(LETTERS[1:5],
           ncol(pbmc_small),
           replace = TRUE)

metadata_table <-
    data.frame(totally_true = sample(c("alive", "dead", "undead"),
                                     ncol(pbmc_small),
                                     replace = TRUE),
               also_useful = sample(1:1000000,
                                    ncol(pbmc_small),
                                    replace = FALSE),
               row.names = colnames(pbmc_small))

pbmc_small_plus <- AddMetaData(pbmc_small, metadata_table)
head(pbmc_small_plus@meta.data)

pbmc_small_plus <- AddMetaData(pbmc_small_plus,
                               colnames(pbmc_small_plus),
                               col.name = "cell_ids")
head(pbmc_small_plus@meta.data)
```

## Get PCA/UMAP/TSNE values
```{r}
tsne_data <- Embeddings(pbmc_small, reduction = "tsne")

pca_data <- Embeddings(pbmc_small, reduction = "pca")
```

## Get cell names for specific cells
```{r}
high_cd79b <- WhichCells(pbmc_small, expression = CD79B > 2)
head(high_cd79b)

group_b <- WhichCells(pbmc_small, idents = "B")
length(group_b)
```

## Get data for specific cells
```{r}
high_cd79b_data <-
    FetchData(pbmc_small,
              vars = rownames(pbmc_small),
              cells = high_cd79b)

high_cd79b_data[1:5, 1:5]

high_cd79b_data <-
    FetchData(pbmc_small,
              vars = c("groups", "letter.idents", "nonsense"),
              cells = high_cd79b)

head(high_cd79b_data)
```

## Save your Idents
```{r}
pbmc_small <- StashIdent(pbmc_small, "my_idents")
head(pbmc_small$my_idents)
```

# Wrapup

A Seurat object in an S4 object

Functions similar to a list of lists, except that the slots are accessed with "@" instead of "$"

The authors of the software wrote a lot of functions to interact with the object, so you don't have to dive into the object manually

But, if you need to, you can!

This should make a lot more sense now!
```{r}
str(pbmc_small)
```

--------------------------------------------------------------------------------

# Activity 3:

Lets redo some of what we did in activity 2, but lets do it the "right" way


Use the `GetAssayData()` method to pull the raw counts out of `rna_assay`, save it as a new variable named "rna_counts" and look at the first 6 rows. Why are there periods? How can you turn the periods into zeros?
```{r}

```

Use a Seurat function to make a vector that contains the names of the cells in the object
```{r}

```

What is the length of this vector? How many cells are in the object?
```{r}

```

Use the `subset()` function to sub-sample the pbmc_small object down to those cells where letter.idents == "A"
```{r}

```

Use the `subset()` function to sub-sample your seurat object down to the 50 cells with the highest expression of the gene "SAT1"
```{r}

```

--------------------------------------------------------------------------------

# Session challenge

Strangely, Seurat has no built in `summary` function, but you can write your own!
Write a function named summary.seurat() that takes a Seurat object as an argument and prints out the following information:

- Number of cells
- Number of features
- Current active idents
- Current default assay
- Print a DimPlot() of the object if the @reductions slot has something in it

```{r}

```

--------------------------------------------------------------------------------

# Helpful references

https://github.com/mojaveazure/seurat-object
https://github.com/satijalab/seurat
https://satijalab.org/seurat/reference/index.html
https://satijalab.org/seurat/index.html


Session info
```{r}
sessionInfo()
```