# Needed packages

# Beginner sessions

python?
scvi-tools

## Cran dependencies:
```{r}
cran_dependencies <-
    c("BiocManager",
      "pak")

install.packages(cran_dependencies, Ncpus = 10)
```

## Biocondutor dependencies:
```{r}
bioc_dependencies <-
    c("GenomeInfoDb",
      "GenomicRanges",
      "IRanges",
      "Rsamtools",
      "S4Vectors",
      "BiocGenerics")
```

## Cran packages:
```{r}
needed_packages_cran <- c(
    "data.table",
    "DESeq2",
    "fgsea",
    "futures",
    "ggrepel",
    "ggthemes",
    "harmony",
    "knitr",
    "MAST",
    "msigdbr",
    "parallel",
    "patchwork",
    "pheatmap",
    "RColorBrewer",
    "RCpp",
    "reticulate",
    "rmarkdown",
    "Seurat",
    "SeuratData",
    "SeuratObject",
    "SeuratWrappers",
    "Signac",
    "testthat",
    "tidyverse"
)
```

## Packages from GitHub
```{r}
github_packages <-
    c("saeyslab/nichenetr",
      "kidcancerlab/rrrSingleCellUtils",
      "NMikolajewicz/scMiko",
      "satijalab/seurat-wrappers",
      "satijalab/azimuth")
```

## Install all packages
```{r install_beginner}
pak::pkg_install(c(bioc_dependencies,
                   needed_packages_cran,
                   github_packages))
```



# Advanced sessions

For testing purposes only:
export R_LIBS_SITE="testLibDir/"
.libPaths(.libPaths()[2])

## Unix utilities
On the Franklin cluster, we require
    HDF5
    python (version likely doesn't matter)
    R4.3.3
    Rstudio

We also need to load

`ml load CMake/3.16.4 XZ/5.2.5 NLopt/2.6.1`

in order for the packages to install without compilation errors.

I doubt the specific versions of cmake, etc. are important, those are just the ones I used.

reticulate::install_python(version = "3.11")

## Cran dependencies:
```{r}
cran_dependencies <-
    c("BiocManager",
      "pak")

install.packages(cran_dependencies, Ncpus = 2)
```

```{r}
bioc_dependencies <-
    c("GenomeInfoDb",
      "GenomicRanges",
      "IRanges",
      "Rsamtools",
      "S4Vectors",
      "BiocGenerics")

cran_packages <-
    c("KernSmooth",
      "pheatmap",
      "tidyverse",
      "qs",
      "RColorBrewer",
      "reticulate",
      "SeuratObject",
      "Signac",
      "Seurat")

bioc_packages <-
    c("AUCell",
      "biovizBase",
      "ComplexHeatmap",
      "doMC",
      "doRNG",
      "DT",
      "EnsDb.Mmusculus.v79",
      "GENIE3",
      "NMF",
      "mixtools",
      "R2HTML",
      "Rtsne",
      "zoo")

github_packages <-
    c("thomasp85/patchwork",
      "bokeh/rbokeh",
      "aertslab/RcisTarget",
      "drieslab/Giotto",
      "saeyslab/nichenetr",
      "kidcancerlab/rrrSingleCellUtils",
      "aertslab/SCopeLoomR",
      "aertslab/SCENIC",
      "satijalab/seurat-data",
      "mojaveazure/seurat-disk",
      "bnprks/BPCells",
      "satijalab/azimuth")

pak::pkg_install(c(bioc_dependencies,
                   cran_packages,
                   bioc_packages,
                   github_packages))
```

