---
title: 'Seurat Lesson: Code'
author: "Yogesh Budhathoki and Matt Gust"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  echo = TRUE,
  cache = TRUE,
  collapse = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE)
```

```{r lib, cache = FALSE}
# Load necessary packages
library(Seurat)
library(tidyverse) # This package contains ggplot2 and dplyr
library(patchwork)
library(ggplot2) # For plotting
library(rrrSingleCellUtils) # For plotting an processing
library(qs2) # For reading and writing .qs files
library(SingleR) # For cell type annotation
library(biomaRt) # For gene annotation
# Set the random generator seed so that results are reproducible.
set.seed(132)
```

# Load in seurtat object: 
***YOU MAY NEED TO MODIFY THE CODE BELOW SO THAT THE FOLDER PATH POINTS R TO 
WHEREVER THE RAW DATA IS LOCATED ON YOUR COMPUTER. Make sure that you include the 
necessary folders in the path to the below folders. 
Be careful with forward slashes. Folders always have a forward slash in front of 
them, and files do not.

# Perform Subsetting for Quality Control
All subsetting will depend on your sample and its quality and conditions. For 
instance, if one sample is a tumor treated with effective chemotherapy, you may 
expect to see a higher percent of mitochondrial genes and may want to include 
more of these dying cells in your subsetting (say, 25% mitochondrial genes). 
All subsetting should depend on the spread of your data, with outliers (both 
high and low) typically targeted for exclusion.
**Feature subsetting:**
nFeature_RNA: number of genes (features) detected per cell. 
Subsetting the number of features per cells can help to remove doublets (two cells
that grouped together as "one" -> high nFeature_RNA) and non-cells, such as 
dead cells or free mRNA that grouped together (low nFeature_RNA). 
nCount_RNA is also related to nFeature_RNA, so it is not necessary to use both
methods.

**nCount subsetting:**
nCount_RNA: number of unique molecule identifiers (UMIs) per cell
High nCount_RNA also indicates doublets, and so is another method for 
filtering out those doublets. Some prefer using this method to using nFeatures to remove 
doublets. Using both, one can be fairly confident of removing doublets. 

**Mitochondrial gene subsetting:**
Subsetting the data by percent mitochondrial genes helps remove poor quality 
cells, such as dead cells or cells with a broken membrane (this causes leakage of
mRNA). These cells have a higher percentage of mitochondrial genes. Typically, 
25% and lower is preferred; however, again, this depends on your data spread 
and your expectations for the sample. 
Note: "^MT-" = all genes beginning with "MT-", which is how all mitochondrial 
gene names begin. "^mt-" for mice gene pattern.

```{r load}
# read in the object
sobj <-
    qs_read("~/workshop_data/rdata/f1_sobj.qs2")

# check the object
sobj
colnames(sobj@meta.data)

# check the metadata
sobj[["percent.mt"]] <- 
    PercentageFeatureSet(sobj, pattern = "^mt-")

#check the metadata
colnames(sobj@meta.data)

# visualize the composition or distribution
# Violin plot to visualize outliers, each dot is a cell
VlnPlot(sobj,
        features = c("nFeature_RNA",
                     "nCount_RNA",
                     "percent.mt"))

# visualize the composition or distribution of ncount (nfeature is related to ncount)
hist(sobj$nCount_RNA,
     breaks = 100)

# visualize the distribution of the percent of mitochondrial genes
hist(sobj$percent.mt,
     breaks = 100)

# subset the dataset to remove the low quality cells
sobj <- 
    subset(sobj,
           subset = nCount_RNA < 25000 &
                    percent.mt < 20)

# check the substetting
hist(sobj$nCount_RNA,
     breaks = 100)
hist(sobj$percent.mt,
        breaks = 100)

# Violin plot to visualize that the subsetting has worked
VlnPlot(sobj,
        features = c("nFeature_RNA",
                     "nCount_RNA",
                     "percent.mt"))

# showing that the nCount and nFeature are related
FeatureScatter(sobj,
                feature1 = "nCount_RNA",
                feature2 = "nFeature_RNA") + 
theme(legend.position="none")

```


# activity1 
**Ribosomal and hemoglobin gene subsetting:**
These are other options for identifying low quality samples based on the percent
of these genes in each cell. Ribosomal genes begin with either "RPS" or "RPL". 
Hemoglobin genes begin with either "HBA" or "HBB".

make a new column in the metadata for the percent of ribosomal genes and hemoglobin genes. 
Then visualize the distribution of these genes using a violin plot and histogram.

Tips: the pattern for the ribosomal genes is "^Rp[sl]", which means that the gene name 
begins with "Rp" and is followed by either "s" or "l". The pattern for the hemoglobin genes
is "^Hb[ab]", which means that the gene name begins with "Hb" and is followed by either "a" or "b".

```{r activity1}

```

# Normalize 
Normalization is the process of adjusting gene expression values to eliminate/reduce 
the influence of technical effects impacting all genes the same way.

# Identify Highly Variable Features
This function identifies the n most highly variable features. By default, n = 2000 
(defined by nfeatures=2000). This is used by the next step, scaling, to speed it
up. Often, many biologically important features will be captured in this step, 
however, you may lose important data, depending on your dataset. 

# Scale Features
Scaling applies linear transformation. Importantly, this step ensures that 
highly-expressed genes do not dominate and drown out relatively lower-expressed 
genes that have biological importance. This is a standard pre-processing step 
prior to principle component analysis (dimensional reduction).
  -Shifts the expression of each gene (feature), so that the mean expression across cells is 0
  -Scales the expression of each gene (feature), so that the variance across cells is 1
  -Results are stored in covid[["RNA"]]@scale.data
  
Note that if the features parameter is undefined, this function will perform 
scaling on only 2,000 of the most variable features by default (may lose data).

```{r normalize_and_scale_data}
# normalize the data
sobj <- 
    NormalizeData(sobj,
                  normalization.method = "LogNormalize",
                  scale.factor = 10000)
# check the normalization, a new layer is added called data, you can compare counts and data layers
sobj
sobj@assays$RNA$counts[1:10,1:10]
sobj@assays$RNA$data[1:10,1:10]

# find the variable features
sobj <- 
    FindVariableFeatures(sobj,
                         selection.method = "vst",
                         nfeatures = 2000)

# visualize the variable features and layers
sobj@assays$RNA

# visualize the variable features
top10 <- 
    head(VariableFeatures(sobj),
         10)
top10

Seurat::LabelPoints(plot = VariableFeaturePlot(sobj),
                    points = top10,
                    repel = TRUE)

# scale the object
sobj <- 
    ScaleData(sobj,
             features = VariableFeatures(sobj))

#check the scaling, a new layer is added called scale.data
sobj 
sobj@assays$RNA$counts[1:10,1:10]
sobj@assays$RNA$data[1:10,1:10]
sobj@assays$RNA$scale.data[1:10,1:10]

```

# activity2
perform normalization using different method and check and comapre how the numbers change
normalization.method	
Method for normalization.
LogNormalize: Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p.
CLR: Applies a centered log ratio transformation
RC: Relative counts. Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. No log-transformation is applied. For counts per million (CPM) set scale.factor = 1e6

perform scaling using using all the genes
Pros of using highly variable features to scale:
Focuses on informative genes: HVFs are the genes that show the most meaningful variation across cells, likely reflecting true biological differences.
Faster & less memory-intensive: You're only scaling a subset of the data.
Reduces noise: Non-variable genes often contribute mostly technical noise.
cons:
You exclude potentially important genes that didn't meet the variance threshold.

pros of using all genes to scale:
Retains all genes: Useful when you need scaled values for genes outside the HVF set (e.g., for scoring gene sets or marker genes not in HVFs).
No bias toward only highly variable genes: For analyses like differential expression or custom scoring, this is sometimes better.
Cons:
Computationally heavier
Noisy genes may dominate

```{r activity2}
# perform normalization using different method and check and comapre how the numbers change

# increase the variable features to 5000 and check the variable features, do the top 10 change?

```


## Perform Principle Component Analysis (PCA)
This step can also be known as linear dimensional reduction. Numerous methods exist for visualization of principle components (PCs). A common method is `ElbowPlot`, which plots the standard deviations of the principle components. The left of the 'elbow' tends to represent the significant dimensions. 

Other visualization methods include `JackStraw` (slow) and `DimHeatmap`. Note that higher dimensions can be chosen, especially if certain rare cell-type signatures are distinguishable in these PCs. `DimHeatmap` is helpful in seeing these genes. More information can be found at: [Seurat - Guided Clustering Tutorial: Perform linear dimensional reduction](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html).

```{r pca}

#### Run PCA ####
sobj <-
    RunPCA(sobj,
          features = VariableFeatures(object = sobj))

#### Elbow Plot ####
# Note the elbow around PC 20
# (more specifically, the PCs from above PC 20 show nearly no difference in y-axis 
# value to that of PC 20). Thus, we will proceed with dims = 1:20.
ElbowPlot(sobj,
          ndims = 50)

```

## Cluster Single Cells.
This portion includes two steps. The first, `FindNeighbors`, takes the principle components selected above and constructs a graph based on the euclidean distance from the PCA and finds the similarities between two cells based on overlap of their local neighborhoods. The selected PCs are defined in the `dims` parameter.

The second step, `FindClusters`, iteratively groups cells together with the ability to set the resolution, or granularity, of the clustering. The higher the resolution, the greater the number of clusters in the output.

```{r cluster}

#### Find Neighbors ####
sobj <-
    FindNeighbors(sobj,
                  dims = 1:20)

#### Find Clusters ####
# Low resolution (low number of clusters)
sobj <-
    FindClusters(sobj,
                resolution = 0.2)
```

## Run Non-linear Dimensional Reduction (UMAP/tSNE)
Using the same PCs and resolution as defined above, this step places similar cells together in a low-dimension (2D) space. Those cells calculated as co-localized in the clusters generated above should co-localize in the UMAP or tSNE space. Cells are colored by their cluster or identity class. As for differences between tSNE and UMAP, UMAP lowers the dimensions of the high dimensional graph from above using compression, whereas t-SNE moves the high  dimensional graph to a lower dimensional space points by points. Their parameters, thus, differ, but both can be used as graph-based methods for non-linear dimensional reduction. Neither is perfect, but UMAP tends to provide better balance between local versus global structure and is more time-effective. For more information, please see [Understanding UMAP](https://pair-code.github.io/understanding-umap/).

`RunTSNE()` is used, similarly to below, to acheive a TSNE plot.

```{r umap}

#### Run UMAP ####
# Use the same dimensions as those in FindNeighbors()
sobj <-
    RunUMAP(sobj,
            dims = 1:20,
            metric = "euclidean")
DimPlot(sobj,
        reduction = "umap")
# Visualize cluster labels on UMAP
## Note: reduction parameter, if not specified, first searches for umap, then tsne, then pca
## By default, then, it will select whichever dimensional reduction method is used above,
## unless both umap and tsne have been generated. 
DimPlot(sobj,
        label = TRUE,
        repel = TRUE)


#run tSNE, a different method of dimensional reduction
sobj <-
    RunTSNE(sobj,
            dims = 1:20,
            metric = "euclidean")

#### Plot UMAP and tSNE together ####
DimPlot(sobj,
        reduction = "umap") +
    DimPlot(sobj,
        reduction = "tsne")

# save processed object
qs_save(sobj, "~/workshop_data/rdata/f1_sobj_processed.qs2") 
```

## Practice Activity 3

Generate a tSNE plot with a higher clustering resolution of 1.0. (You do not need to rerun anything prior to FindClusters).

```{r activity3}





```


# Yogesh's part2
## Celltype Annotation
Marker based annotation is a common method for annotating cell types.
This method uses known marker genes for each cell type to identify the cell types in the dataset.

here we will use dotplot to visualize the expression of the markers in the clusters.
We can also use feature plot to visualize the expression of the markers in the clusters.
We can also use Violin plot to visualize the expression of the markers in the clusters.
```{r using markers}
# Load the reference data
DimPlot(sobj,
        group.by = "seurat_clusters",
        label = TRUE,
        repel = TRUE)

# we will use resolution 0.2 for the clustering for this part
sobj <-
    FindClusters(sobj,
               resolution = 0.2)

# example plot of DotPlot, feature plot and violin plot
DotPlot(sobj,
        features = "Dcn",
        cols = c("blue", "red"))

FeaturePlot(sobj,
            features = "Dcn")


FeaturePlot(sobj,
            features = "Dcn") +
  DimPlot(sobj)

VlnPlot(sobj,
        features = "Dcn",
        group.by = "seurat_clusters") +
    theme(axis.text.x = element_text(angle = 90)) # makes the gene names 90 degrees


# Load the reference data
markers <- list(
    fibroblast = c("Col1a2", "Dcn", "Pdgfra"),
    T_cell = c("Cd2", "Cd3d", "Il7r"),  
    myeloid = c("Cd68", "Fcgr3", "Cd14"),  
    B_cell = c("Cd79a", "Cd79b", "Cd19"),
    NK_cell = c("Nkg7", "Klrk1","Klrd1"),
    endothelial_cell = c("Cldn5", "Pecam1", "Plvap"),  
    plasma_cell = c("Sdc1", "Chst1", "Pou2af1"),  
    brush_cell = c("Serpina1d", "Slc22a7", "Ceacam1"), 
    kidney_epithelial_cell = c("Spp2", "Sox9", "Gpx6"),
    kidney_cortex_artery_cell = c("Sox17", "Fbln5", "Tm4sf1")
)
# plot and test one by one
DotPlot(sobj,
        features = markers$fibroblast,
        cols = c("blue", "red"),
        dot.scale = 8)

# plot and test all the markers
markers_all <-
    markers %>%
    unlist() %>%
    as.character() %>%
    unique()

marker_dotplot <-
    DotPlot(sobj,
            features = markers_all,
            cols = c("blue", "red"),
            dot.scale = 8) +
        theme(axis.text.x = element_text(angle = 90)) # makes the gene names 90 degrees

# save the plot
ggsave("Sessions/beginner_tuesday/markers_dotplot.png",
        plot = marker_dotplot,
       width = 12,
       height = 7,
       bg = "white")

# Add the cell type annotation 
sobj$cell_type <-
    str_replace_all(sobj$seurat_clusters, 
                c("^0$" = "Fibroblast",
                    "^1$" = "Plasma cell",
                    "^2$" = "Plasmatocyte",
                    "^3$" = "Endothelial cell",
                    "^4$" = "Fibroblast",
                    "^5$" = "Fibroblast",
                    "^6$" = "Epithelial cell",
                    "^7$" = "Mesangial cell",
                    "^8$" = "Plasmatocyte",
                    "^9$" = "Macrophage",
                    "^10$" = "Lymphocyte",
                    "^11$" = "Epithelial cell"))

# check the object
dimplot_annotated <- 
    DimPlot(sobj,
            group.by = "cell_type",
            label = TRUE,
            repel = TRUE,
            label.box = T) +
    NoLegend()

# save the plot
ggsave("Sessions/beginner_tuesday/annotated_celltypes.png",
       plot = dimplot_annotated,
       width = 12,
       height = 7,
       bg = "white")

```


## Activity4
The dotplot allows us to visualize the expression of the markers in the clusters.
We can see that the clusters are not well separated. This is because the clusters
are not well defined. The clusters are not well defined because the cells are
not well separated. In this case we can visualize the markers in individual cells
using the feature plot. 
As a part of this exercise we will look at the immune cell types. We will subset the 
macrophages and lymphocytes and reprocess and annoate at a deeper level using the markers given below.
In this analysis we can test if there are other immune cells that are lumped together
into macrophages using feature plot. We can do this by using marker based feature plots. 
The markers for each celltypes are given below.
We will find that there are NK, B and T cells that are lumped together with microglia. 

```{r activity4}
# Subset the macrophages and lymphocytes and reprocess
markers <- list(
    T_cell = c("Cd2", "Cd3d", "Il7r"),  
    macrophage = c("Cd68","Cd163","Cd86"),  
    monocytes = c("Cd14","Fcgr3"),
    dendritic_cell = c("Itgax","Clec9a"),
    B_cell = c("Cd79a", "Cd79b", "Cd19"),
    NK_cell = c("Nkg7", "Klrk1","Klrd1"),
    plasma_cell = c("Sdc1", "Chst1", "Pou2af1")
)

```


## SingleR Annotation, reference based
An alternative approach would be to annotate cell by cell using other methods such as
SingleR or Mapquery. These methods use the reference data to annotate the cells.
```{r singleR}
# reference data
kidney_ref_data <- 
    readRDS("~/workshop_data/rdata/tabula_muris_senis_kidney.rds")
DefaultAssay(kidney_ref_data) <- "RNA"

# check the object labels that will be used for annotation
DimPlot(kidney_ref_data,
        group.by = "cell_type",
        label = TRUE,
        repel = TRUE,
        reduction = "umap") +
    NoLegend()

# singleR annotation
annotation <-
    SingleR::SingleR(test = as.SingleCellExperiment(sobj),
            ref = GetAssayData(kidney_ref_data),
            labels = kidney_ref_data$cell_type,
            aggr.ref = TRUE)

# check the object
sobj$singleR_annotation <- annotation$labels

# cell scores
sobj$singleR_scores <- 
    apply(X = annotation$scores,
          MARGIN = 1,
          function(x) max(x, na.rm = TRUE))

# check thee annotation
DimPlot(sobj,
        group.by = "singleR_annotation",
        label = TRUE,
        label.box = T,
        repel = T) +
    NoLegend()
```



## Load in reference data
Load in the kidney reference Seurat object to serve as the reference for cell typing.

This object has several different levels of granularity. For the examples, we will use `cell_type`.
```{r annotate_seurat}
# Read in the reference Seurat object
kidney_ref_data <- 
    readRDS("~/workshop_data/rdata/tabula_muris_senis_kidney.rds")
DefaultAssay(kidney_ref_data) <- "RNA"

# Set the idents to cell_type so that we can examine which cell types are present in the dataset
Idents(kidney_ref_data) <- "cell_type"

# Visualize the reference dataset
DimPlot(kidney_ref_data, label = T) +NoLegend()

# Find anchors between reference and query datasets
anchors <-
    FindTransferAnchors(reference = kidney_ref_data,
                      query = sobj,
                      dims = 1:30,
                      reference.reduction = "pca")

# Transfer anchors from reference dataset
predictions <-
    TransferData(anchorset = anchors,
               refdata = kidney_ref_data$cell_type,
               dims = 1:30)

# Add query to metadata of Seurat object
sobj <- AddMetaData(sobj,
                    metadata = predictions) 

# Rename "predicted.id" to "cell_types"
sobj$anchor_cell_types <- sobj$predicted.id

# Set active identity to new cell type labels and visualize typed clusters via DimPlot
Idents(sobj) <- sobj$anchor_cell_types
DimPlot(sobj,
        reduction = "umap",
        label = T,
        label.box = T,
        repel = T) + NoLegend()
# Tally number of cells in each cell type 
table(sobj$anchor_cell_types)

# save
qs_save(sobj, "~/workshop_data/rdata/f1_sobj_processed_annotated.qs2") 

```

## Practice Exercise5

Now annotate the dataset using `free_annotation` instead of `cell_type`.

```{r annotate}







```