# activity1 
**Ribosomal and hemoglobin gene subsetting:**
These are other options for identifying low quality samples based on the percent
of these genes in each cell. Ribosomal genes begin with either "RPS" or "RPL". 
Hemoglobin genes begin with either "HBA" or "HBB".

make a new column in the metadata for the percent of ribosomal genes and hemoglobin genes. 
Then visualize the distribution of these genes using a violin plot and histogram.

Tips: the pattern for the ribosomal genes is "^Rp[sl]", which means that the gene name 
begins with "Rp" and is followed by either "s" or "l". The pattern for the hemoglobin genes
is "^Hb[ab]", which means that the gene name begins with "Hb" and is followed by either "a" or "b".

```{r activity1}
sobj[["percent.ribo"]] <- 
        PercentageFeatureSet(sobj, pattern = "^Rp[sl]")

#check the metadata
colnames(sobj@meta.data)

sobj[["percent.hb"]] <- 
        PercentageFeatureSet(sobj, pattern = "^Hb[ab]")

# visualize the composition or distribution
VlnPlot(sobj,
        features = c("percent.mt",
                     "percent.ribo",
                     "percent.hb"))

```


# activity2
perform normalization using different method and check and comapre how the numbers change
normalization.method	
Method for normalization.
LogNormalize: Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p.
CLR: Applies a centered log ratio transformation
RC: Relative counts. Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. No log-transformation is applied. For counts per million (CPM) set scale.factor = 1e6

perform scaling using using all the genes
Pros of using highly variable features to scale:
Focuses on informative genes: HVFs are the genes that show the most meaningful variation across cells, likely reflecting true biological differences.
Faster & less memory-intensive: You're only scaling a subset of the data.
Reduces noise: Non-variable genes often contribute mostly technical noise.
cons:
You exclude potentially important genes that didn't meet the variance threshold.

pros of using all genes to scale:
Retains all genes: Useful when you need scaled values for genes outside the HVF set (e.g., for scoring gene sets or marker genes not in HVFs).
No bias toward only highly variable genes: For analyses like differential expression or custom scoring, this is sometimes better.
Cons:
Computationally heavier
Noisy genes may dominate

```{r activity2}
# perform normalization using different method and check and comapre how the numbers change
new_sobj <- 
    NormalizeData(sobj,
                  normalization.method = "CLR",
                  scale.factor = 10000)

# increase the variable features to 5000 and check the variable features, do the top 10 change?
new_sobj <- 
    FindVariableFeatures(new_sobj,
                         selection.method = "vst",
                         nfeatures = 5000)

top10 <-
    head(VariableFeatures(new_sobj),
            10)

Seurat::LabelPoints(plot = VariableFeaturePlot(new_sobj),
                    points = top10,
                    repel = TRUE)

# perform scaling using all the genes
new_sobj <- 
    ScaleData(new_sobj,
                features = rownames(new_sobj))

# check the scaling, a new layer is added called scale.data
new_sobj
new_sobj@assays$RNA$counts[1:10,1:10]
new_sobj@assays$RNA$data[1:10,1:10]
new_sobj@assays$RNA$scale.data[1:10,1:10]
```

## Practice

Generate a tSNE plot with a higher clustering resolution of 1.0. (You do not need to rerun anything prior to FindClusters).

```{r practice}

#### Find Clusters ####
# Higher resolution (higher number of clusters)
sobj <-
  FindClusters(sobj,
               resolution = 1.0)

#### Run tSNE ####
# Use the same dimensions as those in FindNeighbors()
sobj <-
  RunTSNE(sobj,
          dims = 1:20)

#### Plot tSNE ####
DimPlot(sobj,
        reduction = "tsne",
        label = TRUE,
        repel = TRUE)


# Challenge: plot them side by side using the `group.by` parameter
#
# Hint: `group.by` "RNA_snn_res.0.2" for the low resolution and
#                  "RNA_snn_res.1" for the high resolution

#### Challenge UMAP ####
DimPlot(sobj,
        group.by = "RNA_snn_res.0.2",
        reduction = "umap",
        label = TRUE,
        repel = TRUE) + 
DimPlot(sobj,
        group.by = "RNA_snn_res.1",
        reduction = "umap",
        label = TRUE,
        repel = TRUE)


#### Challenge tSNE ####
DimPlot(sobj,
        group.by = "RNA_snn_res.0.2",
        reduction = "tsne",
        label = TRUE,
        repel = TRUE) + 
DimPlot(sobj,
        group.by = "RNA_snn_res.1",
        reduction = "tsne",
        label = TRUE,
        repel = TRUE)


```

## Activity4
The dotplot allows us to visualize the expression of the markers in the clusters.
We can see that the clusters are not well separated. This is because the clusters
are not well defined. The clusters are not well defined because the cells are
not well separated. In this case we can visualize the markers in individual cells
using the feature plot. 
As a part of this exercise we will look at the immune cell types. We will subset the 
microglia and reprocess and annoate in the deeper level using the markers given below.
In this analysis we can test if there are other immune cells that are lumped together
into microglia using feature plot. We can do this by using marker based feature plots. 
The marekrs for each celltypes are given below.
We will find that there are B and T cells that are lumped together with microglia. 

```{r activity4}
# Subset the macrophages and reprocess
markers <- list(
    T_cell = c("Cd2", "Cd3d", "Il7r"),  
    macrophage = c("Cd68","Cd163","Cd86"),  
    monocytes = c("Cd14","Fcgr3"),
    dendritic_cell = c("Itgax","Clec9a"),
    B_cell = c("Cd79a", "Cd79b", "Cd19"),
    NK_cell = c("Nkg7", "Klrk1","Klrd1"),
    plasma_cell = c("Sdc1", "Chst1", "Pou2af1")
)

# you can subset and pipe all the processing steps together
macrophages <-
    sobj %>%
    subset(seurat_clusters %in% c("9","10")) %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:20) %>%
    FindNeighbors() %>%
    FindClusters(resolution = 1)

qsave(macrophages, "/igm/home/cnh008/projects/scrgot/scrgot_2026/workshop_data/rdata/f1_sobj_MACROPHAGE_processed_annotated.qs") 

# check the object
DimPlot(macrophages,
        group.by = "seurat_clusters",
        label = TRUE,
        repel = TRUE) +
    NoLegend()

DotPlot(macrophages,
        features = markers %>% unlist() %>% as.character() %>% unique(),
        cols = c("blue", "red"),
        #scale = F,
        dot.scale = 8) +
    theme(axis.text.x = element_text(angle = 90)) # makes the gene names 90 degrees

DimPlot(macrophages, 
        group.by = "cell_type",
        label = TRUE,
        repel = TRUE) +
    NoLegend()
```

## Practice Exercise5

Now annotate the dataset using `free_annotation` instead of `cell_type`.

```{r annotate}

# Transfer anchors 
predictions2 <- TransferData(anchorset = anchors,
                             refdata = ref$free_annotation,
                             dims = 1:30)

# Add query to metadata of Seurat object 
sobj <- AddMetaData(sobj,
                    metadata = predictions2) 

# Rename "predicted.id" to "cell_types"
sobj$cell_types2 <- sobj$predicted.id

# Set active identity to new cell type labels and visualize typed clusters via DimPlot
DimPlot(sobj,
        reduction = "umap",
        group.by = "cell_types2")

# Tally number of cells in each cell type 
table(sobj$cell_types2)

# Challenge: plot them side by side using the `group.by` parameter

DimPlot(sobj,
        reduction = "umap",
        group.by = "anchor_cell_types") +
DimPlot(sobj,
        reduction = "umap",
        group.by = "cell_types2")

```