---
title: "Intro to R 2023 Answer Key"
author: "Matt Gust and Luda Popova"
date: '2023-04-23'
output: html_document
---
```{r load_packages}
library(tidyverse)
```
--------------------------------------------------------------------------------

# Part 1

## Activity 1

Create a new Rmd file. Use Code/Insert Chunk, ctrl + alt + i (Windows), or cmd + option + i (macOS).
Within that file, create two variables holding two different numbers.
Knowing that +, -, /, and * work like regular arithmetic operators in R, add your two variables together.

```{r}
num1 <- 3

num2 <- 5

num1 + num2
```

Using the "==" operator, see if your variables are equal to each other. Now use the != operator and observe the output. 

```{r}
num1 == num2

num1 != num2
```

Create two variables holding the following values: "5" and "15". Make sure to enclose your variables in quotes.
Try adding your two new variables together. Observe the error message.

```{r}
num3 <- "5"

num4 <- "15"

num3 + num4
```

Check the class of the variables using the class() function. 

```{r}
class(num3)

class(num4)
```

Using as.numeric() function, try to fix the error.

```{r}
as.numeric(num3) + as.numeric(num4)
```

Create a vector of categorical variables specifying days of the week. Convert them to factors, specifying the correct order.
Use sort() function to make sure everything worked.

```{r}
days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

sort(days)

factor_days <- factor(days, 
                      levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

sort(factor_days)
```

If you're fast:
Think of something interesting you could do with some of your variables. Look up the function that would help you do that and read the documentation. Try it out.

```{r}

```

--------------------------------------------------------------------------------

# Part 2

--------------------------------------------------------------------------------

## Activity 2

-- Make sure to save the summary output to a variable 

Read in the starwars dataset.

Use read.delim()
```{r}
starwars <- read.delim("Sessions/beginner_monday/starwars.txt", sep = " ")
```

Use the `head()` function to get a quick first-glance at the dataset. If things don't look right, try specifying the delimiter in read.delim in the first part.

```{r}
head(starwars)
```

Access the 3rd column of the data frame, i.e. the "mass" column. Save it to a variable. 
```{r}
starwars_mass <- starwars$mass
```

Use the function `max()` to find the tallest height of all the characters.
```{r}
max(starwars$height, na.rm = TRUE)
```

Access the tenth row of the dataset, save the output to a variable. 
```{r}
starwars_10th_row <- starwars[10,]
```

Retrieve the name of the person specified in row 10 using square brackets. 
```{r}
starwars_10th_row[1]

# OR

starwars[10, 1]
```

Create a histogram of `height`. What do you expect to happen?
```{r}
hist(starwars$height)
```

--------------------------------------------------------------------------------

# Part 3

--------------------------------------------------------------------------------

## Activity 3

Go back to our temperatures data frame (in wide format).  

```{r}
temperatures
```

Arrange the data frame by temperature in January from coldest to warmest and by temperature in August from hottest to coolest. 

```{r}
temperatures %>%
  arrange(Jan)

temperatures %>%
  arrange(-Aug)
```

Group the dat frame by country, and calculate the mean values for temperatures in May. 

```{r}
temperatures %>%
  group_by(Country)%>% 
  summarise(mean_May = mean(May))
```

Use filter to make a new data frame named us_temperatures that contains temperatures for cities in the United States only. 

```{r}
us_temperatures <- 
  temperatures %>%
  filter(Country == "United States")

us_temperatures
```

In the us_temperatures data frame, make a new column that includes the temperature in the US in January in Fahrenheit. Note: the formula for the conversion is Fahrenheit = (Celsius * 9/5) + 32. Note: the result will be appended as the last column. 

```{r}
us_temperatures %>%
  mutate(Jan_F = (Jan *9/5) +32)
```

If you are fast: go back to our rain_long data frame. Join the us_temperatures and the rain_long data frames to only include matching rows. Hint: use inner_join. 

```{r}
inner_join(us_temperatures, rain_long)
```

----------------------------------------------------------

# Part 4

## Activity 4 (time permitting)

Using the `iris` dataset, make a scatterplot of `Sepal.Length` vs `Sepal.Width`.
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point()
```

Create the same plot, but color the points by their `Species`, and size them by their `Petal.Length`.
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
    geom_point()
```

Create a histogram of `Petal.Width` with, with one plot for each `Species` of flower.
```{r}
ggplot(iris, aes(x = Petal.Width)) +
    geom_histogram() +
    facet_wrap("Species")
```

Super Duper Hard Impossible Boss-Level Challenge (see slide)

Hints:
    refer to the link from the "theme" subheader to figure out how to change the panel and plot backgrounds

    the panel background color is lavender, the plot background is lightgreen, the panel lines are black

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) +
    geom_boxplot() +
    theme(panel.background = element_rect(fill = "lavender"),
          panel.grid = element_line(color = "black", size = 0.2),
          plot.background = element_rect(fill = "lightgreen")) +
    labs(y = "Sepal Length",
         title = "Super Duper Hard Impossible Boss-Level Challenge Plot")
```