---
title: "Intro to R 2023 Answer Key"
author: "Matt Gust and Luda Popova"
date: '2023-04-23'
output: html_document
---
```{r load_packages}
library(tidyverse)
```
--------------------------------------------------------------------------------

# Part 1

# Activity 1

Create a new Rmd file. Use ctrl+alt+i to insert a new R chunk.

Within that file, create two variables holding numerical data.
Knowing that +, -, /, and * work like regular arithmetic operators in R, add your two variables together.
```{r}

```

Create two variables holding the following values: "10" and "1". Make sure to enclose your variables in quotes.
Try adding your two new variables together. Observe the error message.
```{r eval=FALSE}

```

Using as.numeric() function, try to fix the error.
```{r}

```

Create a set of categorical variables specifying days of the week. Convert them to factors, specifying the correct order.

Use sort() function to make sure everything worked.
```{r}

```

If you're fast:
Think of something interesting you could do with some of your variables. Look up the function that would help you do that and read the documentation. Try it out.
```{r}

```

--------------------------------------------------------------------------------

# Part 2


--------------------------------------------------------------------------------

## Activity 2

-- Make sure to save the summary output to a variable 

Read in the starwars dataset.

Use read.delim()
```{r}
starwars <- read.delim("Sessions/beginner_monday/starwars.txt", sep = " ")
```

Use the `head()` function to get a quick first-glance at the dataset. If things don't look right, try specifying the delimiter in read.delim in the first part.

```{r}
head(starwars)
```

Access the 3rd column of the data frame, i.e. the "mass" column. Save it to a variable. 
```{r}
starwars_mass <- starwars$mass
```

Use the function `max()` to find the tallest height of all the characters.
```{r}
max(starwars$height, na.rm = TRUE)
```

Access the tenth row of the dataset, save the output to a variable. 
```{r}
starwars_10th_row <- starwars[10,]
```

Retrieve the name of the person specified in row 10 using square brackets. 
```{r}
starwars_10th_row[1]

# OR

starwars[10, 1]
```

Create a histogram of `height`. What do you expect to happen?
```{r}
hist(starwars$height)
```

--------------------------------------------------------------------------------

# Part 3

--------------------------------------------------------------------------------

## Activity 3

Read in and check out the Titanic data
```{r}

```
Arrange the dataset by age from youngest to oldest and oldest to youngest
```{r}

```

Use filter to make two new dataframe variables with just male and female only and then calculate the mean/median values for age and fare. Basically, calculate the mean/median age and fare for men vs women
```{r}

```

Group the dataset by gender and calculate mean fare and age
```{r}

```

Make a new data frame selecting the following columns: Name, Age, Fare
```{r}

```

Make a new column 
```{r}

```

# Part 4

## Activity 4 (time permitting)

Using the `iris` dataset, make a scatterplot of `Sepal.Length` vs `Sepal.Width`.
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point()
```

Create the same plot, but color the points by their `Species`, and size them by their `Petal.Length`.
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
    geom_point()
```

Create a histogram of `Petal.Width` with, with one plot for each `Species` of flower.
```{r}
ggplot(iris, aes(x = Petal.Width)) +
    geom_histogram() +
    facet_wrap("Species")
```

Super Duper Hard Impossible Boss-Level Challenge (see slide)

Hints:
    refer to the link from the "theme" subheader to figure out how to change the panel and plot backgrounds

    the panel background color is lavender, the plot background is lightgreen, the panel lines are black

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) +
    geom_boxplot() +
    theme(panel.background = element_rect(fill = "lavender"),
          panel.grid = element_line(color = "black", size = 0.2),
          plot.background = element_rect(fill = "lightgreen")) +
    labs(y = "Sepal Length",
         title = "Super Duper Hard Impossible Boss-Level Challenge Plot")
```