---
title: "Intro to R 2024"
author: "Matt Gust and Luda Popova"
date: '2024-05-06'
output: html_document
---

--------------------------------------------------------------------------------

# Part 1

## Overview

-- RStudio overview and layout.

-- R markdown, creating an Rmd file.

R markdown is a way to combine plain text, your R code, and results from your data analysis (including plots and tables) into a nicely formatted and reproducible document.
https://intro2r.com/what-is-r-markdown.html

-- To create a new R chunk in an Rmd file, use Code/Insert Chunk, ctrl + alt + i (Windows), or cmd + option + i (macOS).

## Variables and basic data types 

"=" or "<-" is an assignment operator, though "<-" is preferred.

Variables in R can be used to store letters, numbers, and various data structures. To create a new variable, simply assign a value to a name.

Fundamental data types in R include: numeric, integer, complex, character, and logical. For our purposes, we will mostly be using numeric, character, and logical data types. 

Here, we create a variable holding character data. Characters are enclosed in quotes.

```{r}
var1 <- "Hello"
```

You can use class(var1) or typeof(var1) to examine to which data type does the variable belong.

```{r}
class(var1)

typeof(var1)
```

Here, we create a variable holding numeric data.

```{r}
count <- 10

class(count)
```

Here, we create a variable holding logical (boolean) data.

```{r}
bool1 <- TRUE

bool2 <- FALSE

class(bool1)

class(bool2)
```

## Operators

Operators in R are used to perform operations on variables. Arithmetic operators (+, -, *, / etc.) perform mathematical operations.

Relational operators, e.g. <, >, <=, =>, are useful for comparisons. Two relational operators, ==, and != are more "unusual". 
"==" checks if the variables are equal. "!=" checks whether they are not equal.  

```{r}
5 == 5

5 != 5

5 == 4
```

## Variable types

### Vectors

- A vector can hold multiple values, which can be joined by using c(value1, value2, value3, value4).
- A vector holds data of a single type (all numbers, all characters, etc..).

```{r}
days <- c("Monday", "Tuesday", "Wednesday", "Thursday")

is.vector(days)
```

### Lists

- Lists are like vectors, but they can hold different types of data.
- Elements in a list can have names.

```{r}
list_of_stuff <-
    list("first",
         2,
         c(TRUE, TRUE, FALSE))
list_of_stuff

other_list <-
    list("name_1" = "first",
         "name_2" = 2,
         "name_3" = c(TRUE, TRUE, FALSE))

other_list$name_3
```

### Data frames

- Data frames are tables of data similar to excel worksheets.
- Each column can be of a different data type, but everything in a column has to be of the same data type.
- A data frame is a table or a two-dimensional array-like structure in which each column contains values of one variable and each row contains one set of values from each column. This is similar to an Excel sheet. 
- Data frames have the following characteristics (https://www.tutorialspoint.com/r/r_data_frames.htm):
    - The column names should be non-empty.
    - The row names should be unique.
    - The data stored in a data frame can be of numeric, factor or character type.
    - Each column should contain same number of data items.
    
```{r}
foods <-
    data.frame(food = c("hamburger", "coffee", "pie"),
               price = c("7", "2", "12"))

foods
```

### Matrices

- Matrices are like vectors, but they can have multiple dimensions.
- Unlike dataframes, all elements in a matrix have to be of the same data type.

```{r}
my_matrix <- 
  matrix(c(1, 2, 3, 4, 5, "bob"),
                    nrow = 2,
                    ncol = 3)
my_matrix

is.numeric(my_matrix)
```

## Factors

Variables can be continuous and categorical. Continuous can assume an infinite number of values within a given interval. Categorical variables have a fixed and known set of possible values.

In R, factors are used to work with categorical variables. They are also useful when you want to display character vectors in a non-alphabetical order (https://r4ds.had.co.nz/factors.html#:~:text=In%20R%2C%20factors%20are%20used,to%20work%20with%20than%20characters).

```{r}
months <- c("January", "February", "March", "April", "May")
  
sort(months)

months_factor <- factor(months,
                        levels = c("January", "February", "March", "April", "May"))

sort(months_factor)
```

## Getting things done with functions

R uses functions to perform operations. For example, sort() and class() from above are functions. Functions are sets of statements that are designed to perform a specific task (https://www.tutorialspoint.com/r/r_functions.htm).

- Functions have a name, which is used to call them.
- They are followed by parentheses, which contain the arguments.
- They can return a value.
- Functions have arguments, which the user can pass to them.
- In R, there are a lot of in-built functions. Some examples of in-built fuctions include mean(), max(), sum(), min() (https://www.tutorialspoint.com/r/r_functions.htm).
- Also, in R users can define their own functions, but we won't go over this in this session.

```{r}
numbers <- c(1, 2, 3, 4, 5)

min(numbers)

sum(numbers)
```

It's impossible (and not necessary) to know all the functions and/or their arguments.You can use the help() function to get more information about any function you want to use.

https://www.r-project.org/help.html
```{r}
help(min)
```

## Naming guidelines in R 

There are a few rules for naming variables in R.

- Names should be short and meaningful.
- Should include numbers, letters, _, and -
- Avoid using special characters in names.
- R is case-sensitive, best practice is to use lowercase.

```{r}
fruit <- "orange"

FRUIT <- "apple"

fruit == FRUIT
```

--------------------------------------------------------------------------------

# Activity 1

Create a new Rmd file. Use Code/Insert Chunk, ctrl + alt + i (Windows), or cmd + option + i (macOS).
Within that file, create two variables holding two different numbers.
Knowing that +, -, /, and * work like regular arithmetic operators in R, add your two variables together.

```{r}

```

Using the "==" operator, see if your variables are equal to each other. Now use the != operator and observe the output. 

```{r}

```

Create two variables holding the following values: "5" and "15". Make sure to enclose your variables in quotes.
Try adding your two new variables together. Observe the error message.

```{r}

```

Check the class of the variables using the class() function. 

```{r}

```

Using as.numeric() function, try to fix the error.

```{r}

```

Create a vector of categorical variables specifying days of the week. Convert them to factors, specifying the correct order.
Use sort() function to make sure everything worked.

```{r}

```

Create a vector of 5 different numbers. Use the functions max() and mean() to find the largest number and the mean, respectively.

```{r}

```

If you're fast: Think of something interesting you could do with some of your variables. Look up the function that would help you do that and read the documentation. Try it out.

```{r}

```

--------------------------------------------------------------------------------

# Part 2

## Directories and Paths

** See powerpoint slides for info on directories and paths **

### Get Working directory

The command `getwd()` will tell you what your current working directory is.

```{r}
getwd()
```

The working directory should be NCH_Coder_Upgrade. We want to make it the beginner_monday folder that's within our Sessions folder. We can do this using `setwd()`. Since we're already within NCH_Coder_Upgrade, we don't need to type the entire path up to beginner_monday, we can use the relative path from within NCH_Coder_Upgrade

```{r}
setwd("Sessions/beginner_monday/")
getwd()
```

Now we should be in beginner_monday. Let's read some files in from our working directory!

## Reading in files

Data are usually available to us in form of larger files, i.e. not single variables. There are functions in R that enable us to read in files. read.delim() and read.table() can be used to read in text (.txt) files.
```{r}
westernQuant <-
    read.table("westernQuant.txt",
               header = TRUE)

westernQuant
```

Once we read in a data frame, we can manipulate the data it contains.

Here, we use an $ operator to access a column in a data frame.
```{r}
westernQuant$group_name
```

Square brackets can be used to subset the data frame.

data_frame[row(s), column(s)]
data_frame[, ] (empty) means everything.
```{r}
# first row, first column (just one element)
westernQuant[1, 1]

# First row, all columns
westernQuant[1, ]

# First column, all rows
westernQuant[, 1]

# First 5 rows, all columns
westernQuant[c(1:5),]
```

The functions max, min, mean, and median will return summary statistics. These functions -- along with many others -- won't return the expected result if there are NA values present. Use the argument `na.rm = TRUE` to get around this issue.

We can use functions to make plots. Here, we use our data to make a boxplot.
```{r plot}
boxplot(intensity_gene1 ~ group_name,
        data = westernQuant)
```

Here, we make a histogram of our data. 
```{r}
hist(westernQuant$intensity_gene1)
```

`plot()` is another commonly used base plotting function for creating scatterplots.


--------------------------------------------------------------------------------

## Activity 2

-- Make sure to save the summary output to a variable 

Read in the starwars dataset.


Use read.table()

```{r}

```

Use the `head()` function to get a quick first-glance at the dataset. If things don't look right, try specifying the delimiter in read.delim in the first part.

```{r}

```

Access the 3rd column of the data frame, i.e. the "mass" column. Save it to a variable. 
```{r}

```

Use the function `max()` to find the tallest height of all the characters.
```{r}

```

Access the tenth row of the dataset, save the output to a variable. 
```{r}

```

Retrieve the name of the person specified in row 10 using square brackets. 
```{r}

```

Create a histogram of `height`. What do you expect to happen?
```{r}

```

--------------------------------------------------------------------------------

# Part 3

## Packages

R packages are a collection of R functions, complied code and sample data (https://www.tutorialspoint.com/r/r_packages.htm). Packages are usually created to address a very specific need in data analysis.

Packages are created by the community and so conventions and quality vary widely.

Packages need to be installed and loaded into the session before they can be used.

Some packages come from The Comprehensive R Archive Network (CRAN) with install.packages(), other packages can be installed through Bioconductor through BiocManager::install(). Best way to find out how to install a given package is to Google the specific package.

Tidyverse is a very useful collection of packages. Tidyverse is an opinionated collection of R packages designed for data science (includes ggplot2, dplyr, and other ones). All packages share an underlying design philosophy, grammar, and data structures.

https://www.tidyverse.org/

```{r}
#install.packages("tidyverse")

library(tidyverse)
```

## A brief introduction to the tidyverse

Tidyverse packages use the pipe operator %>%

This passes data from one function to the next without having to re-call the original data.

data_1 <- function_1(data)
data_2 <- function_2(data_1)
output <- function_3(data_2)

output

or:

output <- function_3(function_2(function_1(data)))

or:

output <- 
    data%>%
    function_1()%>%
    function_2() %>%
    function_3()
    
output


## Several useful tidyverse functions

mutate() adds a new column to a data frame.

```{r}
westernQuant <- 
    westernQuant%>%
    mutate(intensity_gene1_log = log2(intensity_gene1))

westernQuant
```

filter() can be used to filter the data frame based on conditions.

```{r}
control_data <-
  westernQuant%>%
  filter(group_name == "control")

control_data
```

arrange() allow us to order a table by a column's values, in ascending or descending order.

```{r}
arranged1 <- 
  westernQuant %>%
  arrange(intensity_gene1)

arranged1

arranged2 <- 
  westernQuant %>%
  arrange(-intensity_gene1)

arranged2
```

select() allows to select a certain column or several columns.

```{r}
westernQuant %>%
  select(intensity_gene1, group_name)
```

summarize() and group_by() can work together: group() silently divides up rows by categories.
summarize() summarizes data within groups in a data frame. This makes a new column and drops all unnecessary columns.

```{r}
westernQuant %>%
  group_by(group_name) %>%
  summarize(mean_intensity_gene1 = mean(intensity_gene1))
```

## Brief intro into wide data vs long data

"tidy" data - https://www.youtube.com/watch?v=K-ss_ag2k9E explanation starts ~ 6 min in
https://epirhandbook.com/en/images/pivoting/pivot_longer_new.png
Many tidyverse functions require long data.
Wide data is data where the same measurement is in multiple columns split by something (time, etc). The data frame "temperatures" in an example of wide data. 

```{r}
temperatures <- read.delim("weather_temps.txt", header = TRUE)
temperatures
```

Long data has one column for each measurement and column(s) for the variable(s) that define metadata (time, experimental groups, etc). pivot_longer() "lengthens" data, increasing the number of rows and decreasing the number of columns.

```{r}
long_temps <-
    temperatures %>%
    pivot_longer(cols = c(Jan,
                          Feb,
                          Mar,
                          Apr,
                          May,
                          Jun,
                          Jul,
                          Aug,
                          Sep,
                          Oct,
                          Nov,
                          Dec),
                 names_to = "month",
                 values_to = "temp")

long_temps2 <-
    temperatures %>%
    pivot_longer(cols = c(-Country,
                          -City),
                 names_to = "month",
                 values_to = "temp")

long_temps2
```

Make it wider with pivot_wider()

```{r}
wide_again <-
  long_temps %>%
  pivot_wider(names_from = "month",
              values_from = "temp")

wide_again
```

## Joining data frames

left_join, right_join, inner_join, full_join
Graphical explanation of join types:
https://www.softwaretestinghelp.com/wp-content/qa/uploads/2019/05/Capture-1.jpg

Lets get some other data to use

```{r}
rain <- read_tsv("weather_precip.tsv")
rain

rain_long <-
    rain %>%
    pivot_longer(cols = c(-Region,
                          -Country,
                          -City,
                          -Yearly),
                 names_to = "month",
                 values_to = "rainfall")
rain_long
```

Join the data frames together. 

```{r}
colnames(long_temps)
colnames(rain_long)
nrow(long_temps)
nrow(rain_long)

left_join(long_temps,
          rain)

right_join(long_temps,
           rain)

full_join(long_temps,
          rain)

inner_join(long_temps,
           rain)
```


## Writing files out 

```{r}
write.table(wide_again,
            file = "exampleOutput.txt",
            quote = FALSE,
            sep = "\t",
            row.names = FALSE)

write_tsv(wide_again,
          file = "exampleOutputAlso.txt")
```


## General ideas

### Common error messages

```{r, eval=FALSE}
mean(variable1)
#Error in mean(variable1) : object 'variable1' not found
```
R is looking for a variable named variable1, but it doesn't exist
-   Either make it
-   Check spelling/capitalization
-   You forgot to put quotes around a character string

```{r, eval=FALSE}
function1(mtcars)
#Error in function1(mtcars) : could not find function "function1"
```
R is trying to use the function `arbleGarble()`, but it doesn't exist
-   Either load in the library that has it
    -   (or use `coolLib::arbleGarble()`)
-   Check spelling/capitalization

```{r, eval=FALSE}
hist(mtcars$mpg, n = 20

#Error: Incomplete expression: hist(mtcars$mpg, n = 20
#+
#((((((1))))))
```
R thinks you have unfinished business here
-   Hit escape to cancel the command
-   Most likely you're missing a parenthesis or a quotation mark somewhere
-   rainbow brackets!

You can also Google error messages and Google "how do I ..." 

--------------------------------------------------------------------------------

## Activity 3

Go back to our temperatures data frame (in wide format).  

```{r}
temperatures
```

Arrange the data frame by temperature in January from coldest to warmest and by temperature in August from hottest to coolest. 

```{r}

```

Group the dat frame by country, and calculate the mean values for temperatures in May. 

```{r}

```

Use filter to make a new data frame named us_temperatures that contains temperatures for cities in the United States only. 

```{r}

```

In the us_temperatures data frame, make a new column that includes the temperature in the US in January in Fahrenheit. Note: the formula for the conversion is Fahrenheit = (Celsius * 9/5) + 32. Note: the result will be appended as the last column. 

```{r}

```

If you are fast: go back to our rain_long data frame. Join the us_temperatures and the rain_long data frames to only include matching rows. Hint: use inner_join. 

```{r}

```

# Part 4

## GGplot2

This is a package that makes much nicer plots than the base R plotting functions
https://ggplot2.tidyverse.org/reference/

### geoms as layers

When plotting with ggplot2, you're generally building the plot with layers (geoms). A layer is just a word for a component of the plot that represents a specific aspect of the data visualization.

A geom is short for "geometric object", which is ggplot-speak for the elements in your plot that represent your data, such as a point for a scatterplot, a bar for a barplot, etc.

### ggplot
ggplot(data_frame_here,
       aes(x = x_here,
           y = y_here,
           other_stuff)) +
    layer_1 +
    layer_2 +
    layer_3 +
    . +
    . +
    . +
    modify_the_plot_somehow

### Creating a Layer

One argument you always need to provide when creating a ggplot object is the data. This should be a dataframe.

Another important aspect of ggplots is the aesthetics. This is set in the plot above by calling `aes()` in the `ggplot()` function call above. This is necessary for specifying your x and y variables, and it is also useful in changing how geoms in your graph are colored, shaped, and/or sized (more on these last 3 later).


### geom_point()
Add a scatterplot layer to the plot
```{r}
ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point()

# Set the overall look of the plot - I like the white background better than gray
theme_set(theme_bw())
# Put the title in the middle of the top instead of the left
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point() +
  theme_bw()
```

### aes(fill/color/shape/size)

If you put the argument outside of aes()

-   applies to everything in that layer
-   does not show up in the legend

If you put it inside aes()

-   Applies to parts of the layer
-   Shows up in legend

If it is in the aes() call inside of ggplot() `ggplot(aes(...))`

-   It applies to all layers
-   Unless that layer overwrites it

```{r}
mtcars %>%
    ggplot(aes(x = as.factor(cyl),
               y = hp)) +
    geom_boxplot(color = "red",
                 fill = "blue",
                 size = 4)

ggplot(mtcars,
       aes(x = hp,
           y = carb,
           color = as.factor(vs),
           size = disp)) +
  geom_point()
```

#### geom_histogram()
Better looking than hist()
Specify how many bars to show with `bins`
```{r}
hist(storms$pressure, n = 200)

ggplot(storms,
       aes(x = pressure)) +
  geom_histogram(bins = 200)
```


#### geom_col()
Add a barplot layer to the plot
```{r}
mtcars %>%
    rownames_to_column("car_make") %>%
    mutate(car_make = str_remove(car_make," .+")) %>%
    ggplot(aes(x = car_make, y = hp, color = as.factor(car_make))) +
    geom_col(fill = "black")

mtcars %>%
  rownames_to_column("car_make") %>%
  ggplot(aes(y = car_make, x = hp)) +
  geom_col()
```

#### labs()
Add labels
```{r}
ggplot(mtcars) +
  geom_point(aes(x = hp, y = disp)) +
  labs(x = "X label goes here",
       y = "Y label goes here",
       title = "This is a title",
       subtitle = "Subtitle!",
       caption = "Cannon et al, 2034!",
       tag = "A")
```

#### theme()
A good guide to what theme elements are what:
https://henrywang.nl/ggplot2-theme-elements-demonstration/

You can change just about any aspect of how the plot looks with theme()
You'll have to look up how to change each specific bit

### Facet

Sometimes you'll want to create subplots of your data. This is where `facet_wrap` and `facet_grid` come in. These functions allow us to split up a plot based on certain factor variables.

The difference between the two is that facet_wrap will "wrap" a 1-dimensional set of panels (split by only one variable) into 2 dimensions. facet_grid will create a 2-dimensional "grid" of panels in which one variable defines the rows and another defines the columns.

```{r}
ggplot(mtcars) +
    geom_point(aes(x = hp, y = disp)) +
    facet_wrap(facets = "cyl") +
    labs(title = "disp vs. hp Faceted by cyl")
```

IMPORTANT: The variable(s) that your faceting on need to be in quotes in the function call.

### ggsave()
```{r}
ggsave("examplePlot.png",
       width = 10,
       height = 8)

plot_in_variable <-
  ggplot(dplyr::storms,
         aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category)

# ggsave will get the plot size from how big your "plot" windows is in the right pane if you don't specify it!!!!!
ggsave("anotherExamplePlot.png",
       plot = plot_in_variable)
```

### pdf() and png()
```{r}
png("examplePlot.png",
    width = 10,
    height = 8)
hist(storms$pressure, n = 200)
dev.off()

pdf("examplePlot.pdf",
    width = 10,
    height = 8)
hist(storms$pressure, n = 200)
hist(storms$wind, n = 200)
dev.off()
```

## Activity 4 (time permitting)

Using the `iris` dataset, make a scatterplot of `Sepal.Length` vs `Sepal.Width`.
```{r}

```

Create the same plot, but color the points by their `Species`, and size them by their `Petal.Length`.
```{r}

```

Create a histogram of `Petal.Width` with, with one plot for each `Species` of flower.
```{r}

```


Super Duper Hard Impossible Boss-Level Challenges (see slide)

Hints:

For the first plot:
    refer to the link from the "theme" subheader to figure out how to change the panel and plot backgrounds
    the panel background color is lavender, the plot background is lightgreen, the panel lines are black

For the second plot:
    You'll need to reorder your month variable so that it goes in typical calendar order. You can do this with 
    `mutate(month = factor(month, levels = c("Jan",
                                            "Feb",
                                            "Mar",
                                            "Apr",
                                            "May",
                                            "Jun",
                                            "Jul",
                                            "Aug",
                                            "Sep",
                                            "Oct",
                                            "Nov",
                                            "Dec")))`

    Use the `group` aesthetic for the lines

```{r}

```
