---
title: "Intro to R 2023 day 1"
author: "Matt Cannon"
date: '2023-10-23'
output: html_document
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Day 1

RStudio overview
  Projects

Rmd format
  ctrl+alt+i to create R chunk

What's a variable?

What's a function?
  Arguments

Reading function documentation

## Functions for today

### read.delim()
Read a file into your environment and save it as a variable
```{r readAfile}
westernQuant <- read.delim("materials/westernQuant.txt", header = TRUE)
```

### summary()
Generate summary statistics for a variable
```{r summary}
mtcars <- mtcars

my_name <- "Matt"

mtcars = mtcars
summary(mtcars)
```

### plot()
Visualize data
```{r plot}
plot(mtcars$mpg, mtcars$cyl)

plot(as.factor(mtcars$cyl), mtcars$mpg)
```

### boxplot()
```{r}
boxplot(mtcars$mpg ~ mtcars$cyl)
```

### png()/pdf()
Save a plot to an image.
```{r saveAplot}
png(filename = "bob.png")
plot(1, 1)
dev.off()
```

### t.test()
Test for differences between groups
```{r ttest}
t.test(1:10, y = c(7:20))
t.test(extra ~ group, data = sleep)
```

## Activity

See the powerpoint for instructions

### Read in data
Use read.delim()
```{r readData}

```

### Summarize the data.frame
Use summary()
```{r summaryWestern}

```

### Test for group differences
Use t.test() function
```{r testDiff}

```

### Barplots
Use barplot()
```{r plots}

```

### I'm super fast
Try out wilcox.test()

