---
title: "Make Seurat Object"
author: Elizabeth Garfinkle
output: html_document
date: "2025-05-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Set up a Project
Go to File, then New Project
Choose: Version Control
Select: Git
Paste link to NCH_Coder_Upgrade GitHub: https://github.com/kidcancerlab/NCH_Coder_Upgrade
Name project: CoderUpgrade
Click: Create Project
You should now see a data folder and a CoderUpgrade folder in your Home directory

#Load libraries
We need to load the Seurat library and the tidyverse library which contains ggplot2 for plotting and dplyr for data frame manipulation and qs library for reading/saving objects
```{r}
library(Seurat)
library(tidyverse)  # This package contains ggplot2 and dplyr
library(qs) # For reading/saving objects
```

#Check and set working directory
getwd() # This will show you your current working directory
setd("") # You can set your working directory by putting your desired file path in the quotes or click Session, Set Working Directory, Choose Directory, then navigate to your file
```{r}
#check what your current working directory is
getwd() #should look like "/rstudio-workshop/home/username/CoderUpgrade" 
```

#Read 10X data
First, read in the filtered feature matrix output files via the Read10X Seurat function
We will start with reading in just one sample, OX1X, and call it ox1x_raw
```{r}
ox1x_raw <- 
  Read10X("~/data/input/scRNA/GSM6925133_OX1X")
```

#Make Seurat object
Now we will make a Seurat object via the CreateSeuratObject Seurat function from the data we just read in that we named ox1x_raw and call it ox1x_sobj
We also set two standard filters:
min.cells = 3 
  Here, we are filtering out any genes that are detected in fewer than 3 cells to   remove genes that are likely not expressed in a meaningful way
min.features = 200 
  Here, we are filtering out any cells that have less than 200 features (genes)    detected, which could indicate poor quality or dying cells
Both of these filters minimize noise/poor quality data for downstream analyses
```{r}
ox1x_sobj <-
    CreateSeuratObject(
        counts = ox1x_raw,
        min.cells = 3,
        min.features = 200
    )
```

#Rename sample 
By default, the orig.ident (original ident, aka original sample name) is the name that is assigned when making the Seurat object. Often, this name is short/easy to type but we may want a more meaningful/longer name when making plots for figures or publications. Here, we can reassign the orig.ident to have a more descriptive name. 
The use of the $ sign means you are accessing a metadata slot. 
```{r}
#Look at the orig.ident name 
View(ox1x_sobj)

#Reassign name
ox1x_sobj$orig.ident <- "GSM6925133_OX1X"

#Look at the orig.ident name now 
View(ox1x_sobj)

#Set orig.ident to be the active ident
Idents(ox1x_sobj) <- "orig.ident"

#Check
View(ox1x_sobj)
```

#Save Seurat object
We will now save our newly made Seurat object usign the qsave function. The object name you want to save is listed first and then the output file path and name you want to save the object as is listed in quotes. By default, the object will save in your current working directory (important to know what that is at all times!). If you want to save it in a subfolder within your current working directory, you can add the file path before the final name of the saved object. Use .qs to save the final object. 
```{r}
qsave(ox1x_sobj, "~/OX1X_sobj.qs") #save to your home directory 
```

#Remove data and Seurat object
Now that we have saved our object, we can remove the data and object in our Environment using rm.
```{r}
rm(ox1x_raw, ox1x_sobj)
```

#We can read in a saved object any time for additional analyses using the qread function. 
```{r}
ox1x_sobj <-
    qread("~/data/output/rdata_premade/OX1X_sobj.qs") #read in premade object 

#challenge - read in your own made object from your home directory (hint: give it a unique name so you don't overwrite the premade object that's already loaded):

```

#Challenge: 
Make a Seurat object from an additional sample in our data set, GSM6925134_OX2X, and call it ox2x_sobj. After you make the Seurat object, remove both ox1x_sobj and ox2x_sobj from your Environment. 

Write and run your code here:
```{r}

```

