---
title: "Intro to R 2023"
author: "Matt Cannon and Luda Popova"
date: '2023-04-23'
output: html_document
---
```{r load_packages}
library(tidyverse)
```
--------------------------------------------------------------------------------

# Part 1

## Overview

-- RStudio overview and layout.

-- R markdown, creating an Rmd file.

R markdown is a way to combine plain text, your R code, and results from your data analysis (including plots and tables) into a nicely formatted and reproducible document.
https://intro2r.com/what-is-r-markdown.html

-- To create a new R chunk in an Rmd file, use Code/Insert Chunk, ctrl + alt + i (Windows), or cmd + option + i (macOS).

## Variables and basic data types 

"=" or "<-" is an assignment operator, though "<-" is preferred.

Variables in R can be used to store letters, numbers, and various data structures. To create a new variable, simply assign a value to a name.

Fundamental data types in R include: numeric, integer, complex, character, and logical.

Here, we create a variable holding character data. Characters are enclosed in quotes.
```{r}
my_name <- "Matt"
my_name = "Matt"
```

You can use class(my_name) or typeof(my_name) to examine to which data type does the variable my_name belong.
```{r}
class(my_name)
typeof(my_name)
```

Here, we create a variable holding numeric data.
```{r}
count <- 10

class(count)
```

Here, we create a variable holding logical (boolean) data.
```{r}
bool1 <- TRUE

bool2 <- FALSE

class(bool1)

class(bool2)
```

## Factors

Variables can be continuous and categorical. Continuous can assume an infinite number of values within a given interval. Categorical variables have a fixed and known set of possible values.

In R, factors are used to work with categorical variables. They are also useful when you want to display character vectors in a non-alphabetical order (https://r4ds.had.co.nz/factors.html#:~:text=In%20R%2C%20factors%20are%20used,to%20work%20with%20than%20characters)
```{r}
months <- c("Dec", "Apr", "Jan", "Mar")

sort(months)

months_factor <- factor(months,
                        levels = c("Jan", "Mar", "Apr", "Dec"))

sort(months_factor)
```

## Variable types

### Vectors

- A vector can hold multiple values, which can be joined by using c(value1, value2, value3, value4).
- A vector holds data of a single type (all numbers, all characters, etc..)

```{r}
months <- c("Dec", "Apr", "Jan", "Mar")
is.vector(months)
```

### Lists

- Lists are like vectors, but they can hold different types of data.
- Elements in a list can have names

```{r}
list_of_stuff <-
    list("first",
         2,
         c(TRUE, TRUE, FALSE))
list_of_stuff

other_list <-
    list("first" = "first",
         "second" = 2,
         "third" = c(TRUE, TRUE, FALSE))
other_list$third
```

### Data frames

- Data frames are tables of data similar to excel worksheets.
- Each column can be of a different data type, but everything in a column has to be of the same data type.
- A data frame is a table or a two-dimensional array-like structure in which each column contains values of one variable and each row contains one set of values from each column. This is similar to an Excel sheet. 
- Data frames have the following characteristics (https://www.tutorialspoint.com/r/r_data_frames.htm):
    - The column names should be non-empty.
    - The row names should be unique.
    - The data stored in a data frame can be of numeric, factor or character type.
    - Each column should contain same number of data items.

```{r}
important_data <-
    data.frame(names = c("Matt", "Luda", "John"),
               pets = c("dog", "cat", "dog"),
               pie_or_cake = c("pie", "cake", "pie"))
important_data

important_data$names
```

### Matrices

- Matrices are like vectors, but they can have multiple dimensions.
- Unlike dataframes, all elements in a matrix have to be of the same data type.

```{r}
my_matrix <- matrix(c(1, 2, 3, 4, 5, "bob"),
                    nrow = 2,
                    ncol = 3)
my_matrix
is.numeric(my_matrix)
```

## Getting things done with functions

R uses functions to perform operations. For example, sort() and class() from above are functions. Functions are sets of statements that are designed to perform a specific task (https://www.tutorialspoint.com/r/r_functions.htm).

- Functions have a name, which is used to call them.
- They are followed by parentheses, which contain the arguments.
- They can return a value.
- Functions have arguments, which the user can pass to them.
- In R, there are a lot of in-built functions. Some examples of in-built fuctions include mean(), max(), sum(), min() (https://www.tutorialspoint.com/r/r_functions.htm).
- Also, in R users can define their own functions, but we won't go over this in this session.

```{r}
numbers <- c(1, 2, 3, 4, 5)

min(numbers)

max(numbers)

sum(numbers)

mean(numbers)
```

It's impossible (and not necessary) to know all the functions and/or their arguments.You can use the help() function to get more information about any function you want to use.
    - Use "q" to quit the help page.
How to read the help page
```{r}
help(min)

help(max)

help(sum)

help(mean)
```

## Naming guidelines in R 

There are a few rules for naming variables in R.

- Names should be short and meaningful
- Should include numbers, letters, _, and -
- Avoid using special characters in names
- R is case-sensitive, best practice is to use lowercase

```{r}
my_name <- "Matt"
MY_NAME <- "Luda"

my_name == MY_NAME
```

--------------------------------------------------------------------------------

# Activity 1

Create a new Rmd file. Use ctrl+alt+i to insert a new R chunk.

Within that file, create two variables holding numerical data.
Knowing that +, -, /, and * work like regular arithmetic operators in R, add your two variables together.
```{r}

```

Create two variables holding the following values: "10" and "1". Make sure to enclose your variables in quotes.
Try adding your two new variables together. Observe the error message.
```{r eval=FALSE}

```

Using as.numeric() function, try to fix the error.
```{r}

```

Create a set of categorical variables specifying days of the week. Convert them to factors, specifying the correct order.

Use sort() function to make sure everything worked.
```{r}

```

If you're fast:
Think of something interesting you could do with some of your variables. Look up the function that would help you do that and read the documentation. Try it out.
```{r}

```

--------------------------------------------------------------------------------

# Part 2

## Reading in files

Data are usually available to us in form of larger files, i.e. not single variables. There are functions in R that enable us to read in files. read.delim() can be used to read in text (.txt) files.
```{r}
westernQuant <- read_tsv("Sessions/01/westernQuant.txt")

westernQuant
```

Data frames are very useful data structures in R. westernQuant is a data frame.
```{r}
class(westernQuant)
```

Once we read in a data frame, we can manipulate the data it contains.

Here, we use an $ operator to access a column in a data frame.
```{r}
westernQuant$group_name
```

Square brackets can be used to subset the data frame.

data_frame[row(s), column(s)]
data_frame[, ] (empty) means everything.
```{r}
# first row, first column (just one element)
westernQuant[1, 1]

# First row, all columns
westernQuant[1, ]

# First column, all rows
westernQuant[, 1]

# First 5 rows, all columns
westernQuant[c(1:5), ]
```

Also, functions head() and tail() can be used to retrieve the first and the last six rows of a data frame, respectively.
```{r}
head(westernQuant)

tail(westernQuant)
```

We can also use functions to look at data.
```{r}
summary(westernQuant)
```

We can use the functions from the previous section.
```{r}
min(westernQuant$intensity_gene1)

max(westernQuant$intensity_gene1)

mean(westernQuant$intensity_gene1)
```

Many functions will return an error if the data contain NAs. To overcome this issue, you need to include an na.rm = TRUE argument.
```{r}
min(westernQuant$intensity_gene2)

max(westernQuant$intensity_gene2)

mean(westernQuant$intensity_gene2)

min(westernQuant$intensity_gene2, na.rm = TRUE)

max(westernQuant$intensity_gene2, na.rm = TRUE)

mean(westernQuant$intensity_gene2, na.rm = TRUE)
```

We can use functions to make plots. Here, we use our data to make a boxplot.
```{r plot}
boxplot(intensity_gene1 ~ group_name,
        data = westernQuant)
```

Here, we make a histogram of our data. 
```{r}
hist(westernQuant$intensity_gene1)
```

--------------------------------------------------------------------------------

## Activity 2

-- Make sure to save the summary output to a variable 

Read in data the titanic dataset
Use read.delim()
```{r}

```

Summarize the dataset
```{r}

```

Use the head() and tail() functions to take a closer look at the dataset
```{r}

```

Access the 3rd column of the data frame, i.e. the "Name" column. Save it to a variable. 
```{r}

```

Access the third row of the dataset, save the output to a variable. 
```{r}

```
Retrieve the name of the person specified in row 3 using square brackets. 
```{r}

```

Make variables that contain the Mean/median/min/max of age/fare
```{r}

```

--------------------------------------------------------------------------------

# Part 3

## Packages

R packages are a collection of R functions, complied code and sample data (https://www.tutorialspoint.com/r/r_packages.htm). Packages are usually created to address a very specific need in data analysis.

Packages are created by the community and so conventions and quality vary widely.

Packages need to be installed and loaded into the session before they can be used.

Some packages come from The Comprehensive R Archive Network (CRAN) with install.packages(), other packages can be installed through Bioconductor through BiocManager::install(). Best way to find out how to install a given package is to Google the specific package.

Tidyverse is a very useful collection of packages. Tidyverse is an opinionated collection of R packages designed for data science (includes ggplot2, dplyr, and other ones). All packages share an underlying design philosophy, grammar, and data structures.

https://www.tidyverse.org/
```{r}
#install.packages("tidyverse")

library(tidyverse)

# Backup plan if you run into lots of errors there...

#install.packages(c("dplyr", "ggplot2"))

library(dplyr)

library(ggplot2)
```

## A brief intro to the tidyverse
Tidyverse packages use the pipe operator %>%

This passes data from one function to the next without having to re-call the original data

temp_1 <- function_1(my_data)
temp_2 <- function_2(temp_1)
output <- function_3(temp_2)
rm(temp_1, temp_2)

is the same as:

function_3(function_2(function_1(my_data)))

is the same as:

output <-
    function_1(my_data) %>%
    function_2() %>%
    function_3()

Several useful tidyverse functions:

### mutate() adds a new column to a data frame
```{r}
westernQuant <-
    westernQuant %>%
    mutate(intensity_gene1_log = log2(intensity_gene1))
westernQuant
```

### select() allows to select a certain column or several columns.
```{r}
westernQuant %>%
  select(intensity_gene1, group_name)
```

### rename() allows to rename a column.
```{r}
westernQuant %>%
  rename(Group = group_name)
```

### arrange() allow us to order a table by a column's values, in ascending or descending order.
```{r}
arranged1 <- westernQuant %>%
  arrange(intensity_gene1)

head(arranged1)

tail(arranged1)

arranged2 <- westernQuant %>%
  arrange(-intensity_gene1)

head(arranged2)

tail(arranged2)
```

### summarize() and group_by()
group() silently divides up rows by categories
summarize() summarizes data within groups in a data frame. This makes a new column and drops all unnecessary columns.
```{r}
westernQuant %>%
  group_by(group_name) %>%
  summarize(mean_intensity = mean(intensity_gene1)) %>%
  ungroup()
```

### pivot_longer()
wide data vs long data

"tidy" data - https://www.youtube.com/watch?v=K-ss_ag2k9E explanation starts ~ 6 min in
https://epirhandbook.com/en/images/pivoting/pivot_longer_new.png

Many tidyverse functions require long data

wide data is data where the same measurement is in multiple columns split by something (time, etc)
```{r}
temperatures <- read.delim("Sessions/01/weather_temps.txt", header = TRUE)
temperatures
```

long data has one column for each measurement and column(s) for the variable(s) that define metadat (time, experimental groups, etc).

pivot_longer() "lengthens" data, increasing the number of rows and decreasing the number of columns.
```{r}
long_temps <-
    temperatures %>%
    pivot_longer(cols = c(Jan,
                          Feb,
                          Mar,
                          Apr,
                          May,
                          Jun,
                          Jul,
                          Aug,
                          Sep,
                          Oct,
                          Nov,
                          Dec),
                 names_to = "month",
                 values_to = "temp")

long_temps <-
    temperatures %>%
    pivot_longer(cols = c(-Country,
                          -City),
                 names_to = "month",
                 values_to = "temp")

long_temps
```

# Make it wider with pivot_wider()
```{r}
wide_again <-
  long_temps %>%
  pivot_wider(names_from = "month",
              values_from = "temp")

wide_again
```

### Joining dataframes
left_join, right_join, inner_join, full_join
Graphical explanation of join types:
https://www.softwaretestinghelp.com/wp-content/qa/uploads/2019/05/Capture-1.jpg

Lets get some other data to use
```{r}
rain <- read_tsv("Sessions/01/weather_precip.tsv")
rain

rain <-
    rain %>%
    pivot_longer(cols = c(-Region,
                          -Country,
                          -City,
                          -Yearly),
                 names_to = "month",
                 values_to = "rainfall")
rain
```

```{r}
colnames(long_temps)
colnames(rain)
nrow(long_temps)
nrow(rain)

left_join(long_temps,
          rain)

right_join(long_temps,
           rain)

full_join(long_temps,
          rain)

inner_join(long_temps,
           rain)

outer_join(long_temps,
           rain)
```

### Write a table out to a text file
```{r}
write.table(wide_again,
            file = "exampleOutput.txt",
            quote = FALSE,
            sep = "\t",
            row.names = FALSE)

write_tsv(wide_again,
          file = "exampleOutputAlso.txt")
```

## General stuff

### Common error messages
```{r, eval=FALSE}
mean(bob)
#Error in mean(bob) : object 'bob' not found
```
R is looking for a variable named bob, but it doesn't exist
-   Either make it
-   Check spelling/capitalization
-   You forgot to put quotes around a character string

```{r, eval=FALSE}
arbleGarble(mtcars)
#Error in arbleGarble(mtcars) : could not find function "arbleGarble"
```
R is trying to use the function `arbleGarble()`, but it doesn't exist
-   Either load in the library that has it
    -   (or use `coolLib::arbleGarble()`)
-   Check spelling/capitalization

```{r, eval=FALSE}
hist(mtcars$mpg, n = 20
# +
((((((1))))))
```
R thinks you have unfinished business here
-   Hit escape to cancel the command
-   Most likely you're missing a parenthesis or a quotation mark somewhere
-   rainbow brackets!

### googling error messages

### googling "R how do I…"

### = and == have different meaning in R. 

```{r}
a <- 1

b <- 4

a == b

a = b

a

a != 5
```

--------------------------------------------------------------------------------

## Activity 3

Read in and check out the Titanic data
```{r}

```
Arrange the dataset by age from youngest to oldest and oldest to youngest
```{r}

```

Use filter to make two new dataframe variables with just male and female only and then calculate the mean/median values for age and fare. Basically, calculate the mean/median age and fare for men vs women
```{r}

```

Group the dataset by gender and calculate mean fare and age
```{r}

```

Make a new data frame selecting the following columns: Name, Age, Fare
```{r}

```

Make a new column 
```{r}

```

# Part 4

## GGplot2

This is a package that makes much nicer plots than the base R plotting functions
https://ggplot2.tidyverse.org/reference/

### geoms as layers

When plotting with ggplot2, you're generally building the plot with layers (geoms)

### ggplot
ggplot(data_frame_here,
       aes(x = x_here,
           y = y_here,
           other_stuff)) +
    layer_1 +
    layer_2 +
    layer_3 +
    . +
    . +
    . +
    modify_the_plot_somehow

### geom_point()
Add a scatterplot layer to the plot
```{r}
ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point()

# Set the overall look of the plot - I like the white background better than gray
theme_set(theme_bw())
# Put the title in the middle of the top instead of the left
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point() +
  theme_bw()
```

### aes(fill/color/shape/size)
Change colors/shapes/etc

If you put the argument outside of aes()

-   applies to everything in that layer
-   does not show up in the legend

If you put it inside aes()

-   Applies to parts of the layer
-   Shows up in legend

If it is in the aes() call inside of ggplot() `ggplot(aes(...))`

-   It applies to all layers
-   Unless that layer overwrites it

```{r}
mtcars %>%
  ggplot(aes(x = as.factor(cyl),
             y = hp)) +
  geom_boxplot(color = "red",
               fill = "blue",
               size = 4)

ggplot(mtcars,
       aes(x = hp,
           y = carb,
           color = as.factor(vs),
           shape = as.factor(gear),
           size = disp)) +
  geom_point()
```

#### geom_histogram()
Better looking than hist()
Specify how many bars to show with `bins`
```{r}
hist(storms$pressure, n = 200)

ggplot(storms,
       aes(x = pressure)) +
  geom_histogram(bins = 200)
```


#### geom_col()
Add a barplot layer to the plot
```{r}
mtcars %>%
  rownames_to_column("car_make") %>%
  mutate(car_make = str_remove(car_make," .+")) %>%
  ggplot(aes(x = car_make, y = hp)) +
  geom_col(color = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("this is a title") +
  theme_update(plot.title = element_text(hjust = 0.5))

mtcars %>%
  rownames_to_column("car_make") %>%
  ggplot(aes(y = car_make, x = hp)) +
  geom_col()
```

#### labs()
Add labels
```{r}
ggplot(mtcars) +
  geom_point(aes(x = hp, y = disp)) +
  labs(x = "X label goes here",
       y = "Y label goes here",
       title = "This is a title",
       subtitle = "Subtitle!",
       caption = "Cannon et al, 2034!",
       tag = "A")
```


#### theme()
A good guide to what theme elements are what:
https://henrywang.nl/ggplot2-theme-elements-demonstration/

You can change just about any aspect of how the plot looks with theme()
You'll have to look up how to change each specific bit
```{r}
ggplot(storms,
       aes(x = pressure,
           fill = category)) +
  # alpha sets transparency - 0 is clear, 1 is opaque
  geom_density(alpha = 0.5) +
  # The brewer color palettes are pretty, and "Set2" is color-blind friendly
  scale_color_brewer(palette = "Set2") +
  theme(legend.position = c(0.2, 0.9),
        legend.direction = "horizontal",
        legend.key.width = unit(2, "cm"),
        legend.title = element_text(face = "bold"),
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text.x = element_text(size = 30, face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

### ggsave()
```{r}
ggsave("examplePlot.png",
       width = 10,
       height = 8)

plot_in_variable <-
  ggplot(dplyr::storms,
         aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category)

# ggsave will get the plot size from how big your "plot" windows is in the right pane if you don't specify it!!!!!
ggsave("anotherExamplePlot.png",
       plot = plot_in_variable)
```

### pdf() and png()
```{r}
png("examplePlot.png",
    width = 10,
    height = 8)
hist(storms$pressure, n = 200)
dev.off()

pdf("examplePlot.pdf",
    width = 10,
    height = 8)
hist(storms$pressure, n = 200)
hist(storms$wind, n = 200)
dev.off()
```

## Activity

## Recreate plots from paper

Data is from https://doi.org/10.1371/journal.pbio.2005756

Najafov A, Zervantonakis IK, Mookhtiar AK, Greninger P, March RJ, Egan RK, et al. (2018) BRAF and AXL oncogenes drive RIPK3 expression loss in cancer. PLoS Biol 16(8): e2005756. https://doi.org/10.1371/journal.pbio.2005756

I downloaded their figure data and made it easier to import:
-   Fig1A_partial.txt
-   Fig1C.txt
-   Fig2F.txt
-   Fig3F.txt

#### Figure 1A - geom_boxplot()
```{r}

```

#### Figure 2F - geom_point()
```{r}

```

#### Figure 3F - geom_col
use geom_errorbars()
and facet_wrap() to split the plots by cell_line
```{r}

```

## Session challenge
Keep going 🙃

Recreate

- Fig1E
- Fig1F
- Fig2E
- Fig3C
- Fig3D

Download the data from the supplemental data here:

https://doi.org/10.1371/journal.pbio.2005756.s001

