---
title: "Intro to R 2023"
author: "Matt Cannon"
edits: "Luda Popova"
date: '2032-04-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

RStudio overview, R studio layout 
  Projects

Rmd format, how to create a Rmd file
  ctrl+alt+i to create R chunk

What's a variable?

What's a function?
  Arguments

Reading function documentation

## Basic variable types 

"=" or "<-" is an assignment operator

### String 

```{r}
my_name <- "Matt" 
```

### Numeric

```{r}
count = 10 
```

### Boolean 

```{r}
bool = TRUE

bool = FALSE
```

### Factors. In R, factors are used to work with categorical variables, variables that have a fixed and known set of possible values. They are also useful when you want to display character vectors in a non-alphabetical order (https://r4ds.had.co.nz/factors.html#:~:text=In%20R%2C%20factors%20are%20used,to%20work%20with%20than%20characters)

```{r}
months <- c("Dec", "Apr", "Jan", "Mar")

sort(months)

months_factor = factor(months, level = c("Jan", "Mar", "Apr", "Dec"))

sort(months_factor)
```

As you can see, you can join multiple variables by using c(var1, var2)

### Use the following to check the type of variable, e.g trying to add up variables that are not of the same class may not work

```{r}
class(variable)
```

## Functions for today

### sort() and class() from examples above are functions. Functions are sets of statements that are designed to perform a specific task (https://www.tutorialspoint.com/r/r_functions.htm). Functions have arguments, which the user can pass to them. In R, there are a lot of in-built functions. Some examples of in-built fuctions include mean(), max(), sum(), min() (https://www.tutorialspoint.com/r/r_functions.htm). Also, in R users can define their own functions, but we won't go over this in this session. 

```{r}
numbers = c(1,2,3,4,5)
min(numbers)
max(numbers)
sum(numbers)
mean(numbers)
```

## It's impossible (and not necessary) to know all the functions and/or their arguments. 

```{r}
help(min)
help(max)
help(sum)
help(mean)
```

# Activity one 

Create two character, numerical, and two boolean variables. 

Knowing that "+" adds 

Create a set of factor variables , specifying the correct order. 

Knowing that ,    


# Part 2 

## Data are usually available to us in form of larger files, i.e. not single variables. There are functions in R that enable us to read in files. 

### read.delim()

```{r readAfile}
westernQuant <- read.delim("intro_to_R_2022/westernQuant.txt", header = TRUE)
```

### Another useful function is read.csv()

```{r}
data = read.csv("filepath.csv", header = TRUE)
```
-- Not sure how to tie together the reading in a file and using the in-built mtcars dataset. -- One thought -- save the mtcars dataset as a .delim file and replace some of the values with NAs to incorporate Matt's idea to show that functions won't work on NA-containing data

### Mtcars is a data frame (https://www.tutorialspoint.com/r/r_data_frames.htm)

A data frame is a table or a two-dimensional array-like structure in which each column contains values of one variable and each row contains one set of values from each column.

The following are the characteristics of a data frame.

The column names should be non-empty.
The row names should be unique.
The data stored in a data frame can be of numeric, factor or character type.
Each column should contain same number of data items.

## Once we read in the file, we can manipulate the data it contains. 

### Accessing values in a column 

```{r}
mtcars$cyl
```

### Subsetting a data.frame with [ ]
data_frame[row(s), column(s)]
  empty means everything

```{r}
# first row, first column (just one element)
mtcars[1, 1]
# First row, all columns
mtcars[1, ]
# First column, all rows
mtcars[, 1]
mtcars$mpg
# First ten rows, all columns
mtcars[c(1:10), ]
```

### We can also use functions to manipulate the data

### summary()

```{r summary}
summary(mtcars)
```

### We can use the functions from above

```{r}
min(mtcars$cyl)

max(mtcars$mpg)

mean(mtcars$disp)
```
---------------------------
Many of these functions can get angry about missing data (`NA`).

Include na.rm = TRUE as an argument to ignore missing data

#### This doesn't work
```{r dontruns, eval=FALSE}
mean(c(1, 3, 4, NA, 34))
```

#### This works
```{r runs}
mean(c(1, 3, 4, NA, 34), na.rm = TRUE)
```
----------------------------

### We can use finction to make visualizations, e.g. plot()

```{r plot}
plot(mtcars$mpg, mtcars$cyl)

plot(as.factor(mtcars$cyl), mtcars$mpg)
```



## Activity 2

### Use read.csv to read in a 
### Read in data
Use read.delim()
```{r readData}

```

### Summarize the data.frame
Use summary()
```{r summaryWestern}

```

### Test for group differences
Use t.test() function
```{r testDiff}

```

### Barplots
Use barplot()
```{r plots}

```

### I'm super fast
Try out wilcox.test()




# Part 3

## packages -- what is a package? 
# R packages are a collection of R functions, complied code and sample data.  
#https://www.tutorialspoint.com/r/r_packages.htm. Packages are usually created to address a very specific need in data analysis. 

## Tidyverse is a very useful collection of packages
#The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and #data structures.
#https://www.tidyverse.org/

### Packages need to be installed and loaded into the session before they can be used

```{r}
#install.packages("tidyverse", Ncpus = 8)
library(tidyverse)

# Backup plan if you run into lots of errors there...
# By the way, any text that starts with a "#" is a comment and not executed
#install.packages(c("dplyr", "ggplot2"))
#library(dplyr)
#library(ggplot2)
```

### merge
merge() will attach two data.frames side to side, but match data in shared columns
-   by default, does an inner join
-   use all, all.x or all.y to specify other join types

https://www.datasciencemadesimple.com/wp-content/uploads/2017/06/merge-in-R-2.jpg?ezimgfmt=rs:535x142/rscb1/ng:webp/ngcb1
```{r}
band_members
band_instruments
merge(band_members, band_instruments)
merge(band_members, band_instruments, all = TRUE)
```

### mutate
Add a new column or change an existing one
```{r}
km_per_mi <- 1.60934

mtcars %>%
  mutate(kpg = mpg * km_per_mi)

mtcars %>%
  mutate(mpg = mpg * km_per_mi)
```

### summarize() and group_by()
group() silently divides up rows by categories
summarize() data within groups in a dataframe
Makes a new column and drops all columns not in group_by() or created by summarize()
```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))

starwars %>%
  group_by(homeworld, species) %>%
  summarize(mean_height = mean(height),
            max_mass = max(mass))
```

### t()
Rotate a dataframe (or matrix) 90 degrees
Also turns the data into a matrix, so we re-convert it using as.data.frame()
```{r}
rotated_cars <-
  t(mtcars) %>% # |>
  as.data.frame()
```

### pivot_longer()
wide data vs long data
"tidy" data - https://www.youtube.com/watch?v=K-ss_ag2k9E explanation starts ~ 6 min in
https://epirhandbook.com/en/images/pivoting/pivot_longer_new.png
```{r}
# wide data
relig_income

# long data
long_data <-
  relig_income %>%
  pivot_longer(!religion,
               names_to = "income",
               values_to = "count")

# Make it wider
wide_again <-
  long_data %>%
  pivot_wider(names_from = "income",
              values_from = "count")
```

### Write a table out to a text file
```{r}
write.table(wide_again,
            file = "exampleOutput.txt",
            quote = FALSE,
            sep = "\t",
            row.names = FALSE)

write_tsv(wide_again,
          file = "exampleOutputAlso.txt")
```





## Activity

### Read in and check out the Titanic data
```{r}

```

### Make variables that contain the Mean/median/min/max of age/fare
```{r}

```

### Use filter to make two new dataframe variables with just male and female only and then calculate the mean/median values for age and fare
Basically, calculate the mean/median age and fare for men vs women
```{r}

```

### Make a histogram of the age all passengers
```{r}

```

### Make another histogram for only passengers who survived

## If you're super fast

### Use `aov()` to test which factors influenced who survived the titanic disaster
```{r}

```

### Use table() to compare how many women vs men survived
```{r}

```


# Day 3

## General stuff

### Common error messages
```{r, eval=FALSE}
mean(bob)
#Error in mean(bob) : object 'bob' not found
```
R is looking for a variable named bob, but it doesn't exist
-   Either make it
-   Check spelling/capitalization
-   You forgot to put quotes around a character string

```{r, eval=FALSE}
arbleGarble(mtcars)
#Error in arbleGarble(mtcars) : could not find function "arbleGarble"
```
R is trying to use the function `arbleGarble()`, but it doesn't exist
-   Either load in the library that has it
    -   (or use `coolLib::arbleGarble()`)
-   Check spelling/capitalization

```{r, eval=FALSE}
hist(mtcars$mpg, n = 20
# +
((((((1))))))
```
R thinks you have unfinished business here
-   Hit escape to cancel the command
-   Most likely you're missing a parenthesis or a quotation mark somewhere
-   rainbow brackets!

### googling error messages

### googling "R how do I…"

---

## Functions for today's activity

### cbind, rbind
attach two data.frames either side to side or top to bottom
rbind will arrange column names
cbind just puts things together without checking order
  Generally a bad idea to cbind unless you're very careful
```{r}
efficient <-
  mtcars %>%
  filter(mpg > 30)

gas_guzzlers <-
  mtcars %>%
  filter(mpg < 14.5)

rbind(efficient, gas_guzzlers)

# This is total nonsense, but R does it anyways
test <- 
  cbind(efficient, gas_guzzlers)
```



## Activity

### Read in patientGroups.txt and exercise.txt
patientGroups.txt is patient #, treatment groups and final blood glucose measurements apatientGroups.txt is patient #, treatment groups and final blood glucose measurementspatientGroups.txt is patient #, treatment groups and final blood glucose measurementspatientGroups.txt is patient #, treatment groups and final blood glucose measurementspatientGroups.txt is patient #, treatment groups and final blood glucose measurementspatientGroups.txt is patient #, treatment groups and final blood glucose measurementspatientGroups.txt is patient #, treatment groups and final blood glucose measurements
exercise.txt is the how many minutes each patient exercised across five days
```{r}

```


### Combine the datasets into a single data.frame
```{r}

```

### Pivot the data to long form
Columns should be patient, day, exercise_min, glucose
```{r}

```

### Save the pivoted data.frame to a text file with write.table() or write_tsv()
```{r}

```

### Make a new column where glucose is multiplied by 1000
```{r}

```

### Calculate the average daily minutes of exercise per patient
```{r}

```

## If you’re super fast

### Plot glucose levels for each group (treated/control)
```{r}

```

### Test if glucose levels are statistically different between groups
```{r}

```

### Plot daily exercise minutes per group
```{r}

```

### Test if exercise minutes is statistically different between groups
```{r}

```





# Part 4 

## General stuff

## GGplot

### geoms as layers
https://ggplot2.tidyverse.org/reference/

## Functions for today's activity, also for reference

### ggplot
ggplot(data_frame_here, aes(x = x_here, y = y_here, other_stuff)) +
layer_1 +
layer_2 +
layer_3 +
. +
. +
. +
modify_the_plot_somehow

#### geom_point() ##############################################################
Add a scatterplot layer to the plot
```{r}
ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point()

# Set the overall look of the plot - I like the white background better than gray
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point() +
  theme_bw()
```

#### aes(fill/color/shape/size) ################################################
Change colors/shapes/etc

If you put the argument outside of aes()

-   applies to everything in that layer
-   does not show up in the legend

If you put it inside aes()

-   Applies to parts of the layer
-   Shows up in legend

If it is in the aes() call inside of ggplot() `ggplot(aes(...))`

-   It applies to all layers
-   Unless that layer overwrites it

```{r}
mtcars %>%
  ggplot(aes(x = as.factor(cyl),
             y = hp)) +
  geom_boxplot(color = "red",
               fill = "blue",
               size = 4)

ggplot(mtcars,
       aes(x = hp,
           y = carb,
           color = as.factor(vs),
           shape = as.factor(gear),
           size = disp)) +
  geom_point()
```

#### geom_histogram()
Better looking than hist()
Specify how many bars to show with `bins`
```{r}
hist(storms$pressure, n = 200)

ggplot(storms,
       aes(x = pressure)) +
  geom_histogram(bins = 200)
```

#### geom_density()
Smooth your data out
This can actually hide some of the nuance in your data
```{r}
ggplot(storms,
       aes(x = pressure,
           fill = category)) +
  # alpha sets transparency -- 0 is clear, 1 is opaque
  geom_density(alpha = 0.1, adjust = 0.5) +
  # The brewer color palettes are pretty, and "Set2" is color-blind friendly
  scale_color_brewer(palette = "Set2")
```

#### geom_col() ################################################################
Add a barplot layer to the plot
```{r}
mtcars %>%
  rownames_to_column("car_make") %>%
  mutate(car_make = str_remove(car_make," .+")) %>%
  ggplot(aes(x = car_make, y = hp)) +
  geom_col(color = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("this is a title") +
  theme_update(plot.title = element_text(hjust = 0.5))

mtcars %>%
  rownames_to_column("car_make") %>%
  ggplot(aes(y = car_make, x = hp)) +
  geom_col()
```

#### geom_errorbar() ###########################################################
Add error bars as a layer to plot
```{r}
summarized_hp <-
  mtcars %>%
  group_by(cyl) %>%
  summarize(mean_hp = mean(hp),
            sd_hp = sd(hp))

summarized_hp

ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp,
                    xmin = 2,
                    xmax = 2))

# This has error bars in the back since that layer was added first!
ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp),
                color = "red") +
  geom_bar(stat = "identity") +
  geom_point(color = "blue")
```

#### geom_jitter()
Drop-in replacement for geom_point()
Adds noise in x and y directions by default so you can see individual points
```{r}
ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp)) +
  geom_jitter(data = mtcars,
              aes(x = as.factor(cyl),
                  y = hp),
              color = "red", height = 0)
```

#### ggbeeswarm package
Adds noise in x direction and pulls things to the center to make it look nicer than jitter
```{r}
ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp)) +
  ggbeeswarm::geom_beeswarm(data = mtcars,
                            aes(x = as.factor(cyl),
                                y = hp),
                            color = "red")

matt_swarm <- function(...) {
  ggbeeswarm::geom_beeswarm(...)
}
```

#### geom_boxplot() ############################################################
Boxplots!
```{r}
mtcars %>%
  ggplot(aes(x = as.factor(cyl), y = hp)) +
  #geom_boxplot() +
  geom_violin(adjust = 1) +
  ggbeeswarm::geom_beeswarm()


```

#### geom_smooth
Add a regression line to your plot
Can specify the type of regression with `method`
Can remove gray confidence interval with `se = FALSE`
```{r}
ggplot(mtcars,
       aes(x = mpg,
           y = disp)) +
  geom_point() +
  geom_smooth()
```

#### Use multiple datasets in the same plot
Contents of the columns used as axes have to match
Each geom using it's own data needs to have "data = " in it
```{r}
ggplot(summarized_hp,
       aes(x = as.factor(cyl),
           y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp),
                width = 0.5) +
  geom_point(data = mtcars,
             aes(x = as.factor(cyl),
                 y = hp),
             color = "red")
```

#### labs()
Add labels
```{r}
ggplot(mtcars) +
  geom_point(aes(x = hp, y = disp)) +
  labs(x = "X label goes here",
       y = "Y label goes here",
       title = "This is a title",
       subtitle = "Subtitle!",
       caption = "Cannon et al, 2034!",
       tag = "A")
```

#### facet_wrap() ##############################################################
```{r}
dplyr::storms

ggplot(dplyr::storms,
       aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category + status)

ggplot(dplyr::storms,
       aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category, scales = "free_y")

ggplot(dplyr::storms,
       aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category,
             nrow = 1,
             scales = "free_y")
```

#### theme()
A good guide to what theme elements are what:
https://henrywang.nl/ggplot2-theme-elements-demonstration/

You can change just about any aspect of how the plot looks with theme()
You'll have to look up how to change each specific bit
```{r}
ggplot(storms,
       aes(x = pressure,
           fill = category)) +
  # alpha sets transparency - 0 is clear, 1 is opaque
  geom_density(alpha = 0.5) +
  # The brewer color palettes are pretty, and "Set2" is color-blind friendly
  scale_color_brewer(palette = "Set2") +
  theme(legend.position = c(0.2, 0.9),
        legend.direction = "horizontal",
        legend.key.width = unit(2, "cm"),
        legend.title = element_text(face = "bold"),
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text.x = element_text(size = 30, face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

### ggsave()
```{r}
ggsave("examplePlot.png",
       width = 10,
       height = 8)

plot_in_variable <-
  ggplot(dplyr::storms,
         aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category)

# ggsave will get the plot size from how big your "plot" windows is in the right pane if you don't specify it!!!!!
ggsave("anotherExamplePlot.png",
       plot = plot_in_variable)
```

### Include pdf() and png()
### dev.off()

## Activity

## Recreate plots from paper

Data is from https://doi.org/10.1371/journal.pbio.2005756

Najafov A, Zervantonakis IK, Mookhtiar AK, Greninger P, March RJ, Egan RK, et al. (2018) BRAF and AXL oncogenes drive RIPK3 expression loss in cancer. PLoS Biol 16(8): e2005756. https://doi.org/10.1371/journal.pbio.2005756

I downloaded their figure data and made it easier to import:
-   Fig1A_partial.txt
-   Fig1C.txt
-   Fig2F.txt
-   Fig3F.txt

#### Figure 1A – geom_boxplot()
```{r}

```

#### Figure 1C – heatmap()
```{r}

```

#### Figure 2F – geom_point()
```{r}

```

#### Figure 3F – geom_col
use geom_errorbars()
and facet_wrap() to split the plots by cell_line
```{r}

```

## If you’re super fast:
Keep going, do the rest 🙃

Download the data from the supplemental data here:

https://doi.org/10.1371/journal.pbio.2005756.s001

