
## Library
```{r lib, cache = FALSE}
# Load necessary packages
library(Seurat)
library(tidyverse) # This package contains ggplot2 and dplyr
library(patchwork)
library(ggplot2) # For plotting
library(rrrSingleCellUtils) # For plotting an processing
library(qs) # For reading and writing .qs files
library(harmony) # For Harmony integration
library(crescendo)

options(future.globals.maxSize = 10 * 1e9)

# Set the random generator seed so that results are reproducible.
set.seed(132)
```



## load the objects to be merged
In this tutorial, we will demonstrate how to merge two single cell RNA-seq
datasets using the MergeSeurat function in Seurat. Merging single cell RNA-seq 
datasets is a crucial step in the analysis of single cell data. 
It allows researchers to combine data from different sources and 
increase the sample size for more robust and comparative analysis.
Let's say we have many different datasets from different samples or
experiments. We can load the datasets individually and then merge them using
the merge function in Seurat. The merged dataset can then be used for
downstream analysis, such as dimensionality reduction, clustering, and
differential expression analysis.

```{r test}

# read in one object
f1_sobj <-
    qs::qread("workshop_data/rdata/f1_sobj_processed_annotated.qs")
f1_sobj
#list of Seurat objects from the same experiment (GSE232482)
GSE232482_sobj_list <-
    qs::qread("workshop_data/rdata/GSE232482_object_list_processed_annotated.qs")
GSE232482_sobj_list

# merge the Seurat objects
merged_GSE232482_sobj <-
    merge(x= f1_sobj,
          y = GSE232482_sobj_list[1:length(GSE232482_sobj_list)])

# check the merged Seurat object, you will see layers for each seurat object
merged_GSE232482_sobj

# join the layers so that they are all in the same object
merged_GSE232482_sobj <-
    merged_GSE232482_sobj %>%
    JoinLayers()

# check the object now, the layers are now joined or merged
merged_GSE232482_sobj

# need to re-run the normalization and scaling
merged_GSE232482_sobj <- 
    merged_GSE232482_sobj %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData()

# run PCA
merged_GSE232482_sobj <- 
    merged_GSE232482_sobj %>%
    RunPCA()

# run UMAP
merged_GSE232482_sobj <- 
    merged_GSE232482_sobj %>%
    RunUMAP(dims = 1:30)   #use PC1 through PC30

# run clustering
merged_GSE232482_sobj <- 
    merged_GSE232482_sobj %>%
    FindNeighbors(dims = 1:30) %>%    #use PC1 through PC30
    FindClusters()

# all of the above steps can be done in one go using pipe function
merged_GSE232482_sobj <-
    merged_GSE232482_sobj %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:30) %>%  
    FindNeighbors(dims = 1:30) %>%
    FindClusters(resolution = 0.5)

# check the object
merged_GSE232482_sobj

# visualize the clusters
DimPlot(merged_GSE232482_sobj,
        group.by = "seurat_clusters",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

# assessing the batch effect
DimPlot(merged_GSE232482_sobj,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

DimPlot(merged_GSE232482_sobj,
        group.by = "seurat_clusters",
        split.by = "orig.ident",
        ncol = 6,
        label = T,
        repel = T) +
    NoLegend()

# looks like there is no batch effect

#save the object
qs::qsave(merged_GSE232482_sobj,
          "workshop_data/rdata/merged_GSE232482_sobj.qs")

```

## Activity1
We just found out that there is minimal to no batch effect in the merged object. This could 
be because the samples are from the same tissue and the same experiment and source.
Now, we will load another object from a different experiment and source but still the
same tissue. The actvity is to merge the two objects and check for batch effect.
1. Load the second object
```{r activity1}
# read in the second object list

GSE252772_sobj_list <-
    qs::qread("workshop_data/rdata/GSE252772_object_list_processed_annotated_small.qs")

# merge the second object list


#save the batch information


# save the batch information in the merged_GSE232482_sobj as well


#merge the two objects and process it together through plotting dimplot


# check the object


# assess the batch effect presence


# save the object


```

## Harmony integration
It is great to see that there is no batch effect in the previous data. But that is 
not always the case. Most of the time, we will see a batch effect in the data because 
of the different samples or experiments. In this case, we can use the Harmony integration
method to remove the batch effect. Harmony is a method for integrating single-cell RNA-seq
data from multiple samples or experiments. It uses a graph-based approach to identify and
remove batch effects while preserving biological variation. Harmony can be used to integrate
data from different sources, such as different tissues, time points, or experimental conditions.
Harmony only corrects batch effect for plotting and visualization purposes. It does not
change the underlying data. So, we will still need to do batch correction in actual data using
other method like deseq2 or limma.

Here we will load another object from a different expertiment and source but still the 
same tissue. We will first merge and assess the presence of batch effect and then run
Harmony integration to remove the batch effect.

```{r harmony_integration}
# read in the second object list
GSE252772_sobj_list <-
    qs::qread("workshop_data/rdata/GSE252772_object_list_processed_annotated_small.qs")

# merge the second object list
merged_GSE252772_sobj <-
    merge(x= GSE252772_sobj_list[[1]],
          y = GSE252772_sobj_list[[2]])

#save the batch information
merged_GSE252772_sobj$batch <- "GSE252772"

# save the batch information in the merged_sobj as well
merged_GSE232482_sobj <-
    qs::qread("workshop_data/rdata/merged_GSE232482_sobj.qs")
merged_GSE232482_sobj$batch <- "GSE232482"

# merge the two objects and process it together unto plotting dimplot
merged_sobj <-
    merge(x= merged_GSE232482_sobj,
          y = merged_GSE252772_sobj)

merged_sobj <-
    merged_sobj %>%
    JoinLayers()

merged_sobj <-
    merged_sobj %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:30) %>%  
    FindNeighbors(dims = 1:30) %>%
    FindClusters(resolution = 0.5)

qs::qsave(merged_sobj,
          "workshop_data/rdata/merged_sobj_before_harmony.qs")

# assess the batch effect
DimPlot(merged_sobj,
        group.by = "batch",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

DimPlot(merged_sobj,
        group.by = "orig.ident",
        label = T,
        #repel = T,
        label.box = T) +
    NoLegend()


# run Harmony integration
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = "batch") %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters(resolution = 0.5) 

# check the integration
DimPlot(merged_sobj_harmony,
        group.by = "batch",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

# run Harmony integration with different parameters
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = "batch",
               theta = 5) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()
qs::qsave(merged_sobj_harmony,
    "workshop_data/rdata/merged_sobj_harmony_theta.qs")

# check the integration
DimPlot(merged_sobj_harmony,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()
DimPlot(merged_sobj_harmony,
        group.by = "batch",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()
```

## Activity2
In this activity, we will run harmony integration with different parameters/batch.
Use orig.ident as batch variable and see how the integration works.
```{r activity2}

#simply use the same code as above but change the batch variable to orig.ident



# check the integration using dimplot


# run Harmony integration with orig.ident as batch variable and theta value of your preference


# check the integration


```


# run harmony integration with different parameters like multiple batches
Harmony intergration can also be done with multiple batches. Most of the other integration methods
like Seurat and Scanpy can only do one batch at a time. But Harmony can do multiple batches at once.

```{r two_batch_variables}
# run Harmony integration with multiple batch variables
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = c("batch", "orig.ident")) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()
qs::qsave(merged_sobj_harmony,
    "workshop_data/rdata/merged_sobj_harmony_sample_batch.qs")

# check the integration
DimPlot(merged_sobj_harmony,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

DimPlot(merged_sobj_harmony,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

# change theta
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = c("batch", "orig.ident"),
               theta = c(10, 10)) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()
qs::qsave(merged_sobj_harmony,
    "workshop_data/rdata/merged_sobj_harmony_sample_batch_theta.qs")

# check the integration
DimPlot(merged_sobj_harmony,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

DimPlot(merged_sobj_harmony,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

# there are opportunities to explore more of the harmony parameters such as lambda and tau.
# you can also try using multiple theta each time, low and high and see how it works.
# you can also try using multiple lambda each time, low and high and see how it works.

# Clean up a bit to save ram
ls()
rm(f1_sobj, GSE232482_sobj_list, GSE252772_sobj, GSE252772_sobj_list, merged_GSE232482_sobj, merged_GSE252772_sobj, merged_sobj, merged_sobj_harmony)
gc()
```


## RPCA and CCA integration. 

We are going to go back to the merged object with batch effects and test two other integration methods, RPCA, and CCA, which are supported by Seurat v5, i.e. you don't need a separate package to run them. RPCA stands for reciprocal PCA (principal component analysis). CCA strands for canonical correlation analysis. You can read more about these methods here. https://satijalab.org/seurat/articles/seurat5_integration. 
To use these methods, we will need to go back to the merged object that contains 20 samples (with a batch effect). 

```{r}
# Read in the merged object
merged <- qread("workshop_data/rdata/merged_sobj_before_harmony.qs")
```

# Let's remind ourselves what that object looks like. We will make a DimPlot. 

```{r}
# We can group the DimPlot by clusters. 
DimPlot(merged, 
        group.by = "seurat_clusters",
        label = T,
        repel = T,
        label.box = T) +
  NoLegend() 

# We can also group the DimPlot based on the sample. 
DimPlot(merged, 
        group.by = "orig.ident")
```

# We are going to start by using the RPCA integration. 

Out of the two methods, RPCA is faster and more conservative, meaning that in results in smaller batch effect correction. It also runs faster than CCA. 

```{r}
# Lets downsample a bit so it runs quicker
Idents(merged) <- merged$orig.ident
merged <- subset(merged, downsample = 500)

DimPlot(merged,
        group.by = "orig.ident")

# First, we need to split the layers of the merged object to allow for integration to happen. 

merged[["RNA"]] <- split(merged[["RNA"]], f = merged$orig.ident)

# Then, we can perform the actual integration. 

rpca <- IntegrateLayers(
  object = merged, 
  method = RPCAIntegration,
  orig.reduction = "pca", 
  new.reduction = "integrated.rpca",
  verbose = FALSE)
```

# After performing the actual integration, we need to run PCA, UMAP, FindNeighbors() and FindClusters(). 

```{r}
rpca <- 
  rpca%>%
  RunPCA(verbose = FALSE)%>%
  RunUMAP(reduction = "integrated.rpca", dims = 1:30)%>%
  FindNeighbors(dims = 1:30, reduction = "integrated.rpca")%>%
  FindClusters(resolution = 0.1)

qs::qsave(rpca,
    "workshop_data/rdata/merged_sobj_rpca.qs")
```

# Visualize the integrated dataset and compare it to the merged object.

```{r}
# We can group the DimPlot by clusters. 

p1 <-
DimPlot(merged, 
        group.by = "seurat_clusters",
        label = T,
        repel = T,
        label.box = T) +
  NoLegend() 

p2 <-
DimPlot(rpca, 
        group.by = "seurat_clusters",
        label = T,
        repel = T,
        label.box = T) +
  NoLegend() 

p1 + p2

# We can also group the DimPlot based on the sample. 

p1 <-
DimPlot(merged, 
        group.by = "orig.ident")

p2 <-
DimPlot(rpca, 
        group.by = "orig.ident")

p1 + p2
```

# Activity 3. CCA integration. 

Comparing to the RPCA integration, CCA is less conservative, i.e. it results in a greater level of batch correction. It also takes a longer time to run. 

```{r}
# Using the code above, run CCA integration. Hint: you only need to change the method's name in the code for running rpca. If needed, use https://satijalab.org/seurat/articles/seurat5_integration, run CCA integration. 



# Make dimplots to check the integration


```
















