
## Activity1
We just found out that there is minimal to no batch effect in the merged object. This could 
be because the samples are from the same tissue and the same experiment and source.
Now, we will load another object from a different experiment and source but still the
same tissue. The actvity is to merge the two objects and check for batch effect.
1. Load the second object
```{r activity1}
# read in the second object
object1 <-
    qs::qread("input/scRNA/downloaded_source1.qs")

#save the batch information
object1$batch <- "batch1"

# save the batch information in the merged_sobj as well
object2 <-
    qs::qread("input/scRNA/downloaded_source2.qs")
object2$batch <- "batch2"
object2$cell_type <- object2$SingleR_annotation

#merge the two objects and process it together unto plotting dimplot



# assess the batch effect presence




```


## Activity2
In this activity, we will run harmony integration with different parameters/batch.
Use orig.ident as batch variable and see how the integration works.
```{r activity2}

#simply use the same code as above but change the batch variable to orig.ident
harm_obj2 <-
    merge_all %>%
    RunHarmony(group.by.vars = "orig.ident") %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()


# check the integration
DimPlot(harm_obj2,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()


DimPlot(harm_obj2,
        group.by = "cell_type",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()



# run Harmony integration with orig.ident as batch variable and theta value of your preference
harm_obj3 <-
    merge_all %>%
    RunHarmony(group.by.vars = "orig.ident",
               theta = 4) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()

# check the integration
DimPlot(harm_obj3,
        group.by = "orig.ident",
        label = T,
        repel = T,
        label.box = T) +
    NoLegend()

DimPlot(harm_obj3,
        group.by = "cell_type",
        label = T,
        repel = T,
        label.box = T) +




```
