---
title: "Creating additional Seurat Objects"
output: html_notebook
---

# 4 h5ad datasets were downloaded from https://allen-brain-cell-atlas.s3.us-west-2.amazonaws.com/index.html#expression_matrices/WMB-10Xv2/20230630/. 
#TH - thalamus, MB - midbrain, CTXsp - cortical subplate, HY - hypothalamus. 


## Load libraries for analysis. 

```{r}
library(Seurat)

library(anndata)

library(tidyverse)

library(biomaRt)

library(SingleR)
```

## Create Seurat objects from the h5ad data. Preprocess each of the objects and downsample them. Each of the object is filtered using diferent filtering cutoffs. 

```{r}
ctx <- read_h5ad("~/Projects/2025_workshop/WMB-10Xv2-CTXsp-raw.h5ad")

ctx_obj = CreateSeuratObject(counts = t(as.matrix(ctx$X)), 
                             
  meta.data = ctx$obs, 
  
  min.features = 200, 
  
  min.cells = 3, 
  
  project = "CTX")

ctx_obj <- PercentageFeatureSet(ctx_obj, pattern = "^mt", col.name = "percent.mt")

hist(ctx_obj@meta.data$nFeature_RNA)

hist(ctx_obj@meta.data$nCount_RNA)

hist(ctx_obj@meta.data$percent.mt)

ctx_obj <- subset(ctx_obj, subset = nFeature_RNA > 500 &
                    
  nCount_RNA < 20000 &
    
  percent.mt < 5) 

ctx_obj <- NormalizeData(ctx_obj, 
                        
  normalization.method = "LogNormalize", 
  
  scale.factor = 10000)

ctx_obj <- FindVariableFeatures(ctx_obj, 
                                
  selection.method = "vst", 
  
  nfeatures = 2000)

ctx_obj <- ScaleData(ctx_obj, features = rownames(ctx_obj))

ctx_obj <- RunPCA(ctx_obj, features = VariableFeatures(object = ctx_obj))

ctx_obj <- FindNeighbors(ctx_obj, dims = 1:10)

ctx_obj <- FindClusters(ctx_obj, resolution = 0.1)

ctx_obj <- RunUMAP(ctx_obj, dims = 1:10)

DimPlot(ctx_obj)

ctx_obj_subset <- subset(ctx_obj, downsample = 200)

DimPlot(ctx_obj_subset)

saveRDS(ctx_obj_subset, "ctx_obj_downsampled.rds")
```

## Repeat with the MB object. 

```{r}
mb <- read_h5ad("~/Projects/2025_workshop/WMB-10Xv2-MB-raw.h5ad")

mb_obj <- CreateSeuratObject(counts = t(as.matrix(mb$X)), 
                             
  meta.data = mb$obs, 
  
  min.features = 200, 
  
  min.cells = 3, 
  
  project = "MB")

mb_obj <- PercentageFeatureSet(mb_obj, pattern = "^mt", col.name = "percent.mt")

hist(mb_obj@meta.data$nFeature_RNA)

hist(mb_obj@meta.data$nCount_RNA)

hist(mb_obj@meta.data$percent.mt)

mb_obj <- subset(mb_obj, subset = nFeature_RNA > 500 &
                   
  nCount_RNA < 20000 &
    
  percent.mt < 5)

mb_obj <- NormalizeData(mb_obj, 
                        
  normalization.method = "LogNormalize", 
  
  scale.factor = 10000)

mb_obj <- FindVariableFeatures(mb_obj, 
                              
  selection.method = "vst", 
  
  nfeatures = 2000)

mb_obj <- ScaleData(mb_obj, features = rownames(mb_obj))

mb_obj <- RunPCA(mb_obj, features = VariableFeatures(object = mb_obj))

mb_obj <- FindNeighbors(mb_obj,  dims = 1:10)

mb_obj <- FindClusters(mb_obj, resolution = 0.1)

mb_obj <- RunUMAP(mb_obj, dims = 1:10)

DimPlot(mb_obj)

mb_obj_subset <- subset(mb_obj, downsample = 200)

DimPlot(mb_obj_subset)

saveRDS(mb_obj_subset, "mb_obj_downsampled.rds")
```

## Repeat with the HY object. 

```{r}
hy <- read_h5ad("~/Projects/2025_workshop/WMB-10Xv2-HY-raw.h5ad")

hy_obj <- CreateSeuratObject(counts = t(as.matrix(hy$X)), 
                            
  meta.data = hy$obs, 
  
  min.features = 200, 
  
  min.cells = 3, 
  
  project = "HY")

hy_obj <- PercentageFeatureSet(hy_obj, pattern = "^mt", col.name = "percent.mt")

hist(hy_obj@meta.data$nFeature_RNA)

hist(hy_obj@meta.data$nCount_RNA)

hist(hy_obj@meta.data$percent.mt)

hy_obj <- subset(hy_obj, subset = nFeature_RNA > 500 &
                   
  nCount_RNA < 20000 &
    
  percent.mt < 5)

hy_obj <- NormalizeData(hy_obj, 
                        
  normalization.method = "LogNormalize", 
  
  scale.factor = 10000)

hy_obj <- FindVariableFeatures(hy_obj, 
                               
  selection.method = "vst", 
  
  nfeatures = 2000)

hy_obj <- ScaleData(hy_obj, features = rownames(hy_obj))

hy_obj <- RunPCA(hy_obj, 
                
  features = VariableFeatures(object = hy_obj))

hy_obj <- FindNeighbors(hy_obj, dims = 1:10)

hy_obj <- FindClusters(hy_obj, resolution = 0.1)

hy_obj <- RunUMAP(hy_obj, dims = 1:10)

DimPlot(hy_obj)

hy_obj_subset <- subset(hy_obj, downsample = 200)

DimPlot(hy_obj_subset)

saveRDS(hy_obj_subset, "hy_obj_downsampled.rds")
```

## Repeat with the TH specimen. 

```{r}
th <- read_h5ad("~/Projects/2025_workshop/WMB-10Xv2-TH-raw.h5ad")

th_obj <- CreateSeuratObject(counts = t(as.matrix(th$X)), 
                            
  meta.data = th$obs, 
  
  min.features = 200, 
  
  min.cells = 3, 
  
  project = "TH")

th_obj <- PercentageFeatureSet(th_obj, pattern = "^mt", col.name = "percent.mt")

hist(th_obj@meta.data$nFeature_RNA)

hist(th_obj@meta.data$nCount_RNA)

hist(th_obj@meta.data$percent.mt)

VlnPlot(th_obj, features <- c("percent.mt", "n_FeatureRNA", "n_CountRNA"))

th_obj <- subset(th_obj, subset = nFeature_RNA > 500 &
                   
  nCount_RNA < 30000 &
    
  percent.mt < 5)

th_obj <- NormalizeData(th_obj, 
                        
  normalization.method = "LogNormalize", 
  
  scale.factor = 10000)

th_obj <- FindVariableFeatures(th_obj, 
                              
  selection.method = "vst", 
  
  nfeatures = 2000)

th_obj <- ScaleData(th_obj, features = rownames(th_obj))

th_obj <- RunPCA(th_obj, features = VariableFeatures(object = th_obj))

th_obj <- FindNeighbors(th_obj, dims = 1:10)

th_obj <- FindClusters(th_obj, resolution = 0.1)

th_obj <- RunUMAP(th_obj, dims = 1:10)

DimPlot(th_obj)

th_obj_subset <- subset(th_obj, downsample = 200)

DimPlot(th_obj_subset)

saveRDS(th_obj_subset, "~/Projects/2025_workshop/th_obj_downsampled.rds")
```

## Read in thw downsampled objects and convert Ensembl IDs to Gene Symbols. 

```{r}
ctx_obj_subset <- readRDS("ctx_obj_downsampled.rds")

mb_obj_subset <- readRDS("mb_obj_downsampled.rds")

hy_obj_subset <- readRDS("hy_obj_downsampled.rds")

th_obj_subset <- readRDS("th_obj_downsampled.rds")
```

## Process the objects. 

```{r}
list_new <-

    list(ctx_obj_subset, mb_obj_subset, hy_obj_subset, th_obj_subset)

mart = useMart("ensembl")

mart = useDataset("mmusculus_gene_ensembl", mart)

new_list_obj <- list()

for (item in list_new) {

    matrix <-
      
        GetAssayData(item,

                layer = "counts")

    gene_symbol <-

        biomaRt::getBM(attributes = c("ensembl_gene_id", "mgi_symbol"),

                    filters = "ensembl_gene_id",

                    values = rownames(matrix),

                    mart = mart)

    matrix_new <-

        matrix %>%

        as.data.frame() %>%

        rownames_to_column(var = "ensembl_gene_id") %>%

        left_join(gene_symbol) %>%

        dplyr::group_by(mgi_symbol) %>%

        slice_head(n=1) %>%

        filter(!is.na(mgi_symbol)) %>%

        filter(mgi_symbol != "") %>%

        column_to_rownames(var = "mgi_symbol") %>%

        dplyr::select(-ensembl_gene_id) %>%

        as.matrix()

    seurat_object <-

        CreateSeuratObject(counts = matrix_new,

                        meta.data = item@meta.data)

    new_list_obj <-

        append(new_list_obj,

            list(seurat_object))    

}
```

## Save the objects. 

```{r}
saveRDS(new_list_obj, "additional_objects.rds")
```







