
## Library
```{r lib, cache = FALSE}
# Load necessary packages
library(Seurat)
library(tidyverse) # This package contains ggplot2 and dplyr
library(patchwork)
library(ggplot2) # For plotting
library(rrrSingleCellUtils) # For plotting an processing
library(qs) # For reading and writing .qs files
library(harmony) # For Harmony integration
# Set the random generator seed so that results are reproducible.
set.seed(132)
```



## load the objects to be merged
In this tutorial, we will demonstrate how to merge two single cell RNA-seq
datasets using the MergeSeurat function in Seurat. Merging single cell RNA-seq 
datasets is a crucial step in the analysis of single cell data. 
It allows researchers to combine data from different sources and 
increase the sample size for more robust and comparative analysis.
Let's say we have many different datasets from different samples or
experiments. We can load the datasets individually and then merge them using
the MergeSeurat function in Seurat. The merged dataset can then be used for
downstream analysis, such as dimensionality reduction, clustering, and
differential expression analysis.

```{r test}
# read in the pre-made Seurat objects to be merge
sobj1 <- 
    qs::qread("output/rdata_premade/OX1X_sobj_annotated.qs")

#check the object
sobj1

# read in the pre-made Seurat objects to be merge 
# this is a list of Seurat objects
sobj_list <-
    qs::qread("output/rdata_premade/scRNA_objs_list.qs")

#check the list
sobj_list

# merge the Seurat objects
merged_sobj <-
    merge(x= sobj1, 
          y = sobj_list[1:length(sobj_list)])

# check the merged Seurat object, you will see layers for each seurat object
merged_sobj

# join the layers so that they are all in the same object
merged_sobj <-
    merged_sobj %>%
    JoinLayers()

# check the object now, the layers are now joined or merged
merged_sobj

# need to re-run the normalization and scaling
merged_sobj <- 
    merged_sobj %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData()

# run PCA
merged_sobj <- 
    merged_sobj %>%
    RunPCA()

# run UMAP
merged_sobj <- 
    merged_sobj %>%
    RunUMAP(dims = 1:30)

# run clustering
merged_sobj <- 
    merged_sobj %>%
    FindNeighbors(dims = 1:30) %>%
    FindClusters(resolution = 0.5)
# check the object
merged_sobj

# check the metadata
DimPlot(merged_sobj,
              group.by = "cell_type",
              #split.by = "orig.ident",
              #ncol = 6,
              label = T) +
    NoLegend()

harm <-
    RunHarmony(merged_sobj,
              group.by.vars = "orig.ident")

# check the object
harm <-
    RunUMAP(harm,
            reduction = "harmony",
            dims = 1:30) %>%
    FindNeighbors(reduction = "harmony",
                  dims = 1:30) %>%
    FindClusters(resolution = 0.5)

DimPlot(harm,
              group.by = "cell_type",
              #split.by = "orig.ident",
              #ncol = 6,
              label = T) +
    NoLegend()
```