---
title: 'Seurat Lesson: Code'
author: "Emily Franz and Jack Hedberg"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  echo = TRUE,
  cache = TRUE,
  collapse = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE)
```

```{r lib, cache = FALSE}
# Load necessary packages
library(Seurat)
library(tidyverse) # This package contains ggplot2 and dplyr
library(patchwork)

# Set the random generator seed so that results are reproducible.
set.seed(132)
```


Link to download the example dataset we are using: https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz


# Load in scRNAseq dataset: 
***YOU WILL NEED TO MODIFY THE CODE BELOW SO THAT THE FOLDER PATH POINTS R TO 
WHEREVER THE RAW DATA IS LOCATED ON YOUR COMPUTER. Make sure that you include the 
necessary folders in the path to the below folders. 
Be careful with forward slashes. Folders always have a forward slash in front of 
them, and files do not.
```{r load}
# Load pbmc scRNAseq dataset
pbmc.data <- Read10X(data.dir = "pbmc3k_filtered_gene_bc_matrices/hg19")
```

# Create Seurat Object
```{r createobj}
pbmc <- CreateSeuratObject(counts = pbmc.data,
                           project = "pbmc3k",
                           min.cells = 3,
                           min.features = 200)

# We can now remove the previous object
rm(pbmc.data)
```

# Perform Subsetting for Quality Control
All subsetting will depend on your sample and its quality and conditions. For 
instance, if one sample is a tumor treated with effective chemotherapy, you may 
expect to see a higher percent of mitochondrial genes and may want to include 
more of these dying cells in your subsetting (say, 25% mitochondrial genes). 
All subsetting should depend on the spread of your data, with outliers (both 
high and low) typically targeted for exclusion.
**Feature subsetting:**
nFeature_RNA: number of genes (features) detected per cell. 
Subsetting the number of features per cells can help to remove doublets (two cells
that grouped together as "one" -> high nFeature_RNA) and non-cells, such as 
dead cells or free mRNA that grouped together (low nFeature_RNA).

**Count subsetting:**
nCount_RNA: number of unique molecule identifiers (UMIs) per cell
High nCount_RNA also indicates doublets, and so is another method for 
filtering out those doublets. Some prefer using this method to using nFeatures to remove 
doublets. Using both, one can be fairly confident of removing doublets. 

**Mitochondrial gene subsetting:**
Subsetting the data by percent mitochondrial genes helps remove poor quality 
cells, such as dead cells or cells with a broken membrane (this causes leakage of
mRNA). These cells have a higher percentage of mitochondrial genes. Typically, 
25% and lower is preferred; however, again, this depends on your data spread 
and your expectations for the sample. 
Note: "^MT-" = all genes beginning with "MT-", which is how all mitochondrial 
gene names begin.

**Ribosomal and hemoglobin gene subsetting:**
These are other options for identifying low quality samples based on the percent
of these genes in each cell. Ribosomal genes begin with either "RPS" or "RPL". 
Hemoglobin genes begin with either "HBA" or "HBB".
```{r qc}
#### Visualize Outliers ####
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc[["percent.ribo"]] <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]")
pbmc[["percent.hb"]] <- PercentageFeatureSet(pbmc, pattern = "^HB[AB]")

# Violin plot to visualize outliers
VlnPlot(pbmc,
        features = c("nFeature_RNA",
                     "nCount_RNA",
                     "percent.mt"))
VlnPlot(pbmc,
        features = c("percent.mt",
                     "percent.ribo",
                     "percent.hb"))

# Scatter plots to visualize outliers using combinations of criteria for subsetting
## For instance, cells with both high nCount and nFeature are likely doublets.
## Low nCount and high mitochondrial percentage can indicate dead/dying cells 
## where membrane leakage has occured.
plot1 <- FeatureScatter(pbmc,
                        feature1 = "nCount_RNA",
                        feature2 = "percent.mt") + 
  theme(legend.position="none")
plot2 <- FeatureScatter(pbmc,
                        feature1 = "nCount_RNA",
                        feature2 = "nFeature_RNA") + 
  theme(legend.position="none")
# Plots can be assigned to an object and then patchwork uses the below 
# to plot these side by side
plot1 + plot2

#### Subset ####
pbmc <- subset(pbmc,
               subset = nFeature_RNA > 300 &
                 nFeature_RNA < 2000 &
                 nCount_RNA < 8000 &
                 percent.mt < 10 &
                 percent.hb < 2)
```

# Normalize
Normalization is the process of adjusting gene expression values to eliminate/reduce 
the influence of technical effects impacting all genes the same way.
```{r normalize}
pbmc <- NormalizeData(pbmc,
                      normalization.method = "LogNormalize",
                      scale.factor = 10000)
```

# Identify Highly Variable Features
This function identifies the n most highly variable features. By default, n = 2000 
(defined by nfeatures=2000). This is used by the next step, scaling, to speed it
up. Often, many biologically important features will be captured in this step, 
however, you may lose important data, depending on your dataset. 
```{r varfeatures}
pbmc <- FindVariableFeatures(pbmc,
                             selection.method = "vst",
                             nfeatures = 2000)
```

# Scale Features
Scaling applies linear transformation. Importantly, this step ensures that 
highly-expressed genes do not dominate and drown out relatively lower-expressed 
genes that have biological importance. This is a standard pre-processing step 
prior to principle component analysis (dimensional reduction).
  -Shifts the expression of each gene (feature), so that the mean expression across cells is 0
  -Scales the expression of each gene (feature), so that the variance across cells is 1
  -Results are stored in pbmc[["RNA"]]@scale.data
  
Note that if the features parameter is undefined, this function will perform 
scaling on only 2,000 of the most variable features by default (may lose data).
```{r scale}
#### Scale data ####
# The rownames of the pbmc object are ALL of the genes (features).
pbmc <- ScaleData(pbmc,
                  features = rownames(pbmc))
```

# Perform Principle Component Analysis (PCA)
This step can also be known as linear dimensional reduction. Numerous methods 
exist for visualization of principle components (PCs). A common method is 
*ElbowPlot*, which plots the standard deviations of the principle components. 
The left of the 'elbow' tends to represent the significant dimensions. 

Other visualization methods include *JackStraw* (slow) and *DimHeatmap*. Note that 
higher dimensions can be chosen, especially if certain rare cell-type signatures 
are distinguishable in these PCs. *DimHeatmap* is helpful in seeing these genes.
More information can be found at: 
[Seurat - Guided Clustering Tutorial: Perform linear dimensional reduction](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html).
```{r pca}
#### Run PCA ####
pbmc <- RunPCA(pbmc,
               features = VariableFeatures(object = pbmc))

#### Elbow Plot ####
# Note the elbow around PC 9 or 10
# (more specifically, the PCs from above PC 10 show nearly no difference in y-axis 
# value to that of PC 10). Thus, we will proceed with dims = 1:10.
ElbowPlot(pbmc,
          ndims = 25)
```

# Cluster Single Cells.
This portion includes two steps. The first, *FindNeighbors*, takes the principle 
components selected above and constructs a graph based on the euclidean distance 
from the PCA and finds the similarities between two cells based on overlap of 
their local neighborhoods. The selected PCs are defined in the *dims* parameter. 
The second step, *FindClusters*, iteratively groups cells together with the ability 
to set the resolution, or granularity, of the clustering. The higher the 
resolution, the greater the number of clusters in the output.
```{r cluster}
#### Find Neighbors ####
pbmc <- FindNeighbors(pbmc,
                      dims = 1:10)

#### Find Clusters ####
# Low resolution (low number of clusters)
pbmc <- FindClusters(pbmc,
                     resolution = 0.1)
```

# Run Non-linear Dimensional Reduction (UMAP/tSNE)
Using the same PCs and resolution as defined above, this step places similar 
cells together in a low-dimension (2D) space. Those cells calculated as co-localized 
in the clusters generated above should co-localize in the UMAP or tSNE space. 
Cells are colored by their cluster or identity class.

As for differences between tSNE and UMAP, UMAP lowers the dimensions of the high 
dimensional graph from above using compression, whereas t-SNE moves the high 
dimensional graph to a lower dimensional space points by points. Their parameters,
thus, differ, but both can be used as graph-based methods for non-linear 
dimensional reduction. Neither is perfect, but UMAP tends to provide better balance
between local versus global structure and is more time-effective. For more information, 
please see [Comparing UMAP vs t-SNE in Single-cell RNA-Seq Data Visualization, Simply Explained](https://blog.bioturing.com/2022/01/14/umap-vs-t-sne-single-cell-rna-seq-data-visualization/) 
and [Understanding UMAP](https://pair-code.github.io/understanding-umap/).

*RunTSNE()* is used, similarly to below, to acheive a TSNE plot.
```{r umap}
#### Run UMAP ####
# Use the same dimensions as those in FindNeighbors()
pbmc <- RunUMAP(pbmc,
                dims = 1:10)

#### Plot UMAP ####
DimPlot(pbmc,
        reduction = "umap")

# Visualize cluster labels on UMAP
## Note: reduction parameter, if not specified, first searches for umap, then tsne, then pca
## By default, then, it will select whichever dimensional reduction method is used above,
## unless both umap and tsne have been generated. 
DimPlot(pbmc,
        label = TRUE,
        repel = TRUE)
```

# Practice 1
Generate a dimplot with a higher clustering resolution of 0.5. (You do not need 
to rerun anything prior to this clustering section).

```{r practice1}

```

# Differentially Expressed Genes
*Tomorrow's lesson will be devoted to exploring this topic in more depth.*

*FindAllMarkers* finds all the genes that make each cluster most different from 
all of the other clusters. If you want to look at the genes that make cluster (A)
most different from another cluster (B), you can use *FindMarkers* and set ident.1 
equal to the cluster of interest (A) and ident.2 equal to its comparison (B).

```{r idents}
# Make sure the identities are what you want (clustering, cell ids, etc.)
# You can find current identity labels using:
levels(pbmc)
head(Idents(pbmc))

# You can access the metadata labels using "$" (tab to see meta.data options)
# Here, we choose our lower resolution clustering assignments from above
Idents(pbmc) <- pbmc$RNA_snn_res.0.1
```

```{r deg}
#### FindAllMarkers ####
# Find differential markers for every cluster compared to all remaining cells
# Report only the positive values and set a log-fold-change threshold of 0.25
# Note that different statistical tests can be used; default is wilcox
pbmc.markers <- FindAllMarkers(pbmc,
                               only.pos = TRUE,
                               logfc.threshold = 0.25)
#### View Markers ####
pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 5,
              order_by = avg_log2FC)
```

# Feature Plotting Techniques 
Seurat and other packages offer numerous methods for plotting genes of interest.
Here we present some of the most popular Seurat-based methods for plotting.

```{r idents-repeated}
# Make sure the identities are what you want (clustering, cell ids, etc.)
# You can find current identity labels using:
levels(pbmc)
head(Idents(pbmc))

# You can access the metadata labels using "$" (tab to see meta.data options)
# Here, we choose our lower resolution clustering assignments from above
Idents(pbmc) <- pbmc$RNA_snn_res.0.1
```

## Feature Plot
Colors single cells on a dimensional reduction plot by feature (typically gene expression level,
but could be  PC scores, number of genes detected, module score, etc.)
```{r feature}
#### Default FeaturePlot ####
# Plots each feature separately
FeaturePlot(pbmc,
            features = c("MS4A1",
                         "NKG7",
                         "CD3E",
                         "LYZ"),
            label = TRUE,
            repel = TRUE)
```

```{r rename}
# Change cluster names - add identities
pbmc <- RenameIdents(pbmc,
                     "0" = "T cells",
                     "1" = "Monocytes",
                     "2" = "NK cells",
                     "3" = "B cells")
# Stash/save current identities into the metadata under a name
pbmc <- StashIdent(pbmc,
                   save.name = 'cellid')

# Check under the new metadata category name
levels(pbmc$cellid)

# Plot with new names (uses our active.ident, which is currently 'cellid' labels)
DimPlot(pbmc,
        label = TRUE,
        pt.size = 0.5) +
  NoLegend()
```

```{r}
#### Split.by ####
# split.by plots each feature in each identity separately (labels would still
# be defined as the default, or that specified in group.by)
## very helpful for different treatment groups
FeaturePlot(pbmc,
            features = c("MS4A1",
                         "NKG7"),
            split.by = "cellid")

# Plots by the defined groups even if the active identity is a different
Idents(pbmc) <- pbmc$RNA_snn_res.0.1
FeaturePlot(pbmc,
            features = c("MS4A1",
                         "NKG7"),
            split.by = "cellid",
            label = T)
```

```{r}
# Cutoffs help visualize features if signal too weak or too strong in certain cells
# Note that the expression units are arbitrary units, specific for each feature
FeaturePlot(pbmc,
            features = "LYZ")

#### Max cutoff ####
FeaturePlot(pbmc,
            features = "LYZ",
            max.cutoff = 4)

#### Min cutoff ####
FeaturePlot(pbmc,
            features = "LYZ",
            min.cutoff = 2)
```

```{r}
#### Change plot colors ####
# (note that a linear color scale works best)
FeaturePlot(pbmc,
            features = "LYZ",
            cols = c("gray", "red"))

```

## Violin Plot

```{r vln}
#### Default VlnPlot ####
# Each feature is plotted separately
# violins are colored by feature (can change to color by identity with fill.by)
VlnPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"))
```

```{r}
#### Group.by and RotateAxis() ####
# group.by allows expression to be shown for identity classes of interest
# RotatedAxis() rotates x-axis labels
VlnPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"),
        group.by = "cellid") +
  RotatedAxis()
```

## Ridge Plot

```{r ridge}
### Default ridgeplot ####
# plots by feature
RidgePlot(pbmc,
          features = c("MS4A1",
                       "NKG7",
                       "CD3E",
                       "LYZ"))
```

```{r}
#### Group.by, fill.by, sort ####
# Can group by certain identities, sort by average expression, and more
RidgePlot(pbmc,
          features = c("MS4A1",
                       "NKG7",
                       "CD3E",
                       "LYZ"),
          group.by = "cellid",
          sort = TRUE,
          fill.by = "feature") # fill.by is also an option in VlnPlot()
```

## DotPlot
Be careful using this plotting method - it works to emphasize differences and 
therefore may appear misleading (visual differences are exaggerated)

```{r dot}
#### Default DotPlot ####
DotPlot(pbmc,
        features = c("MS4A1",
                       "NKG7",
                       "CD3E",
                       "LYZ"),
          group.by = "cellid")
```

```{r}
#### Remove plot scaling ####
# (less exaggerated, more representative of true differences)
DotPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"),
        group.by = "cellid",
        scale = FALSE)
```

```{r}
#### Change Dotplot colors ####
# (note that a linear color scale works best)
DotPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"),
        group.by = "cellid",
        scale = FALSE,
        cols = c("gray", "red"))
```

# Practice 2
Change your active identities to your resolution of 0.5 (RNA_snn_res.0.5).
Plot the expression of "CD8A" to identify CD8 T cells using a feature plot 
(label cluster numbers on plot), violin plot (group by cellid and rotate the x-axis 
labels), ridgeplot, and dotplot (not scaled).

```{r practice2}

```

# Saving Seurat Objects

```{r seuratsave}
# Save
save(pbmc,
     file = "Data/pbmc.RData")
# saveRDS
saveRDS(pbmc,
     file = "Data/pbmc.rds")
```

# Loading Seurat Objects

```{r seuratload}
# Load
load("Data/pbmc.RData")
# readRDS
pbmc <- readRDS("Data/pbmc.rds")
```

# Bonus: Cell Type Annotation
There are many methods of cell type annotation. Here we show a reference-
based technique, *SingleR*. A reference based technique uses a
list of genes associated with a certain cell type to suggest cell identities.
A simpler, but often time-consuming approach, is to analyze the top 
differentially expressed genes for each cluster (shown above) or certain unique 
genes your lab or others have used to identify specific cell types, if 
applicable (see techniques for feature plotting). Note: this is not from 
the Seurat package.

## SingleR
```{r singler}
# Automatic prediction of cell types using SingleR
library(SingleCellExperiment)
library(celldex)
library(SingleR)

# Convert the Seurat object to a SCE object
pbmc_sce <- as.SingleCellExperiment(pbmc)

# Load human reference of interest from celldex
## Mouse reference & immune-specific mouse reference are helpful for mouse data.
## ref_mouse <- celldex::MouseRNAseqData()
## ref_mouse_immune <- celldex::ImmGenData()
ref1 <- celldex::HumanPrimaryCellAtlasData()

# Make cell type predictions using SingleR
## Note that you can use ref1$label.fine for more specific predictions
predictions <- SingleR(test = pbmc_sce,
                       ref = list(ref1),
                       labels = list(ref1$label.main))

# Clean up environment
rm(pbmc_sce, ref1)
```

```{r singler_cont, dependson= "singler"}
# Transfer labels back to the Seurat object metadata and plot
pbmc$singleR <- predictions$labels

# Label cells with singlr label
Idents(pbmc) <- pbmc$singleR 

# Plot on UMAP
## Note the use of group.by to label cells by certain identifiers (singleR here)
## Seurat uses ggplot, so plot titles (and more) can be edited with ggplot2 functions
## NoLegend() (Seurat) removes DimPlot legend, useful if plot is labeled

DimPlot(pbmc,
        reduction = "umap",
        group.by = "singleR",
        label = T,
        repel = T) +
  ggtitle("Human Reference Annotation") +
  NoLegend()
```

# Challenge:
Perform single cell analysis of a human bone marrow mononuclear cell dataset (10x): 

## Load in bone marrow dataset
```{r} 
datasetbone_marrow_data <- Read10X(data.dir = "Challenge Dataset/hg19") 
```

## Create Seurat Object
```{r} 

```

## Perform Subsetting for Quality Control
```{r qc} 

```

## Normalize
```{r normalize} 

```

## Identify Highly Variable Features
```{r varfeatures} 

```

## Scale Data
```{r} 

```

## Run PCA
```{r pca} 

```

## Find Neighbors and Find Clusters
```{r cluster} 

```

## Run UMAP
```{r umap} 

```

## FindMarkers of Clusters
```{r} 

```

## DotPlot:
Dot plot showing cluster-wise expression of selected genes:
MS4A1, NKG7, CD3E, LYZ, CD4
```{r} 

```

## FeaturePlot:
Feature Plots showing expression of selected genes:
MS4A1, NKG7, CD3E, LYZ
```{r}

```

## Bonus: Let's make a Nebulosa density plot!
Density Plot (Nebulosa Package) Showing Density of LYZ Expression

Use this code to install Nebulosa and get started
```{r} 
if (!require("BiocManager", quietly = TRUE)) 
  install.packages("BiocManager") 
BiocManager::install("Nebulosa") 
library(Nebulosa) 

#plot_density(seurat object, "GENE SYMBOL") 
#above is an example of how to use the density plot function from the Nebulosa package 
#Below, make your own density plot of LYZ
 
```

```{r}
sessionInfo()
```

