---
title: 'Seurat Lesson: Code'
author: "Emily Franz and Jack Hedberg"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  echo = TRUE,
  cache = TRUE,
  collapse = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE)
```

```{r lib, cache = FALSE}
# Load necessary packages
library(Seurat)
library(tidyverse) # This package contains ggplot2 and dplyr
library(patchwork)

# Set the random generator seed so that results are reproducible.
set.seed(132)
```

# Load in scNRAseq dataset
```{r load}
# Load pbmc scRNAseq dataset


```

# Perform Subsetting for Quality Control
```{r qc}

```

# Normalize
```{r normalize}

```

# Identify Highly Variable Features
```{r varfeatures}

```

# Scale Features
Scaling applies linear transformation. Importantly, this step ensures that 
highly-expressed genes do not dominate and drown out relatively lower-expressed 
genes that have biological importance. This is a standard pre-processing step 
prior to principle component analysis (dimensional reduction).
  -Shifts the expression of each gene (feature), so that the mean expression across cells is 0
  -Scales the expression of each gene (feature), so that the variance across cells is 1
  -Results are stored in pbmc[["RNA"]]@scale.data
```{r scale}
#### Scale data ####
# The rownames of the pbmc object are ALL of the genes (features).
# Note that if the features parameter is undefined, this function will perform 
# scaling on only 2,000 of the most variable features by default (may lose data).
pbmc <- ScaleData(pbmc,
                  features = rownames(pbmc))
```

# Perform Principle Component Analysis (PCA)
This step can also be known as linear dimensional reduction. Numerous methods 
exist for visualization of principle components (PCs). A common method is 
*ElbowPlot*, which plots the standard deviations of the principle components. 
The left of the 'elbow' tends to represent the significant dimensions. 
Other visualization methods include *JackStraw* (slow) and *DimHeatmap*. Note that 
higher dimensions can be chosen, especially if certain rare cell-type signatures 
are distinguishable in these PCs. *DimHeatmap* is helpful in seeing these genes.
More information can be found at: 
[Seurat - Guided Clustering Tutorial: Perform linear dimensional reduction](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html).
```{r pca}
#### Run PCA ####
pbmc <- RunPCA(pbmc,
               features = VariableFeatures(object = pbmc))

#### Elbow Plot ####
# Note the elbow around PC 9 or 10
# (more specifically, the PC 11 point shows nearly no difference in y-axis 
# value to that of PC 10). Thus, we will proceed with dims = 1:10.
ElbowPlot(pbmc,
          ndims = 25)
```

# Cluster Single Cells.
This portion includes two steps. The first, *FindNeighbors*, takes the principle 
components selected above and constructs a graph based on the euclidean distance 
from the PCA and finds the similarities between two cells based on overlap of 
their local neighborhoods. The selected PCs are defined in the *dims* parameter. 
The second step, *FindClusters*, iteratively groups cells together with the ability 
to set the resolution, or granularity, of the clustering. The higher the 
resolution, the greater the number of clusters in the output.
```{r cluster}
#### Find Neighbors ####
pbmc <- FindNeighbors(pbmc,
                      dims = 1:10)

#### Find Clusters ####
pbmc <- FindClusters(pbmc,
                     resolution = 0.5)
```

# Run Non-linear Dimensional Reduction (UMAP/tSNE)
Using the same PCs and resolution as defined above, this step places similar 
cells together in a low-dimension (2D) space. Those cells calculated as co-localized 
in the clusters generated above should co-localize in the UMAP or tSNE space. 
Cells are colored by their cluster or identity class.

As for differences between tSNE and UMAP, UMAP lowers the dimensions of the high 
dimensional graph from above using compression, whereas t-SNE moves the high 
dimensional graph to a lower dimensional space points by points. Their parameters,
thus, differ, but both can be used as graph-based methods for non-linear 
dimensional reduction. Neither is perfect, but UMAP tends to provide better balance
between local versus global structure and is more time-effective. For more information, 
please see [Comparing UMAP vs t-SNE in Single-cell RNA-Seq Data Visualization, Simply Explained](https://blog.bioturing.com/2022/01/14/umap-vs-t-sne-single-cell-rna-seq-data-visualization/) 
and [Understanding UMAP](https://pair-code.github.io/understanding-umap/).
```{r umap}
#### Run UMAP ####
# Use the same dimensions as those in FindNeighbors()
pbmc <- RunUMAP(pbmc,
                dims = 1:10)

#### Plot UMAP ####
DimPlot(pbmc,
        reduction = "umap")

# Visualize cluster labels on UMAP
## Note: reduction parameter, if not specified, first searches for umap, then tsne, then pca
## By default, then, it will select whichever dimensional reduction method is used above,
## unless both umap and tsne have been generated. 
DimPlot(pbmc,
        label = TRUE,
        repel = TRUE)
```

# Cell Type Annotation
There are many methods of cell type annotation. Here we show a reference-
based technique, *SingleR*. A reference based technique uses a
list of genes associated with a certain cell type to suggest cell identities.
A simpler, but often time-consuming approach, is to analyze the top 
differentially expressed genes for each cluster (shown here) or certain unique 
genes your lab or others have used to identify specific cell types, if 
applicable (see below techniques for feature plotting).

## SingleR
```{r singler}
# Automatic prediction of cell types using SingleR
library(SingleCellExperiment)
library(celldex)
library(SingleR)

# Convert the Seurat object to a SCE object
pbmc_sce <- as.SingleCellExperiment(pbmc)

# Load human reference of interest from celldex
## Mouse reference & immune-specific mouse reference are helpful for mouse data.
## ref_mouse <- celldex::MouseRNAseqData()
## ref_mouse_immune <- celldex::ImmGenData()
ref1 <- celldex::HumanPrimaryCellAtlasData()

# Make cell type predictions using SingleR
## Note that you can use ref1$label.fine for more specific predictions
predictions <- SingleR(test = pbmc_sce,
                       ref = list(ref1),
                       labels = list(ref1$label.main))

# Clean up environment
rm(pbmc_sce, ref1)
```

```{r singler_cont, dependson= "singler"}
# Transfer labels back to the Seurat object metadata and plot
pbmc$singleR <- predictions$labels

# Label cells with singlr label
Idents(pbmc) <- pbmc$singleR 

# Plot on UMAP
## Note the use of group.by to label cells by certain identifiers (singleR here)
## Seurat uses ggplot, so plot titles (and more) can be edited with ggplot2 functions
## NoLegend() (Seurat) removes DimPlot legend, useful if plot is labeled
DimPlot(pbmc,
        reduction = "umap",
        group.by = "singleR",
        label = T,
        repel = T) +
  ggtitle("Human Reference Annotation") +
  NoLegend()
```

## Differentially Expressed Genes
*Tomorrow's lesson will be devoted to exploring this topic in more depth.*

*FindAllMarkers* finds all the genes that make each cluster most different from 
all of the other clusters. If you want to look at the genes that make cluster (A)
most different from another cluster (B), you can use *FindMarkers* and set ident.1 
equal to the cluster of interest (A) and ident.2 equal to its comparison (B).
```{r deg}
#### FindAllMarkers ####
# Find differential markers for every cluster compared to all remaining cells
# Report only the positive values and set a log-fold-change threshold of 0.25
# Note that different statistical tests can be used; default is wilcox
pbmc.markers <- FindAllMarkers(pbmc,
                               only.pos = TRUE,
                               logfc.threshold = 0.25)
#### View Markers ####
pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 5,
              order_by = avg_log2FC)
```

# Feature Plotting Techniques 
Seurat and other packages offer numerous methods for plotting genes of interest.
Here we present some of the most popular Seurat-based methods for plotting.

## Feature Plot
Colors single cells on a dimensional reduction plot by feature (typically gene expression level,
but could be  PC scores, number of genes detected, module score, etc.)
```{r feature}
#### Default FeaturePlot ####
# Plots each feature separately
FeaturePlot(pbmc,
            features = c("MS4A1",
                         "NKG7",
                         "CD3E",
                         "LYZ"))
#### Group.by ####
# group.by plots each feature separately, but labels by identities of interest
FeaturePlot(pbmc,
            features = c("MS4A1",
                         "NKG7",
                         "CD3E",
                         "LYZ"),
            group.by = "singleR",
            label = TRUE,
            repel = TRUE)

#### Split.by ####
# split.by plots each feature in each identity separately (labels would still
# be defined as the default, or that specified in group.by)
## very helpful for different treatment groups
FeaturePlot(pbmc,
            features = c("MS4A1",
                         "NKG7"),
            split.by = "singleR")

# Cutoffs help visualize features if signal too weak or too strong in certain cells
# Note that the expression units are arbitrary units, specific for each feature
#### Max cutoff ####
FeaturePlot(pbmc,
            features = "LYZ",
            max.cutoff = 4)

#### Min cutoff ####
FeaturePlot(pbmc,
            features = "LYZ",
            min.cutoff = 2)


#### Change plot colors ####
# (note that a linear color scale works best)
FeaturePlot(pbmc,
            features = "LYZ",
            cols = c("gray", "red"))

```

## Violin Plot

```{r vln}
#### Default VlnPlot ####
# Each feature is plotted separately
# violins are colored by feature (can change to color by identity with fill.by)
VlnPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"))

#### Group.by and RotateAxis() ####
# group.by allows expression to be shown for identity classes of interest
# RotatedAxis() rotates x-axis labels
VlnPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"),
        group.by = "singleR") +
  RotatedAxis()
```

## Ridge Plot

```{r ridge}
### Default ridgeplot ####
# plots by feature
RidgePlot(pbmc,
          features = c("MS4A1",
                       "NKG7",
                       "CD3E",
                       "LYZ"))

#### Group.by, fill.by, sort ####
# Can group by certain identities, sort by average expression, and more
RidgePlot(pbmc,
          features = c("MS4A1",
                       "NKG7",
                       "CD3E",
                       "LYZ"),
          group.by = "singleR",
          sort = TRUE,
          fill.by = "ident") # fill.by is also an option in VlnPlot()
```

## DotPlot
Be careful using this plotting method - it works to emphasize differences and 
therefore may appear misleading (visual differences are exaggerated)
```{r dot}
#### Default DotPlot ####
DotPlot(pbmc,
        features = c("MS4A1",
                       "NKG7",
                       "CD3E",
                       "LYZ"),
          group.by = "singleR")

#### Remove plot scaling ####
# (less exaggerated, more representative of true differences)
DotPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"),
        group.by = "singleR",
        scale = FALSE)

#### Change Dotplot colors ####
# (note that a linear color scale works best)
DotPlot(pbmc,
        features = c("MS4A1",
                     "NKG7",
                     "CD3E",
                     "LYZ"),
        group.by = "singleR",
        scale = FALSE,
        cols = c("gray", "red"))
```

# Saving Seurat Objects

```{r save}

```

# Loading Seurat Objects

```{r load}

```

# Adding/Modifying Metadata

```{r metadata}
# The [[ operator can add columns to object metadata.
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc,
                                             pattern = "^MT-")
```

```{r}
sessionInfo()
```

