---
title: "Coder Upgrade 2026: Spatial Analysis"
author: "Corinne Strawser"
date: "2026-05-26"
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 5
        number_sections: false
        code_fmaleing: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      cache.lazy = FALSE)

# load libraries
library(Seurat)
library(tidyverse)
library("hdf5r")
library(patchwork)
library(qs2)
library(SPOTlight)
library(spacexr)
library(fgsea)
library(msigdbr)
library(data.table)

# set random generator seed to facilitate reproducibility
addTaskCallback(function(...) {set.seed(999);TRUE})

# load some colors
colorset <-
  c("#0166d8", "#ec102f", "#3aa609", "#6524ff", "#f8d147", "#5ce8ef", "#fd6ca0",
    "#476ba5", "#fea27a", "#add465", "#b775e1", "#fe8f06", "#5cfdc4", "#feafda",
    "#00228b", "#fd5925", "#166d2a", "#a90aa1", "#f3c011", "#5ac7dc", "#c1c2f5",
    "#1f29a1", "#e13224", "#86ec5a", "#672396", "#e1c637", "#069668", "#fcb2c7",
    "#1288da", "#c26a6b", "#6e9d50", "#1e0e76", "#fea53b", "#15729c", "#ffa8ff",
    "#270fe2", "#8a1341", "#285d28", "#956bed", "#f79302", "#116966", "#d945c2",
    "#94a6fd", "#e13219", "#01c472", "#afc6fe", "#e3c60b", "#aedfca", "#c2639f",
    "#0e2450", "#6d1a44", "#91fc40", "#5858e1", "#fca283", "#1eefc9", "#fb2076", 
    "#8bd0eb")

```

# Session Description

This session will provide an introduction to 10X Visium spatial transcriptomics analysis with Seurat. We will also cover experimental design considerations and adaptability of this code to other spatial technologies.

# Session Learning Objectives

After completion, you should be able to load 10X Visium spatial transcriptomics data into Seurat, perform normalization and unsupervised clustering, identify spatially-defined tissue domains, integrate with scRNA-seq data (spot deconvolution), subset out regions of interest (ROI), and perform differential expression and gene set enrichment analyses. 

# Section 1: Overview (advanced_tuesday.pptx)

* What is spatial transcriptomics?
* Benefits of spatial transcriptomics
* Challenges of spatial transcriptomics
* 10X Visium Technology
* 10X Visium Data Requirements
* 10X Visium Space Ranger Output
* Today's Demo Data

--------------------------------------------------------------------------------

# Section 2: Load 10X Visium data

Visium data can easily be loaded into Seurat using the spaceranger output directory. Load the B slice from the female mouse.

```{r load_female_data}
# load data
W3_F_Kidney <-
  Load10X_Spatial(data.dir = "~/workshop_data/raw/GSE252772/GSM8704839_20201201-NMK-F-Fc1U4Z1B1",
                  slice = "W3_F_Kidney")

# look at data
View(W3_F_Kidney)
```

How is the Visium data organized in the Seurat object?

* Spatial Assay (S4 object of class Assay5)
  - layers: counts (cells x features)
* Metadata (data frame)
* Image with our given slice name (S4 object of class VisiumV1)
  - image: hires or lowres image supplied to spaceranger
  - scale.factors: scale factors used for mapping pixel positions 
  - coordinates: tissue (1 = inside, 0 = outside), array position (col, row) and pixel position (imagecol, imagerow)
  
--------------------------------------------------------------------------------

# Section 3: QC

Are there differences in QC parameters based on anatomical differences?

```{r qc_plots}
# add % of reads mapping to mitochondrial genes as a measure of cell health
W3_F_Kidney[["percent.mt"]] <-
  PercentageFeatureSet(object = W3_F_Kidney, 
                       pattern = "^Mt\\d")
head(W3_F_Kidney)

# add mouse age to meta data
W3_F_Kidney$age <- "w3"

# add sex to meta data
W3_F_Kidney$sex <- "female"

# make QC violin plots
VlnPlot(object = W3_F_Kidney,
        features = c("nCount_Spatial",
                     "nFeature_Spatial",
                     "percent.mt"))

# make QC spatial plots
SpatialFeaturePlot(object = W3_F_Kidney,
                   features = c("nCount_Spatial",
                                "nFeature_Spatial",
                                "percent.mt"),
                   pt.size.factor = 3000,
                   image.alpha = 0)
```


SCTransform takes into account variation in number of molecules detected within each spot. Standard log normalization forces each spot to have the save underlying size after normalization. Since there may be variation in the amount of transcription occuring in certain regions of the tissue, I prefer to use log normalization.

--------------------------------------------------------------------------------

# Section 4: Normalization, dimensionality reduction, and clustering 

```{r female_normalization}
W3_F_Kidney <-
  W3_F_Kidney %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData(vars.to.regress = "percent.mt") 
W3_F_Kidney_clustered <-
  RunPCA(object = W3_F_Kidney,
         npcs = 30) %>%
  RunUMAP(dims = 1:30) %>%
  FindNeighbors(k.parm = 30) %>%
  FindClusters(resolution = 0.5)
```

# Section 5: Visualization

```{r female_vis}
# set a color palette
colors <-
  colorset[1:length(levels(W3_F_Kidney_clustered))] 
names(colors) <-
  levels(W3_F_Kidney_clustered)

# plot clusters as a umap
DimPlot(object = W3_F_Kidney_clustered,
        cols = colors)

# plot as a spatial plot
a <- 
  SpatialDimPlot(object = W3_F_Kidney_clustered,
                 cols = colors,
                   pt.size.factor = 3000,
                   image.alpha = 0) &
  ggtitle("Default")

# adjust point size parameter
b <- 
  SpatialDimPlot(object = W3_F_Kidney_clustered,
                 cols = colors,,
                   pt.size.factor = 5000,
                   image.alpha = 0) &
  ggtitle("pt.size.factor")

# adjust image alpha parameter
c <- 
  SpatialDimPlot(object = W3_F_Kidney_clustered,
                 cols = colors,
                   pt.size.factor = 3000,
                   image.alpha = 0.5) &
  ggtitle("image.alpha")

# adjust spot alpha parameter
d <-
  SpatialDimPlot(object = W3_F_Kidney_clustered,
                 cols = colors,
                 alpha = 0.4,
                   pt.size.factor = 3000,
                   image.alpha = 0) &
  ggtitle("alpha")

(a+b)/(c+d)
```

--------------------------------------------------------------------------------

# Activity 1:

Let's look at the same slice from a male mouse

## Load in a male mouse and add a column to the meta data indicating it's from a male mouse and what age it is

```{r load_male_data}

```

## Does the male kidney have a similar distribution of reads and percent of molecules mapping to mitochondrial genes?

```{r qc_male}

```

## Perform normalization, dimensionality reduction, and clustering.

```{r male_normalization}

```

--------------------------------------------------------------------------------

# Sidebar: There are many other ways to cluster your spatial data! 

Some other popular methods you might want to look at:
1. BANKSY: https://doi.org/10.1038/s41588-024-01664-3 & https://github.com/prabhakarlab/Banksy
2. BayesSpace: https://doi.org/10.1038/s41587-021-00935-2 & https://github.com/edward130603/BayesSpace
3. GraphST: https://doi.org/10.1038/s41467-023-36796-3 & https://github.com/JinmiaoChenLab/GraphST

--------------------------------------------------------------------------------

# Section 6: Identify spatially variable features using cluster identities

```{r spatial_features}
# make a vector of genes we're interested in
bad_genes <-
  c(rownames(W3_F_Kidney_clustered)[str_ends(string = rownames(W3_F_Kidney_clustered), 
                                          pattern = "Rik")],
    rownames(W3_F_Kidney_clustered)[str_starts(string = rownames(W3_F_Kidney_clustered),
                                            pattern = "Gm[:digit:]")])
genes_we_like <-
  rownames(W3_F_Kidney_clustered)[!rownames(W3_F_Kidney_clustered) %in% bad_genes]

# perform DE analysis on clusters
Idents(W3_F_Kidney_clustered) <- "seurat_clusters"
DefaultAssay(W3_F_Kidney_clustered) <- "Spatial"
cluster_markers <-
  FindAllMarkers(object = W3_F_Kidney_clustered,
                 only.pos = TRUE,
                 features = genes_we_like,
                 min.pct = 0.1,
                 logfc.threshmale = 0.2)

# make a dot plot of top 3 genes (by adjusted p value) per cluster
top3_genes <-
  cluster_markers %>%
  filter(p_val_adj <= 0.05) %>%
  group_by(cluster) %>%
  slice_min(order_by = p_val_adj,
            n = 50) %>%
  slice_max(order_by = avg_log2FC,
            n = 3) %>%
  pull(gene) %>%
  unique()
DotPlot(object = W3_F_Kidney_clustered,
        features = top3_genes) &
  scale_color_continuous(type = "viridis") &
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   vjust = 1))

# make a spatial feature plot of top gene (by adjusted p value and avg_log2FC) per cluster
top1_gene <-
  cluster_markers %>%
  filter(p_val_adj <= 0.05) %>%
  group_by(cluster) %>%
  slice_min(order_by = p_val_adj,
            n = 50) %>%
  slice_max(order_by = avg_log2FC,
            n = 1) %>%
  pull(gene) %>%
  unique()
SpatialFeaturePlot(object = W3_F_Kidney_clustered,
                   features = top1_gene,
                   pt.size.factor = 3000,
                   image.alpha = 0)
```

```{r}
rm(W3_F_Kidney_clustered, W3_M_Kidney_clustered)
gc()
```

--------------------------------------------------------------------------------

# Section 7: Merge male and female mouse data

```{r merge_data}
# make a list of the data to integrate
obj_list <-
  list("W3_F_Kidney" = W3_F_Kidney,
       "W3_M_Kidney" = W3_M_Kidney)

# merge normalized, but not clustered objects
spatial <-
  merge(x = obj_list[[1]],
        y = obj_list[2:length(obj_list)],
        add.cell.ids = names(obj_list))

View(spatial)

# notice we have two layers called counts in the spatial assay and two models in the RNA assay 

# select variable genes from both objects by ranking the features by the number of datasets they are deemed variable in, breaking ties by the median variable feature rank across datasets
var_feats <-
  SelectIntegrationFeatures(object = obj_list,
                            nfeatures = 4000)
VariableFeatures(object = spatial) <- var_feats

rm(obj_list, W3_F_Kidney, W3_M_Kidney)
gc()

# perform dimensionality reduction and clustering
DefaultAssay(object = spatial) <- "Spatial"
spatial <- 
  spatial %>%
  NormalizeData() %>%
  ScaleData() %>%
  RunPCA(npcs = 30) %>%
  RunUMAP(dims = 1:30) %>%
  FindNeighbors(k.parm = 30) %>%
  FindClusters(res = 0.6)

# join Spatial data layers
spatial <- 
  JoinLayers(object = spatial,
             assay = "Spatial")

# what do our clusters look like now?
cluster_colors <-
  colorset[1:length(levels(spatial$seurat_clusters))]
names(cluster_colors) <-
  levels(spatial$seurat_clusters)
SpatialDimPlot(object = spatial,
               group.by = "seurat_clusters",
               cols = cluster_colors,
               pt.size.factor = 3000,
               image.alpha = 0)

# do we have any batch effects?
DimPlot(object = spatial,
        group.by = c("sex","seurat_clusters"),
        cols = c(c("male" = "#4deeea",
                   "female" = "#fb2076"),
                 cluster_colors))

# yup
```

--------------------------------------------------------------------------------

# Section 8: Perform spot deconvolution

```{r deconvolution}
head(spatial@meta.data)

# load in reference data
ref <-
  readRDS("~/workshop_data/rdata/tabula_muris_senis_kidney.rds")

# take a look at the data
head(ref@meta.data)
DimPlot(object = ref,
        group.by = "cell_type")

# spacexr doesn't like labels with "/" in it, so let's change that
ref$cell_type <-
  gsub(pattern = "/", 
       replacement = "-", 
       x = ref$cell_type)

# create the spacexr reference object
spacexr_reference <-
    spacexr::Reference(counts = GetAssayData(object = ref, 
                                             layer = "counts"),
                       cell_types = as.factor(ref$cell_type))

rm(ref)
gc()

# get x and y coordinates for our visium data
coords <- 
  rbind(GetTissueCoordinates(object = spatial, 
                             image = "W3_F_Kidney"),
        GetTissueCoordinates(object = spatial, 
                             image = "W3_M_Kidney")) %>%
  select(-cell)

# create the spacexr data object
my_data <- 
  spacexr::SpatialRNA(coords = coords,
                      counts = GetAssayData(object = spatial, 
                                            layer = "counts",
                                            assay = "Spatial"))

# create the rctd object 
rctd_obj <- 
  spacexr::create.RCTD(spatialRNA = my_data, 
                       reference = spacexr_reference,
                       CELL_MIN_INSTANCE = 1, # minimum number of cells required per cell type (default = 25)
                       fc_cutoff = 0.25, # minimum log-fmale-change (across cell types) for genes to be included in the platform effect normalization step (default = 0.5)
                       UMI_min = 10) # minimum UMI per pixel included in the analysis (default = 100)

# run RCTD
rctd_out <- 
  spacexr::run.RCTD(RCTD = rctd_obj, 
                    doublet_mode = "full") # default is "doublet" which fits at most two cell types per pixel, full is any number

# qs_save(rctd_out, "~/workshop_data/rdata/rctd_out.qs2")

# rctd_out <- qs_read("~/workshop_data/rdata/rctd_out.qs2")

# normalize the cell type proportions to sum to 1
norm_weights <- 
  spacexr::normalize_weights(weights = rctd_out@results$weights)

head(norm_weights)

# add the deconvolution weights to the visium meta data 
spatial <- 
  AddMetaData(object = spatial, 
              metadata = norm_weights)

head(spatial@meta.data)

# qs_save(spatial, "~/workshop_data/rdata/merged_spatial_post_deconv.qs2")


## Read in here to save time
spatial <- qs_read("~/workshop_data/rdata/merged_spatial_post_deconv.qs2")
```

```{r}
rm(spacexr_reference, rctd_obj, rctd_out, norm_weights)
gc()
```

# Activity 2:

Use some of the spatial plotting functions to look at the deconvolution weight of some of the reference cell types in the kidney. Are any of them spatially-conserved?

Does the male mouse kidney have more of any of the cell types compared to the female kidney?

```{r deconvolution_visualization}

```

```{r cell_type_comparison}

```

--------------------------------------------------------------------------------

# Section 9: Visualize spot deconvolution with SPOTlight

```{r spotlight}
split <- 
  SplitObject(spatial, 
              split.by = "sex")

celltype_colors <-
  colorset[1:length(unique(ref$cell_type))]
names(celltype_colors) <- 
  levels(ref$cell_type)

scatterpies <- list()
for (sex in names(split)){
  spe <- split[[sex]]
  mat <- 
    spe@meta.data[,unique(ref$cell_type)]
  # reduce anything less than 10% to 0
  mat[mat < 0.1] <- 0
  Idents(spe) <- 
    rownames(spe@meta.data)
  DefaultAssay(spe) <- "Spatial"
  spe <- 
    GetTissueCoordinates(spe)[,c("x","y")]
  dim <-
    spe %>%
    as.data.frame()
  scatterpies[[sex]] <-
    plotSpatialScatterpie(
      x = spe,
      y = mat,
      cell_types = colnames(mat),
      img = FALSE,
      scatterpie_alpha = 0.8,
      pie_scale = 0.3) +
      scale_y_reverse() +
      scale_fill_manual(values = celltype_colors,
                        breaks = names(celltype_colors)) +
      theme(text = element_text(size = 14)) +
      ggtitle(sex) +
      theme(plot.title = element_text(hjust = 0.5)) +
      coord_fixed(ratio = 0.75)
}

wrap_plots(scatterpies)
```

```{r}
rm(split)
gc()
```

--------------------------------------------------------------------------------

# Section 10: Are specific clusters made up of specific cell types?

```{r cluster_celltypes}
deconv_mat <-
  spatial@meta.data %>%
  dplyr::select(seurat_clusters, unique(ref$cell_type)) %>%
  pivot_longer(cols = !c(seurat_clusters),
               values_to = "deconvolution_weight",
               names_to = "cell_type") %>%
  group_by(seurat_clusters, cell_type) %>%
  summarize(mean = mean(deconvolution_weight, na.rm = T)) %>%
  pivot_wider(names_from = cell_type,
              values_from = mean) %>%
  column_to_rownames(var = "seurat_clusters") %>%
  t()

pheatmap::pheatmap(mat = deconv_mat,
                   color = colorRampPalette(c("navy", "white","red"))(100),
                   main = "Deconvolution\nWeight",
                   scale = "row")
```

```{r}
rm(deconv_mat)
gc()
```

--------------------------------------------------------------------------------

# Section 11: Subset regions of interest

Let's look at the distribution of cells within the isocortex.

```{r subsetting_regions}
# which clusters roughly make up the isocortex?
SpatialDimPlot(object = spatial,
               group.by = "seurat_clusters",
               cols = cluster_colors,
               label = T,
               label.box = T,
               repel = T,
               image.alpha = 0,
               pt.size.factor = 3000)

# subset to those clusters
Idents(spatial) <- "seurat_clusters"
brush_cells <-
  subset(spatial, 
         seurat_clusters %in% c("4","5","6"))

# look at distribution of cell types within this region
brush_cells@meta.data %>%
  dplyr::select(sex, unique(ref$cell_type)) %>%
  pivot_longer(cols = !sex,
               values_to = "deconvolution_weight",
               names_to = "cell_type") %>% 
  ggplot(aes(x = deconvolution_weight,
             y = cell_type)) +
  geom_point(color = "grey",
             position = position_jitter(height = 0.2),
             alpha = 0.5) +
  geom_boxplot(alpha = 0) +
  scale_x_continuous(expand = c(0,NA)) +
  theme_bw()

# is it different between male and female mice?
brush_cells@meta.data %>%
  dplyr::select(sex, unique(ref$cell_type)) %>%
  pivot_longer(cols = !sex,
               values_to = "deconvolution_weight",
               names_to = "cell_type") %>% 
  ggplot(aes(y = deconvolution_weight,
             x = sex,
             color = sex)) +
  geom_point(alpha = 0.5) +
  geom_boxplot(alpha = 0) +
  facet_wrap(~cell_type) +
  scale_y_continuous(expand = c(0,NA)) +
  scale_color_manual(values = c("male" = "#4deeea",
                                "female" = "#fb2076")) +
  theme_bw()

# other ways to pick spots:

# interactive spatial plotting to pick coordinates
Idents(spatial) <- "x"
SpatialDimPlot(object = spatial,
               interactive = T,
               images = "W3_F_Kidney",
               image.alpha = 0,
               pt.size.factor = 3000)
SpatialDimPlot(object = spatial,
               interactive = T,
               images = "W3_M_Kidney",
               image.alpha = 0,
               pt.size.factor = 3000)
spatial <-
  AddMetaData(object = spatial,
              metadata = coords)
spots_to_keep <- 
  WhichCells(spatial, 
             expression = seurat_clusters == 5 & imagecol < 11000 & imagerow > 4000)
  

# loupe browser

```

```{r}
rm(spatial)
gc()
```

--------------------------------------------------------------------------------

# Section 12: Perform Differential Expression (DE) Analysis

Let's look at genes that are differentially expressed in the brush cells of the male and female mouse kidney 

```{r de}
brush_degs <-
  FindMarkers(object = brush_cells,
              ident.1 = "female",
              ident.2 = "male",
              group.by = "sex",
              assay = "Spatial") %>%
  rownames_to_column(var = "gene")

View(brush_degs)

SpatialFeaturePlot(object = brush_cells,
                   features = "Calr",
                   pt.size.factor = 3000,
                   image.alpha = 0)
```

--------------------------------------------------------------------------------

# Section 13: Perform Gene Set Enrichment Analysis (GSEA)

Let's look at pathways that are differentially regulated in the female vs. male isocortex 

```{r gsea}
# set up genesets
go_bp <- 
  msigdbr(species = "Mus musculus",
          category = "C5",
          subcategory = "GO:BP") %>% 
  split(x = .$gene_symbol,
        f = .$gs_name)
hallmark <- 
  msigdbr(species = "Mus musculus",
          category = "H") %>% 
  split(x = .$gene_symbol,
        f = .$gs_name)
genesets <- 
  c(go_bp, hallmark)

# add q value metric to do GSEA with
brush_degs <- 
  brush_degs %>%
  mutate(qvalue = (1-p_val_adj)*avg_log2FC)

# perform gsea
gsea_ranks <- 
  brush_degs %>%
  arrange(desc(qvalue)) %>%
  dplyr::select(gene, qvalue) %>%
  deframe()
gsea_res <- 
  fgseaMultilevel(genesets,
                  gsea_ranks,
                  minSize = 15,
                  maxSize = 500) %>%
      arrange(padj) # save with fwrite

# make GSEA dotplot
gsea_res_small <- 
  gsea_res %>%
  slice_min(order_by = padj,
            n = 20,
            with_ties = F) %>%
  mutate(type = ifelse(test = NES < 0,
                       yes = "downregulated",
                       no = "upregulated"),
         NES = abs(NES),
         pathway = gsub(pattern = "_",
                        replacement = " ",
                        x = pathway)) %>%
  filter(!is.na(pathway),
         !is.na(padj)) 

ggplot(gsea_res_small,
       aes(x = NES,
           y = fct_reorder(pathway, NES))) + 
  geom_point(aes(size = size,
                 color = padj)) +
  scale_color_gradient2(high = "#762A83",
                        mid = "#f7f7f7",
                        low = "#1B7837",
                        midpoint = 0.05) +
  ylab(NULL) +
  xlab("Absolute Value NES") +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 50)) +
  ggtitle("Brush Cell GSEA") +
  theme_bw(base_size = 11) +
  facet_grid(.~type) +
  theme(plot.title = element_text(hjust=0.5, 
                                  face="bmale"),
        strip.background = element_rect(fill = "white")) 
```

--------------------------------------------------------------------------------

# Section 14: Experimental Design Considerations (advanced_tuesday.pptx)

--------------------------------------------------------------------------------

# Section 15: Adapting to Other Technologies

There are new spatial technologies and analytical tools coming out all the time! We've successfully implemented workflows similar to that presented above for Bruker (Nanostring) GeoMx data, Takara (Curio) Trekker data, and 10X Visium HD data. Some modifications are necessary for GeoMx and HD data.

1. For GeoMx data, you can make use of GeomxTools and this vignette as a starting point: https://www.bioconductor.org/packages/release/bioc/vignettes/GeomxTools/inst/doc/GeomxSet_coercions.html
2. For Trekker data, the output of their nextflow pipeline is a Seurat object, so no coercion necessary.
3. For Visium HD data, you need to specify bin size and likely need to use Sketch-based approaches. This vignette is a good starting point: https://satijalab.org/seurat/articles/visiumhd_analysis_vignette

