---
title: "Prepping Workshop Data"
author: "Corinne Strawser"
date: "2026-05-28"
output: html_document
---

# For Monday:

1) "f1_sobj.qs" : Seurat object made from /workshop_data/raw/GSE232482/GSM7337030_mouse_scRNA-seq_P0_F1/filtered_feature_bc_matrix

```{r}
f1_raw <- 
  Read10X("workshop_data/raw/GSE232482/GSM7337030_mouse_scRNA-seq_P0_F1/filtered_feature_bc_matrix")
f1_sobj <-
    CreateSeuratObject(
        counts = f1_raw,
        min.cells = 3,
        min.features = 200,
    )
f1_sobj$orig.ident <- "GSM7337030_mouse_scRNA-seq_P0_F1"
qsave(f1_sobj, "workshop_data/rdata/f1_sobj.qs") #save to data/output/rdata directory
```

2) "m5_sobj.qs" : Seurat object made from /workshop_data/raw/GSE232482/GSM7337034_mouse_scRNA-seq_P0_M5/filtered_feature_bc_matrix

```{r}
m5_raw <- 
  Read10X("workshop_data/raw/GSE232482/GSM7337034_mouse_scRNA-seq_P0_M5/filtered_feature_bc_matrix")
m5_sobj <-
    CreateSeuratObject(
        counts = m5_raw,
        min.cells = 3,
        min.features = 200,
    )
m5_sobj$orig.ident <- "GSM7337034_mouse_scRNA-seq_P0_M5"
qsave(m5_sobj, "workshop_data/rdata/m5_sobj.qs") #save to data/output/rdata directory
```

3) "f1_sobj_processed_annotated.qs" : Seurat object processed and annotated

```{r}
sobj <-
  qs::qread("workshop_data/rdata/f1_sobj.qs")

sobj[["percent.mt"]] <- 
  PercentageFeatureSet(sobj, pattern = "^mt-")

sobj <- 
  subset(sobj,
         subset = nCount_RNA < 20000 &
           percent.mt < 20)

sobj <-
  NormalizeData(sobj,
                normalization.method = "LogNormalize",
                scale.factor = 10000) %>%
  FindVariableFeatures(selection.method = "vst",
                       nfeatures = 2000) %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:20) %>%
  FindClusters(resolution = 0.2) %>%
  RunUMAP(dims = 1:20,
          metric = "euclidean") %>%
  RunTSNE(dims = 1:20,
          metric = "euclidean")

qsave(sobj, "workshop_data/rdata/f1_sobj_processed.qs")

kidney_ref_data <- 
    readRDS("workshop_data/rdata/tabula_muris_senis_kidney.rds")

# singleR annotation
annotation <-
    SingleR(test = as.SingleCellExperiment(sobj),
            ref = GetAssayData(kidney_ref_data),
            labels = kidney_ref_data$cell_type,
            aggr.ref = TRUE)

# check the object
sobj$singleR_annotation <- annotation$labels

sobj$singleR_scores <- 
    apply(X = annotation$scores,
          MARGIN = 1,
          function(x) max(x, na.rm = TRUE))

DefaultAssay(kidney_ref_data) <- "RNA"

# Set the idents to cell_type so that we can examine which cell types are present in the dataset
Idents(kidney_ref_data) <- "cell_type"

# Find anchors between reference and query datasets
anchors <-
    FindTransferAnchors(reference = kidney_ref_data,
                      query = sobj,
                      dims = 1:30,
                      reference.reduction = "pca")

# Transfer anchors from reference dataset
predictions <-
    TransferData(anchorset = anchors,
               refdata = kidney_ref_data$cell_type,
               dims = 1:30)

# Add query to metadata of Seurat object
sobj <- AddMetaData(sobj,
                    metadata = predictions) 

# Rename "predicted.id" to "cell_types"
sobj$anchor_cell_types <- sobj$predicted.id

qsave(sobj, "workshop_data/rdata/f1_sobj_processed_annotated.qs") 
```

# For Tuesday:

4) "tabula_muris_senis_kidney.rds" : reference data

5) "f1_sobj_processed.qs" : processed from Monday ** generated above **

6) "W3_F_Kidney.qs" : Visium object

```{r}
W3_F_Kidney <-
  Load10X_Spatial(data.dir = paste0("workshop_data/raw/GSE252772/GSM8704839_20201201-NMK-F-Fc1U4Z1B1"),
                  slice = "W3_F_Kidney")
# add % of reads mapping to mitochondrial genes as a measure of cell health
W3_F_Kidney[["percent.mt"]] <-
  PercentageFeatureSet(object = W3_F_Kidney, 
                       pattern = "^Mt\\d")
head(W3_F_Kidney)

# add mouse age to meta data
W3_F_Kidney$age <- "w3"

# add sex to meta data
W3_F_Kidney$sex <- "female"

W3_F_Kidney <-
  W3_F_Kidney %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData(vars.to.regress = "percent.mt") 
W3_F_Kidney_clustered <-
  RunPCA(object = W3_F_Kidney,
         npcs = 30) %>%
  RunUMAP(dims = 1:30) %>%
  FindNeighbors(k.parm = 30) %>%
  FindClusters(resolution = 0.5)

qsave(W3_F_Kidney_clustered, "workshop_data/rdata/W3_F_Kidney.qs") 
```

7) "W3_M_Kidney.qs" : Visium object

```{r}
W3_M_Kidney <-
  Load10X_Spatial(data.dir = paste0("workshop_data/raw/GSE252772/GSM8704840_20201201-NMK-M-Fc1U1Z1B1/"),
                  slice = "W3_M_Kidney")
W3_M_Kidney$age <- "w3"
W3_M_Kidney$sex <- "male"
W3_M_Kidney[["percent.mt"]] <-
  PercentageFeatureSet(object = W3_M_Kidney, 
                       pattern = "^Mt\\d")

W3_M_Kidney <-
  W3_M_Kidney %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData(vars.to.regress = "percent.mt") 
W3_M_Kidney_clustered <-
  RunPCA(object = W3_M_Kidney,
         npcs = 30) %>%
  RunUMAP(dims = 1:30) %>%
  FindNeighbors(k.parm = 30) %>%
  FindClusters(resolution = 0.5)

qsave(W3_M_Kidney_clustered, "workshop_data/rdata/W3_M_Kidney.qs") 
```

8) "spatial_obj_list.qs" : Visium object list ** didn't make **

9) "spatial_merged" : W3_F_Kidney.qs and W3_M_Kidney.qs merged Visium object 

```{r}
# make a list of the data to integrate
obj_list <-
  list("W3_F_Kidney" = W3_F_Kidney,
       "W3_M_Kidney" = W3_M_Kidney)

# merge normalized, but not clustered objects
spatial <-
  merge(x = obj_list[[1]],
        y = obj_list[2:length(obj_list)],
        add.cell.ids = names(obj_list))

var_feats <-
  SelectIntegrationFeatures(object = obj_list,
                            nfeatures = 4000)
VariableFeatures(object = spatial) <- var_feats

# perform dimensionality reduction and clustering
DefaultAssay(object = spatial) <- "Spatial"
spatial <- 
  spatial %>%
  NormalizeData() %>%
  ScaleData() %>%
  RunPCA(npcs = 30) %>%
  RunUMAP(dims = 1:30) %>%
  FindNeighbors(k.parm = 30) %>%
  FindClusters(res = 0.6)

# join Spatial data layers
spatial <- 
  JoinLayers(object = spatial,
             assay = "Spatial")

# load in reference data
ref <-
  readRDS(file = paste0("workshop_data/rdata/tabula_muris_senis_kidney.rds"))

# spacexr doesn't like labels with "/" in it, so let's change that
ref$cell_type <-
  gsub(pattern = "/", 
       replacement = "-", 
       x = ref$cell_type)

# create the spacexr reference object
spacexr_reference <-
    spacexr::Reference(counts = GetAssayData(object = ref, 
                                             layer = "counts"),
                       cell_types = as.factor(ref$cell_type))

# get x and y coordinates for our visium data
coords <- 
  rbind(GetTissueCoordinates(object = spatial, 
                             image = "W3_F_Kidney"),
        GetTissueCoordinates(object = spatial, 
                             image = "W3_M_Kidney")) %>%
  dplyr::select(!cell)

# create the spacexr data object
my_data <- 
  spacexr::SpatialRNA(coords = coords,
                      counts = GetAssayData(object = spatial, 
                                            layer = "counts",
                                            assay = "Spatial"))

# create the rctd object 
rctd_obj <- 
  spacexr::create.RCTD(spatialRNA = my_data, 
                       reference = spacexr_reference,
                       CELL_MIN_INSTANCE = 1, # minimum number of cells required per cell type (default = 25)
                       fc_cutoff = 0.25, # minimum log-fold-change (across cell types) for genes to be included in the platform effect normalization step (default = 0.5)
                       UMI_min = 10) # minimum UMI per pixel included in the analysis (default = 100)

# run RCTD
rctd_out <- 
  spacexr::run.RCTD(RCTD = rctd_obj, 
                    doublet_mode = "full") # default is "doublet" which fits at most two cell types per pixel, full is any number

# normalize the cell type proportions to sum to 1
norm_weights <- 
  spacexr::normalize_weights(weights = rctd_out@results$weights)

head(norm_weights)

# add the deconvolution weights to the visium meta data 
spatial <- 
  AddMetaData(object = spatial, 
              metadata = norm_weights)

qsave(spatial, "workshop_data/rdata/spatial_merged.qs") 
```



# For Wednesday: 

10) "GSE232482_object_list_processed_annotated.qs" : processed data from GSE232482 with 1 female and 2 male

```{r}
samples <- 
  c("GSM7337031_mouse_scRNA-seq_P0_F2",
    "GSM7337034_mouse_scRNA-seq_P0_M5",
    "GSM7337035_mouse_scRNA-seq_P0_M6")

object_list <- list()
for (smp in samples) {
  raw <- 
    Read10X(paste0("workshop_data/raw/GSE232482/", smp, "/filtered_feature_bc_matrix"))
  obj <-
    CreateSeuratObject(counts = raw,
                       min.cells = 3,
                       min.features = 200)
  obj$orig.ident <- smp
  obj[["percent.mt"]] <- 
    PercentageFeatureSet(obj, pattern = "^mt-")
  obj <- 
    subset(obj,
         subset = nCount_RNA < 20000 &
           percent.mt < 20)
  obj <-
    NormalizeData(obj,
                  normalization.method = "LogNormalize",
                  scale.factor = 10000) %>%
    FindVariableFeatures(selection.method = "vst",
                         nfeatures = 2000) %>%
    ScaleData() %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:20) %>%
    FindClusters(resolution = 0.2) %>%
    RunUMAP(dims = 1:20,
            metric = "euclidean") %>%
    RunTSNE(dims = 1:20,
            metric = "euclidean")
  Idents(ref) <- "cell_type"
  DefaultAssay(ref) <- "RNA"
  anchors <-
      FindTransferAnchors(reference = ref,
                        query = obj,
                        dims = 1:30,
                        reference.reduction = "pca")
  predictions <-
      TransferData(anchorset = anchors,
                 refdata = ref$cell_type,
                 dims = 1:30)
  obj <- AddMetaData(obj,
                      metadata = predictions) 
  
  obj$anchor_cell_types <- obj$predicted.id
  object_list[[smp]] <- obj
}

qsave(object_list, "workshop_data/rdata/GSE232482_object_list_processed_annotated.qs")
```


11) "merged_GSE232482_sobj.qs" : merged and processed GSE232482 data 

```{r}
merged <-
  merge(x = object_list[[1]],
        y = object_list[2:length(object_list)],
        add.cell.ids = names(object_list))

qsave(merged, "workshop_data/rdata/merged_GSE232482_sobj.qs")
```

12) "GSE252772_object_list_processed_annotated_small.qs" : merged and processed snRNA-seq data from GSE252772 with 1 p0 females, 1 p0 males, 1 w3 female, and 1 w3 male

```{r}
samples <- 
  c("GSE252772_20210419-NMK3F",
    "GSE252772_20210419-NMK3M",
    "GSE252772_20220129-NMK0F",
    "GSE252772_20220129-NMK0M")

object_list <- list()
for (smp in samples) {
  raw <- 
    Read10X(paste0("/igm/home/cnh008/ref/GSE252772/snrna_inputs/", smp, "/filtered_feature_bc_matrix"))
  obj <-
    CreateSeuratObject(counts = raw$`Gene Expression`,
                       min.cells = 3,
                       min.features = 200)
  obj$orig.ident <- smp
  obj[["percent.mt"]] <- 
    PercentageFeatureSet(obj, pattern = "^mt-")
  obj <- 
    subset(obj,
         subset = nCount_RNA < 20000 &
           percent.mt < 20)
  obj <-
    subset(obj,
           downsample = 2000)
  obj <-
    NormalizeData(obj,
                  normalization.method = "LogNormalize",
                  scale.factor = 10000) %>%
    FindVariableFeatures(selection.method = "vst",
                         nfeatures = 2000) %>%
    ScaleData() %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:20) %>%
    FindClusters(resolution = 0.2) %>%
    RunUMAP(dims = 1:20,
            metric = "euclidean") %>%
    RunTSNE(dims = 1:20,
            metric = "euclidean")
  Idents(ref) <- "cell_type"
  DefaultAssay(ref) <- "RNA"
  anchors <-
      FindTransferAnchors(reference = ref,
                        query = obj,
                        dims = 1:20,
                        reference.reduction = "pca")
  predictions <-
      TransferData(anchorset = anchors,
                 refdata = ref$cell_type,
                 dims = 1:20)
  obj <- AddMetaData(obj,
                      metadata = predictions) 
  
  obj$anchor_cell_types <- obj$predicted.id
  object_list[[smp]] <- obj
}

qsave(object_list, "workshop_data/rdata/GSE252772_object_list_processed_annotated_small.qs")
```

13) "merged_sobj_before_harmony.qs" : merged, processed, annotated, GSE232482 and GSE252772 object before harmony integration

```{r}
# read in the second object list
GSE252772_sobj_list <-
    qs::qread("workshop_data/rdata/GSE252772_object_list_processed_annotated_small.qs")

# merge the second object list
merged_GSE252772_sobj <-
  merge(x = GSE252772_sobj_list[[1]],
        y = GSE252772_sobj_list[2:length(GSE252772_sobj_list)],
        add.cell.ids = names(GSE252772_sobj_list))

#save the batch information
merged_GSE252772_sobj$batch <- "GSE252772"

# save the batch information in the merged_sobj as well
merged_GSE232482_sobj <-
    qs::qread("workshop_data/rdata/merged_GSE232482_sobj.qs")
merged_GSE232482_sobj$batch <- "GSE232482"

# merge the two objects and process it together unto plotting dimplot
merged_sobj <-
    merge(x= merged_GSE232482_sobj,
          y = merged_GSE252772_sobj)

merged_sobj <-
    merged_sobj %>%
    JoinLayers()

merged_sobj <-
    merged_sobj %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:30) %>%  
    FindNeighbors(dims = 1:30) %>%
    FindClusters(resolution = 0.5)

qs::qsave(merged_sobj,
          "workshop_data/rdata/merged_sobj_before_harmony.qs")
```

14) "merged_sobj_harmony_batch.qs"
```{r}
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = "batch") %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters(resolution = 0.5) 

qs::qsave(merged_sobj,
          "workshop_data/rdata/merged_sobj_harmony_batch.qs")
```

15) "merged_sobj_harmony_theta.qs" : harmony integration with theta = 5

```{r}
# run Harmony integration with different parameters
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = "batch",
               theta = 5) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()
qs::qsave(merged_sobj_harmony,
    "workshop_data/rdata/merged_sobj_harmony_theta.qs")
```

16) "merged_sobj_harmony_sample.qs" : sample as batch variable in harmony integration

```{r}
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = "orig.ident") %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters(resolution = 0.5) 

qs::qsave(merged_sobj,
          "workshop_data/rdata/merged_sobj_harmony_sample.qs")
```

17) "merged_sobj_harmony_sample_theta.qs" : harmony integration with theta = 5 and sample as batch variable

```{r}
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = "orig.ident",
               theta = 5) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters(resolution = 0.5) 

qs::qsave(merged_sobj,
          "workshop_data/rdata/merged_sobj_harmony_sample_theta.qs")
```

18) "merged_sobj_harmony_sample_batch.qs" : harmony integration with sample AND seq batch as batch variables

```{r}
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = c("batch", "orig.ident")) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()
qs::qsave(merged_sobj_harmony,
    "workshop_data/rdata/merged_sobj_harmony_sample_batch.qs")
```

19) "merged_sobj_harmony_sample_batch_theta.qs" : harmony integration with sample AND seq batch as batch variables and theta = 10

```{r}
merged_sobj_harmony <-
    merged_sobj %>%
    RunHarmony(group.by.vars = c("batch", "orig.ident"),
               theta = c(10, 10)) %>%
    RunUMAP(reduction = "harmony", dims = 1:30) %>%
    FindNeighbors(reduction = "harmony", dims = 1:30) %>%
    FindClusters()
qs::qsave(merged_sobj_harmony,
    "workshop_data/rdata/merged_sobj_harmony_sample_batch_theta.qs")
```

20) "merged_sobj_rpca.qs" : RPCA integration

```{r}
# Lets downsample a bit so it runs quicker
Idents(merged_sobj) <- merged_sobj$orig.ident
merged_sobj <- subset(merged_sobj, downsample = 500)

# First, we need to split the layers of the merged object to allow for integration to happen. 

merged_sobj[["RNA"]] <- split(merged_sobj[["RNA"]], f = merged_sobj$orig.ident)

# Then, we can perform the actual integration. 

rpca <- IntegrateLayers(
  object = merged_sobj, 
  method = RPCAIntegration,
  orig.reduction = "pca", 
  new.reduction = "integrated.rpca",
  verbose = FALSE)

rpca <- rpca%>%
  RunPCA(verbose = FALSE)%>%
  RunUMAP(reduction = "integrated.rpca", dims = 1:30)%>%
  FindNeighbors(dims = 1:30, reduction = "integrated.rpca")%>%
  FindClusters(resolution = 0.1)

qs::qsave(rpca,
    "workshop_data/rdata/merged_sobj_rpca.qs")
```

20) "merged_sobj_cca.qs" : CCA integration
```{r}
cca <- IntegrateLayers(
  object = merged_sobj, 
  method =CCAIntegration,
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  verbose = FALSE)

cca <- cca%>%
  RunPCA(verbose = FALSE)%>%
  RunUMAP(reduction = "integrated.cca", dims = 1:30)%>%
  FindNeighbors(dims = 1:30, reduction = "integrated.cca")%>%
  FindClusters(resolution = 0.1)

qs::qsave(cca,
    "workshop_data/rdata/merged_sobj_cca.qs")
```

# For Thursday:

21) "annotations.qs" : annotated atac data

```{r}

```


22) "mouse_kidney_processed.qs" : fully processed atac data