## Activity 1

We have another dataset for mouse kidney nuclei. They were aligned to the same genome as the lesson dataset.

### 1.1

Read in the gene expression and chromatin accessibility matrices using Read10X_h5()*.

```{r activity-1-1}
data_path <- "~/workshop_data/data/raw/"
session_path <- "Sessions/adv_thursday/"

my_h5_data <-
    Read10X_h5(paste0(
        data_path,
        "mouse_kidney_nuc_multi/M_Kidney_Chromium_Nuc_Isolation_vs_SaltyEZ_vs_ComplexTissueDP_filtered_feature_bc_matrix.h5"
    ))

# downsample h5 data
set.seed(17)
keep <- sample(colnames(my_h5_data[[1]]), 5000)
my_h5_data[[1]] <- my_h5_data[[1]][ , keep]
my_h5_data[[2]] <- my_h5_data[[2]][ , keep]
```

### 1.2

Create a Seurat object with an RNA and ATAC assay

```{r activity-1-2}
my_seurat_obj <-
    CreateSeuratObject(
        counts = my_h5_data$`Gene Expression`,
        assay = "RNA",
        project = "10x_multiomics"
    ) %>%
    PercentageFeatureSet(
        pattern = "^mt-",
        col.name = "percent_mt_rna",
        assay = "RNA"
    )

my_frag_file <-
    paste0(
        data_path,
        "mouse_kidney_nuc_multi/M_Kidney_Chromium_Nuc_Isolation_vs_SaltyEZ_vs_ComplexTissueDP_atac_fragments.tsv.gz"
    )

my_seurat_obj[["ATAC"]] <-
    CreateChromatinAssay(
        counts = my_h5_data$Peaks,
        sep = c(":", "-"),
        fragments = my_frag_file,
        min.cells = 5,
        annotation = annotations
    )
DefaultAssay(my_seurat_obj) <- "ATAC"
rm(my_h5_data)
gc()
```
# Activity 2

## 2.1

Compute clusters for the ATAC data. Use `resolution = 0.1` in find clusters.

```{r activity-2}
DefaultAssay(my_seurat_obj) <- "ATAC"
my_seurat_obj <- my_seurat_obj %>% 
    RunTFIDF() %>%
    FindTopFeatures(min.cutoff = "q0") %>%
    RunSVD() %>%
    FindNeighbors(reduction = "lsi") %>%
    FindClusters(algorithm = 3, resolution = 0.1) %>%
    RunUMAP(
        reduction = "lsi",
        dims = 2:30,
        reduction.name = "umap_atac"
    )
my_seurat_obj$atac_clusters <- my_seurat_obj$seurat_clusters
```

## 2.2

Explore the different clusters based on the ATAC data and the RNA seq data. Does anything stick out to you? Try plotting rna clusters with the atac umap embeddings and vice versa.

```{r activity-2-2}
DimPlot(my_seurat_obj, group.by = c("rna_clusters", "atac_clusters"), reduction = "umap_rna")
DimPlot(my_seurat_obj, group.by = c("rna_clusters", "atac_clusters"), reduction = "umap_atac")
```


# Activity 3

# Activity 3

I'm thinking for this we'll do...
    - differential accessibility between two groups in our dataset
    - get top 5 upregulated regions for each group
    - create ATAC feature plot of these regions
    - create joint umap
    - link peaks to genes
    - plot expression of peak-linked genes

## 3.1

In our `new_obj` object from activity 2, find the top 5 differentially accessibile peaks between the two groups as determined by log-fold change, as well as the top differentially expressed genes.

```{r activity-3-1}
DefaultAssay(new_obj) <- "ATAC"
Idents(new_obj) <- new_obj$atac_clusters
diff_peaks <- FindAllMarkers(new_obj,
    assay = "ATAC",
    min.pct = .1
) %>%
    subset(p_val_adj < 0.05)

top_5_each <- group_by(diff_peaks, cluster) %>%
    arrange(desc(avg_log2FC)) %>%
    slice_head(n = 5)
```

## 3.2

Create a coverage plot of these regions. Note you'll have to go through the steps of linking peaks to genes, adding motif information to your object, etc. Refer to the code from before to do this.

```{r activity-3-2}

# link peaks to genes first
new_obj <- 
  AddMotifs(
        object = new_obj,
        genome = BSgenome.Mmusculus.UCSC.mm39::BSgenome.Mmusculus.UCSC.mm39,
        pfm = freq_matrix
    )

FindMotifs(new_obj,)


links <-
    Links(new_obj) %>%
    as.data.frame() %>%
    as_tibble() %>%
    arrange(pvalue)

CoveragePlot(
    object = new_obj,
    region = top_5_each$gene[1:4],
    idents = c("1", "2"),
    peaks.group.by = "sig_peaks",
    expression.assay = "RNA",
    extend.upstream = 100000,
    extend.downstream = 100000
)
```

## 3.3

Identify what genes are associated with these peaks. Plot a few of their expressions and the associated peaks' accessibility.

```{r activity-3-3}
degs <-
    FindMarkers(
        new_obj,
        assay = "RNA",
        ident.1 = "1",
        ident.2 = "2",
        logfc.threshold = 2
    ) %>%
    dplyr::filter(p_val_adj < 0.000001 & # Using a really low number since the p-values are so skewed
        abs(pct.1 - pct.2) > 0.6) %>%
    arrange(desc(abs(pct.1 - pct.2)))

# Getting this down to a smallish number so it doesn't take forever to run
nrow(degs)

# link peaks to top 10 genes as determined by the difference in pct expression
new_obj <-
    LinkPeaks(
        object = new_obj,
        peak.assay = "ATAC",
        expression.assay = "RNA",
        min.distance = 50000,
        genes.use = rownames(degs)
    )


links <-
    Links(new_obj) %>%
    as.data.frame() %>%
    as_tibble() %>%
    arrange(pvalue)
```

## 3.4

Find what motifs are enriched in our differentially accessible peaks.

```{r activity-3-4}
new_obj <-
    AddMotifs(
        object = new_obj,
        genome = BSgenome.Mmusculus.UCSC.mm39::BSgenome.Mmusculus.UCSC.mm39,
        pfm = freq_matrix
    )


matching_motifs <-
    FindMotifs(
        object = new_obj,
        features = links$peak %>% unique()
    )

dplyr::select(
    matching_motifs,
    percent.observed,
    percent.background,
    motif.name,
    p.adjust
) %>%
    arrange(p.adjust) %>%
    head(n = 15)
```
