---
title: "Coder Upgrade 2023: Cell Cell interaction analysis"
author: "Matt Cannon"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 5
        number_sections: false
        code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      cache.lazy = FALSE)
```

Load libraries
```{r libraries, cache=FALSE, warning=FALSE, error=FALSE, message=FALSE,}
library(rrrSingleCellUtils)
library(tidyverse)
library(Seurat)
library(nichenetr)

.libPaths("/gpfs0/export/apps/opt/scrgot/23/lib/R")

```

--------------------------------------------------------------------------------

# Section 1: Overview


## What is the goal of cell-cell interaction analysis?

## What are common tools used to do this?

CellphoneDB
Cellchat


## Nichenetr

Nichenetr relies on a pre-defined model of ligand - target interactions
It applies this model to your data to identify possible interactions


## Original method

## Get their pre-defined model data
From:
"https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"
"https://zenodo.org/record/3260758/files/lr_network.rds"
"https://zenodo.org/record/3260758/files/weighted_networks.rds"
```{r}
ligand_target_matrix <-
    readRDS("/gpfs0/home1/gdworkshop/lab/session_data/ligand_target_matrix.rds")

lr_network <-
    readRDS("/gpfs0/home1/gdworkshop/lab/session_data/lr_network.rds")

weighted_networks <-
    readRDS("/gpfs0/home1/gdworkshop/lab/session_data/weighted_networks.rds")

weighted_networks_lr <-
    weighted_networks$lr_sig %>%
    inner_join(lr_network %>% distinct(from, to),
               by = c("from", "to"))
```

## Get data to work with
From "https://zenodo.org/record/5840787/files/seurat_obj_subset_integrated_zonation.rds"
```{r}
seurat_obj <-
    readRDS("/gpfs0/home/gdworkshop/lab/session_data/seuratObj.rds")

DimPlot(seurat_obj, group.by = "celltype", label = TRUE)
```

## Convert gene names in model to mouse
```{r}
lr_network <-
    lr_network %>%
    mutate(from = convert_human_to_mouse_symbols(from),
           to = convert_human_to_mouse_symbols(to)) %>%
    drop_na()

colnames(ligand_target_matrix) <-
    ligand_target_matrix %>%
    colnames() %>%
    convert_human_to_mouse_symbols()

rownames(ligand_target_matrix) <-
    ligand_target_matrix %>%
    rownames() %>%
    convert_human_to_mouse_symbols()

ligand_target_matrix <-
    ligand_target_matrix %>%
    .[!is.na(rownames(ligand_target_matrix)),
        !is.na(colnames(ligand_target_matrix))]

weighted_networks_lr <-
    weighted_networks_lr %>%
    mutate(from = convert_human_to_mouse_symbols(from),
           to = convert_human_to_mouse_symbols(to)) %>%
    drop_na()
```

## Define sender and reciever cells
Also make sure they're in our model
```{r}
## receiver
receiver <- "CD8 T"
expressed_genes_receiver <-
    get_expressed_genes(receiver,
                        seuratObj,
                        pct = 0.10)

background_expressed_genes <-
    expressed_genes_receiver %>%
    .[. %in% rownames(ligand_target_matrix)]

## sender
sender_celltypes <- c("CD4 T",
                      "Treg",
                      "Mono",
                      "NK",
                      "B",
                      "DC")

list_expressed_genes_sender <-
    sender_celltypes %>%
    # lapply to get the expressed genes of every sender cell type separately here
    lapply(get_expressed_genes, seuratObj, 0.10)

expressed_genes_sender <-
    list_expressed_genes_sender %>%
    unlist() %>%
    unique()
```

## Use FindMarkers to define out genes of interest
Can use other methods to define the group of genes you want to look at
Get just the reciever cells and do DE based on conditions ("aggregate" in this case)
Then make sure these genes are in our model
```{r}
seurat_obj_receiver <- subset(seuratObj, idents = receiver)
seurat_obj_receiver <-
    SetIdent(seurat_obj_receiver,
             value = seurat_obj_receiver[["aggregate"]])

DE_table_receiver <-
    FindMarkers(object = seurat_obj_receiver,
                ident.1 = "LCMV",
                ident.2 = "SS",
                min.pct = 0.10) %>%
    rownames_to_column("gene")

geneset_oi <-
    DE_table_receiver %>%
    filter(p_val_adj <= 0.05 &
            abs(avg_log2FC) >= 0.25) %>%
    pull(gene)

geneset_oi <-
    geneset_oi %>%
    .[. %in% rownames(ligand_target_matrix)]
```

## Define potential ligands

```{r}
ligands <-
    lr_network %>%
    pull(from) %>%
    unique()

receptors <-
    lr_network %>%
    pull(to) %>%
    unique()

expressed_ligands <- intersect(ligands, expressed_genes_sender)
expressed_receptors <- intersect(receptors, expressed_genes_receiver)

potential_ligands <-
    lr_network %>%
    filter(from %in% expressed_ligands &
            to %in% expressed_receptors) %>%
    pull(from) %>%
    unique()
```

## 
```{r}
ligand_activities <-
    predict_ligand_activities(geneset = geneset_oi,
                              background_expressed_genes = background_expressed_genes,
                              ligand_target_matrix = ligand_target_matrix,
                              potential_ligands = potential_ligands)

ligand_activities <-
    ligand_activities %>%
    arrange(-pearson) %>%
    mutate(rank = rank(desc(pearson)))

```



## Easy mode

This will work if you are comparing between two conditions such as treated/untreated
List of genes to 
```{r}
nichenet_output <-
    nichenet_seuratobj_aggregate(seurat_obj = seurat_obj,
                                 receiver = "CD8 T",
                                 condition_colname = "aggregate",
                                 condition_oi = "LCMV",
                                 condition_reference = "SS",
                                 sender = c("CD4 T",
                                            "Treg",
                                            "Mono",
                                            "NK",
                                            "B",
                                            "DC"),
                                 ligand_target_matrix = ligand_target_matrix,
                                 lr_network = lr_network,
                                 weighted_networks = weighted_networks,
                                 organism = "mouse")
```



# Activity (1:30 - 2ish)
Run nichenetr on os-shark?
# Section 2: Visualizing nichenetr results


# Activity (2:30 - 3ish)
Make plots from nichenetr outputs? 

# Section 3: Interpreting nichenetr results
Activity (3:30 - 4ish)
 
# Session challenge
Run nichenetr on your own data or data from some random paper
Run cell-cell interaction analysis with another package
Compare the two


# Resources

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9184522/
https://www.nature.com/articles/s41592-019-0667-5


















# Extra stuff to run Differential NicheNetr
```{r}
lr_network <-
    readRDS("/gpfs0/home1/gdworkshop/lab/session_data/lr_network.rds") %>%
    mutate(bonafide = !database %in% c("ppi_prediction",
                                        "ppi_prediction_go")) %>%
    dplyr::rename(ligand = from, receptor = to) %>%
    distinct(ligand, receptor, bonafide)

```

## Get data to work with
From "https://zenodo.org/record/5840787/files/seurat_obj_subset_integrated_zonation.rds"
```{r}
seurat_obj <-
    readRDS("/gpfs0/home/gdworkshop/lab/session_data/seurat_obj_subset_integrated_zonation.rds")
```

## Convert the species from human to mouse
```{r}
head(lr_network)
lr_network <-
    lr_network %>%
    mutate(ligand = convert_human_to_mouse_symbols(ligand),
           receptor = convert_human_to_mouse_symbols(receptor)) %>%
    drop_na()
head(lr_network)

colnames(ligand_target_matrix) <-
    ligand_target_matrix %>%
    colnames() %>%
    convert_human_to_mouse_symbols()

rownames(ligand_target_matrix) <-
    ligand_target_matrix %>%
    rownames() %>%
    convert_human_to_mouse_symbols()

ligand_target_matrix <-
    ligand_target_matrix %>%
    .[!is.na(rownames(ligand_target_matrix)),
        !is.na(colnames(ligand_target_matrix))]
```

## Define which cells we're comparing
```{r}
niche_list <- list(
    "KC_niche" = list("sender" = c("LSECs_portal",
                                   "Hepatocytes_portal",
                                   "Stellate cells_portal"),
                      "receiver" = c("KCs")),
    "MoMac2_niche" = list("sender" = c("Cholangiocytes",
                                       "Fibroblast 2"),
                          "receiver" = c("MoMac2")),
    "MoMac1_niche" = list("sender" = c("Capsule fibroblasts",
                                       "Mesothelial cells"),
                          "receiver" = c("MoMac1"))
  )
```

## Do DE to define genes of interest

### There were originally multiple SCT-transformed assays, so we need to integrate them
```{r}
assay_name <- "SCT"

seurat_obj <-
    PrepSCTFindMarkers(seurat_obj,
                       assay = assay_name,
                       verbose = FALSE)
```

### Do DE between niches, then process the results
Run DE analysis, then get rid of infinite values and calculate further statistics for DE genes

```{r}
expression_pct <- 0.10
de_sender_processed <-
    calculate_niche_de(seurat_obj = seurat_obj %>%
                            subset(features = lr_network$ligand %>%
                                intersect(rownames(seurat_obj))),
                       niches = niche_list,
                       type = "sender",
                       assay_oi = assay_name) %>%
    # Fix any infinite values by making them the max/min finite value
    mutate(avg_log2FC = ifelse(avg_log2FC == Inf,
                               max(avg_log2FC[is.finite(avg_log2FC)]),
                               ifelse(avg_log2FC == -Inf,
                                      min(avg_log2FC[is.finite(avg_log2FC)]),
                                      avg_log2FC))) %>%
    process_niche_de(niches = niche_list,
                     expression_pct = expression_pct,
                     type = "sender")

de_receiver_processed <-
    calculate_niche_de(seurat_obj = seurat_obj %>%
                            subset(features = lr_network$receptor %>%
                                   unique()),
                       niches = niche_list,
                       type = "receiver",
                       assay_oi = assay_name) %>%
    # Fix any infinite values by making them the max/min finite value
    mutate(avg_log2FC = ifelse(avg_log2FC == Inf,
                               max(avg_log2FC[is.finite(avg_log2FC)]),
                               ifelse(avg_log2FC == -Inf,
                                      min(avg_log2FC[is.finite(avg_log2FC)]),
                                      avg_log2FC))) %>%
    process_niche_de(niches = niche_list,
                     expression_pct = expression_pct,
                     type = "receiver")
```

```{r}
de_sender_receiver <-
    combine_sender_receiver_de(de_sender_processed,
                               de_receiver_processed,
                               lr_network,
                               specificity_score = "min_lfc")
```

```{r}
lfc_cutoff <- 0.15

de_receiver_processed_targets <-
    calculate_niche_de_targets(seurat_obj = seurat_obj,
                               niches = niche_list,
                               lfc_cutoff = lfc_cutoff,
                               expression_pct = expression_pct,
                               assay_oi = assay_name) %>%
    process_receiver_target_de(niches = niche_list,
                               expression_pct = expression_pct,
                               specificity_score = "min_lfc")

background <-
    de_receiver_processed_targets  %>%
    pull(target) %>%
    unique()

geneset_KC <-
    de_receiver_processed_targets %>%
    filter(receiver == niche_list$KC_niche$receiver &
            target_score >= lfc_cutoff &
            target_significant == 1 &
            target_present == 1) %>%
    pull(target) %>%
    unique()

geneset_MoMac2 <-
    de_receiver_processed_targets %>%
    filter(receiver == niche_list$MoMac2_niche$receiver &
            target_score >= lfc_cutoff &
            target_significant == 1 &
            target_present == 1) %>%
    pull(target) %>%
    unique()

geneset_MoMac1 <-
    de_receiver_processed_targets %>%
    filter(receiver == niche_list$MoMac1_niche$receiver &
            target_score >= lfc_cutoff &
            target_significant == 1 &
            target_present == 1) %>%
    pull(target) %>%
    unique()
```

```{r}
top_n_target <- 250

niche_geneset_list <- list(
    "KC_niche" = list("receiver" = "KCs",
                      "geneset" = geneset_KC,
                      "background" = background),
    "MoMac1_niche" = list("receiver" = "MoMac1",
                          "geneset" = geneset_MoMac1,
                          "background" = background),
    "MoMac2_niche" = list("receiver" = "MoMac2",
                          "geneset" = geneset_MoMac2,
                          "background" = background))

ligand_activities_targets <-
    get_ligand_activities_targets(niche_geneset_list = niche_geneset_list,
                                  ligand_target_matrix = ligand_target_matrix,
                                  top_n_target = top_n_target)

```

## Get table of normalized expression values for genes of interest

```{r}
features_oi <-
    union(lr_network$ligand,
          lr_network$receptor) %>%
    union(ligand_activities_targets$target) %>%
    setdiff(NA)

exprs_tbl <-
    suppressWarnings(Seurat::DotPlot(seurat_obj %>%
                                     subset(idents = niche_list %>%
                                            unlist() %>%
                                            unique()),
                                     features = features_oi,
                                     assay = assay_name))$data %>%
    as_tibble() %>%
    rename(celltype = id,
           gene = features.plot,
           expression = avg.exp,
           expression_scaled = avg.exp.scaled,
           fraction = pct.exp) %>%
    mutate(fraction = fraction / 100) %>%
    as_tibble() %>%
    select(celltype,
           gene,
           expression,
           expression_scaled,
           fraction) %>%
    distinct() %>%
    arrange(gene) %>%
    mutate(gene = as.character(gene))
```

```{r}
exprs_tbl_ligand <-
    exprs_tbl %>%
    filter(gene %in% lr_network$ligand) %>%
    rename(sender = celltype,
           ligand = gene,
           ligand_expression = expression,
           ligand_expression_scaled = expression_scaled,
           ligand_fraction = fraction) %>%
    mutate(scaled_ligand_expression_scaled =
                scale_quantile_adapted(ligand_expression_scaled),
           ligand_fraction_adapted = ligand_fraction) %>%
    mutate_cond(ligand_fraction >= expression_pct,
                ligand_fraction_adapted = expression_pct)  %>%
    mutate(scaled_ligand_fraction_adapted =
                scale_quantile_adapted(ligand_fraction_adapted))

exprs_tbl_receptor <-
    exprs_tbl %>%
    filter(gene %in% lr_network$receptor) %>%
    rename(receiver = celltype,
           receptor = gene,
           receptor_expression = expression,
           receptor_expression_scaled = expression_scaled,
           receptor_fraction = fraction) %>%
    mutate(scaled_receptor_expression_scaled =
                scale_quantile_adapted(receptor_expression_scaled),
           receptor_fraction_adapted = receptor_fraction) %>%
    mutate_cond(receptor_fraction >= expression_pct,
                receptor_fraction_adapted = expression_pct) %>%
    mutate(scaled_receptor_fraction_adapted =
                scale_quantile_adapted(receptor_fraction_adapted))

exprs_tbl_target <-
    exprs_tbl %>%
    filter(gene %in% ligand_activities_targets$target) %>%
    rename(receiver = celltype,
           target = gene,
           target_expression = expression,
           target_expression_scaled = expression_scaled,
           target_fraction = fraction)
```

```{r}
exprs_sender_receiver <-
    lr_network %>%
    inner_join(exprs_tbl_ligand, by = c("ligand"), multiple = "all") %>%
    inner_join(exprs_tbl_receptor, by = c("receptor"), multiple = "all") %>%
    inner_join(de_sender_receiver %>%
                    distinct(niche, sender, receiver), multiple = "all")

```


```{r}
ligand_scaled_receptor_expression_fraction_df <-
    exprs_sender_receiver %>%
    group_by(ligand, receiver) %>%
    mutate(rank_receptor_expression = dense_rank(receptor_expression),
           rank_receptor_fraction  = dense_rank(receptor_fraction)) %>%
    mutate(ligand_scaled_receptor_expression_fraction =
                0.5 * ((rank_receptor_fraction / max(rank_receptor_fraction)) +
                     ((rank_receptor_expression / max(rank_receptor_expression)))))  %>%
    distinct(ligand,
             receptor,
             receiver,
             ligand_scaled_receptor_expression_fraction,
             bonafide) %>%
    distinct() %>%
    ungroup()
```

```{r}
prioritizing_weights <-
    c("scaled_ligand_score" = 5,
      "scaled_ligand_expression_scaled" = 1,
      "ligand_fraction" = 1,
      "scaled_ligand_score_spatial" = 2,
      "scaled_receptor_score" = 0.5,
      "scaled_receptor_expression_scaled" = 0.5,
      "receptor_fraction" = 1,
      "ligand_scaled_receptor_expression_fraction" = 1,
      "scaled_receptor_score_spatial" = 0,
      "scaled_activity" = 0,
      "scaled_activity_normalized" = 1,
      "bona_fide" = 1)
```

### Make fake spatial data since we don't have it
```{r}
spatial_info <-
    tibble(celltype_region_oi = c("LSECs_portal",
                                  "Hepatocytes_portal",
                                  "Stellate cells_portal"),
           celltype_other_region = c("LSECs_central",
                                     "Hepatocytes_central",
                                     "Stellate cells_central")) %>%
  mutate(niche =  "KC_niche",
         celltype_type = "sender")

sender_spatial_de_processed <-
    get_non_spatial_de(niches = niche_list,
                       spatial_info = spatial_info,
                       type = "sender",
                       lr_network = lr_network) %>%
    mutate(scaled_ligand_score_spatial =
            scale_quantile_adapted(ligand_score_spatial))

receiver_spatial_de_processed <-
    get_non_spatial_de(niches = niche_list,
                       spatial_info = spatial_info,
                       type = "receiver",
                       lr_network = lr_network) %>%
    mutate(scaled_receptor_score_spatial =
            scale_quantile_adapted(receptor_score_spatial))
```

```{r}
output <-
    list(DE_sender_receiver = de_sender_receiver,
         ligand_scaled_receptor_expression_fraction_df =
                ligand_scaled_receptor_expression_fraction_df,
         sender_spatial_DE_processed = sender_spatial_de_processed,
         receiver_spatial_DE_processed = receiver_spatial_de_processed,
         ligand_activities_targets = ligand_activities_targets,
         DE_receiver_processed_targets = de_receiver_processed_targets,
         exprs_tbl_ligand = exprs_tbl_ligand,
         exprs_tbl_receptor = exprs_tbl_receptor,
         exprs_tbl_target = exprs_tbl_target)

prioritization_tables <- get_prioritization_tables(output, prioritizing_weights)
qs::qsave(prioritization_tables, "Sessions/06/prioritization_tables.rds")
```

