---
title: "Coder Upgrade 2023: Cell Cell interaction analysis"
author: "Matt Cannon"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 5
        number_sections: false
        code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      cache.lazy = FALSE)
```

Load libraries
```{r libraries, cache=FALSE, warning=FALSE, error=FALSE, message=FALSE,}
library(rrrSingleCellUtils)
library(tidyverse)
library(Seurat)
library(nichenetr)
```

--------------------------------------------------------------------------------

# Section 1: Overview


## What is the goal of cell-cell interaction analysis?

## What are common tools used to do this?

CellphoneDB
Cellchat


### Nichenetr

Nichenetr relies on a pre-defined model of ligand - target interactions
It applies this model to your data to identify possible interactions

## Get their pre-defined model data
From "https://zenodo.org/record/3260758/files/ligand_target_matrix.rds" and
"https://zenodo.org/record/3260758/files/lr_network.rds"
```{r}
ligand_target_matrix <-
    readRDS("/gpfs0/home1/gdworkshop/lab/session_data/ligand_target_matrix.rds")

lr_network <-
    readRDS("/gpfs0/home1/gdworkshop/lab/session_data/lr_network.rds") %>%
    mutate(bonafide = !database %in% c("ppi_prediction",
                                        "ppi_prediction_go")) %>%
    dplyr::rename(ligand = from, receptor = to) %>%
    distinct(ligand, receptor, bonafide)
```

## Convert the species from human to mouse
```{r}
head(lr_network)
lr_network <-
    lr_network %>%
    mutate(ligand = convert_human_to_mouse_symbols(ligand),
           receptor = convert_human_to_mouse_symbols(receptor)) %>%
    drop_na()
head(lr_network)

colnames(ligand_target_matrix) <-
    ligand_target_matrix %>%
    colnames() %>%
    convert_human_to_mouse_symbols()

rownames(ligand_target_matrix) <-
    ligand_target_matrix %>%
    rownames() %>%
    convert_human_to_mouse_symbols()

ligand_target_matrix <-
    ligand_target_matrix %>%
    .[!is.na(rownames(ligand_target_matrix)),
        !is.na(colnames(ligand_target_matrix))]
```

## Get data to work with
From "https://zenodo.org/record/5840787/files/seurat_obj_subset_integrated_zonation.rds"
```{r}
seurat_obj <-
    readRDS("/gpfs0/home/gdworkshop/lab/session_data/seurat_obj_subset_integrated_zonation.rds")
```

## Define which cells we're comparing
```{r}
niche_list <- list(
    "KC_niche" = list("sender" = c("LSECs_portal",
                                   "Hepatocytes_portal",
                                   "Stellate cells_portal"),
                      "receiver" = c("KCs")),
    "MoMac2_niche" = list("sender" = c("Cholangiocytes",
                                       "Fibroblast 2"),
                          "receiver" = c("MoMac2")),
    "MoMac1_niche" = list("sender" = c("Capsule fibroblasts",
                                       "Mesothelial cells"),
                          "receiver" = c("MoMac1"))
  )
```

## Do DE to define genes of interest

### There were originally multiple SCT-transformed assays, so we need to integrate them
```{r}
assay_name <- "SCT"

seurat_obj <-
    PrepSCTFindMarkers(seurat_obj,
                       assay = assay_name,
                       verbose = FALSE)
```

### Do DE between niches, then process the results
Run DE analysis, then get rid of infinite values and calculate further statistics for DE genes

```{r}
expression_pct <- 0.10
de_sender_processed <-
    calculate_niche_de(seurat_obj = seurat_obj %>%
                            subset(features = lr_network$ligand %>%
                                intersect(rownames(seurat_obj))),
                       niches = niche_list,
                       type = "sender",
                       assay_oi = assay_name) %>%
    # Fix any infinite values by making them the max/min finite value
    mutate(avg_log2FC = ifelse(avg_log2FC == Inf,
                               max(avg_log2FC[is.finite(avg_log2FC)]),
                               ifelse(avg_log2FC == -Inf,
                                      min(avg_log2FC[is.finite(avg_log2FC)]),
                                      avg_log2FC))) %>%
    process_niche_de(niches = niche_list,
                     expression_pct = expression_pct,
                     type = "sender")

de_receiver_processed <-
    calculate_niche_de(seurat_obj = seurat_obj %>%
                            subset(features = lr_network$receptor %>%
                                   unique()),
                       niches = niche_list,
                       type = "receiver",
                       assay_oi = assay_name) %>%
    # Fix any infinite values by making them the max/min finite value
    mutate(avg_log2FC = ifelse(avg_log2FC == Inf,
                               max(avg_log2FC[is.finite(avg_log2FC)]),
                               ifelse(avg_log2FC == -Inf,
                                      min(avg_log2FC[is.finite(avg_log2FC)]),
                                      avg_log2FC))) %>%
    process_niche_de(niches = niche_list,
                     expression_pct = expression_pct,
                     type = "receiver")
```

```{r}
de_sender_receiver <-
    combine_sender_receiver_de(de_sender_processed,
                               de_receiver_processed,
                               lr_network,
                               specificity_score = "min_lfc")
```

```{r}
lfc_cutoff <- 0.15

de_receiver_processed_targets <-
    calculate_niche_de_targets(seurat_obj = seurat_obj,
                               niches = niche_list,
                               lfc_cutoff = lfc_cutoff,
                               expression_pct = expression_pct,
                               assay_oi = assay_name) %>%
    process_receiver_target_de(niches = niche_list,
                               expression_pct = expression_pct,
                               specificity_score = "min_lfc")

background <-
    de_receiver_processed_targets  %>%
    pull(target) %>%
    unique()

geneset_KC <-
    de_receiver_processed_targets %>%
    filter(receiver == niche_list$KC_niche$receiver &
            target_score >= lfc_cutoff &
            target_significant == 1 &
            target_present == 1) %>%
    pull(target) %>%
    unique()

geneset_MoMac2 <-
    de_receiver_processed_targets %>%
    filter(receiver == niche_list$MoMac2_niche$receiver &
            target_score >= lfc_cutoff &
            target_significant == 1 &
            target_present == 1) %>%
    pull(target) %>%
    unique()

geneset_MoMac1 <-
    de_receiver_processed_targets %>%
    filter(receiver == niche_list$MoMac1_niche$receiver &
            target_score >= lfc_cutoff &
            target_significant == 1 &
            target_present == 1) %>%
    pull(target) %>%
    unique()
```

```{r}
top_n_target <- 250

niche_geneset_list <- list(
    "KC_niche" = list("receiver" = "KCs",
                      "geneset" = geneset_KC,
                      "background" = background),
    "MoMac1_niche" = list("receiver" = "MoMac1",
                          "geneset" = geneset_MoMac1,
                          "background" = background),
    "MoMac2_niche" = list("receiver" = "MoMac2",
                          "geneset" = geneset_MoMac2,
                          "background" = background))

ligand_activities_targets <-
    get_ligand_activities_targets(niche_geneset_list = niche_geneset_list,
                                  ligand_target_matrix = ligand_target_matrix,
                                  top_n_target = top_n_target)

```

## Get table of normalized expression values for genes of interest

```{r}
features_oi <-
    union(lr_network$ligand,
          lr_network$receptor) %>%
    union(ligand_activities_targets$target) %>%
    setdiff(NA)

exprs_tbl <-
    suppressWarnings(Seurat::DotPlot(seurat_obj %>%
                                     subset(idents = niche_list %>%
                                            unlist() %>%
                                            unique()),
                                     features = features_oi,
                                     assay = assay_name))$data %>%
    as_tibble() %>%
    rename(celltype = id,
           gene = features.plot,
           expression = avg.exp,
           expression_scaled = avg.exp.scaled,
           fraction = pct.exp) %>%
    mutate(fraction = fraction / 100) %>%
    as_tibble() %>%
    select(celltype,
           gene,
           expression,
           expression_scaled,
           fraction) %>%
    distinct() %>%
    arrange(gene) %>%
    mutate(gene = as.character(gene))
```

```{r}
exprs_tbl_ligand <-
    exprs_tbl %>%
    filter(gene %in% lr_network$ligand) %>%
    rename(sender = celltype,
           ligand = gene,
           ligand_expression = expression,
           ligand_expression_scaled = expression_scaled,
           ligand_fraction = fraction) %>%
    mutate(scaled_ligand_expression_scaled =
                scale_quantile_adapted(ligand_expression_scaled),
           ligand_fraction_adapted = ligand_fraction) %>%
    mutate_cond(ligand_fraction >= expression_pct,
                ligand_fraction_adapted = expression_pct)  %>%
    mutate(scaled_ligand_fraction_adapted =
                scale_quantile_adapted(ligand_fraction_adapted))

exprs_tbl_receptor <-
    exprs_tbl %>%
    filter(gene %in% lr_network$receptor) %>%
    rename(receiver = celltype,
           receptor = gene,
           receptor_expression = expression,
           receptor_expression_scaled = expression_scaled,
           receptor_fraction = fraction) %>%
    mutate(scaled_receptor_expression_scaled =
                scale_quantile_adapted(receptor_expression_scaled),
           receptor_fraction_adapted = receptor_fraction) %>%
    mutate_cond(receptor_fraction >= expression_pct,
                receptor_fraction_adapted = expression_pct) %>%
    mutate(scaled_receptor_fraction_adapted =
                scale_quantile_adapted(receptor_fraction_adapted))

exprs_tbl_target <-
    exprs_tbl %>%
    filter(gene %in% ligand_activities_targets$target) %>%
    rename(receiver = celltype,
           target = gene,
           target_expression = expression,
           target_expression_scaled = expression_scaled,
           target_fraction = fraction)
```

```{r}
exprs_sender_receiver <-
    lr_network %>%
    inner_join(exprs_tbl_ligand, by = c("ligand"), multiple = "all") %>%
    inner_join(exprs_tbl_receptor, by = c("receptor"), multiple = "all") %>%
    inner_join(de_sender_receiver %>%
                    distinct(niche, sender, receiver), multiple = "all")

```


```{r}
ligand_scaled_receptor_expression_fraction_df <-
    exprs_sender_receiver %>%
    group_by(ligand, receiver) %>%
    mutate(rank_receptor_expression = dense_rank(receptor_expression),
           rank_receptor_fraction  = dense_rank(receptor_fraction)) %>%
    mutate(ligand_scaled_receptor_expression_fraction =
                0.5 * ((rank_receptor_fraction / max(rank_receptor_fraction)) +
                     ((rank_receptor_expression / max(rank_receptor_expression)))))  %>%
    distinct(ligand,
             receptor,
             receiver,
             ligand_scaled_receptor_expression_fraction,
             bonafide) %>%
    distinct() %>%
    ungroup()
```

```{r}
prioritizing_weights <-
    c("scaled_ligand_score" = 5,
      "scaled_ligand_expression_scaled" = 1,
      "ligand_fraction" = 1,
      "scaled_ligand_score_spatial" = 2,
      "scaled_receptor_score" = 0.5,
      "scaled_receptor_expression_scaled" = 0.5,
      "receptor_fraction" = 1,
      "ligand_scaled_receptor_expression_fraction" = 1,
      "scaled_receptor_score_spatial" = 0,
      "scaled_activity" = 0,
      "scaled_activity_normalized" = 1,
      "bona_fide" = 1)
```

### Make fake spatial data since we don't have it
```{r}
spatial_info <-
    tibble(celltype_region_oi = c("LSECs_portal",
                                  "Hepatocytes_portal",
                                  "Stellate cells_portal"),
           celltype_other_region = c("LSECs_central",
                                     "Hepatocytes_central",
                                     "Stellate cells_central")) %>%
  mutate(niche =  "KC_niche",
         celltype_type = "sender")

sender_spatial_de_processed <-
    get_non_spatial_de(niches = niche_list,
                       spatial_info = spatial_info,
                       type = "sender",
                       lr_network = lr_network) %>%
    mutate(scaled_ligand_score_spatial =
            scale_quantile_adapted(ligand_score_spatial))

receiver_spatial_de_processed <-
    get_non_spatial_de(niches = niche_list,
                       spatial_info = spatial_info,
                       type = "receiver",
                       lr_network = lr_network) %>%
    mutate(scaled_receptor_score_spatial =
            scale_quantile_adapted(receptor_score_spatial))
```

```{r}
output <-
    list(DE_sender_receiver = de_sender_receiver,
         ligand_scaled_receptor_expression_fraction_df =
                ligand_scaled_receptor_expression_fraction_df,
         sender_spatial_DE_processed = sender_spatial_de_processed,
         receiver_spatial_DE_processed = receiver_spatial_de_processed,
         ligand_activities_targets = ligand_activities_targets,
         DE_receiver_processed_targets = de_receiver_processed_targets,
         exprs_tbl_ligand = exprs_tbl_ligand,
         exprs_tbl_receptor = exprs_tbl_receptor,
         exprs_tbl_target = exprs_tbl_target)

prioritization_tables <- get_prioritization_tables(output, prioritizing_weights)
qs::qsave(prioritization_tables, "Sessions/06/prioritization_tables.rds")
```


# Activity (1:30 - 2ish)
Run nichenetr on os-shark? 
# Section 2: Visualizing nichenetr results


# Activity (2:30 - 3ish)
Make plots from nichenetr outputs? 

# Section 3: Interpreting nichenetr results
Activity (3:30 - 4ish)
 
# Homework
Run nichenetr on your own data or data from some random paper
Run cell-cell interaction analysis with another package
Compare the two


# Resources

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9184522/
https://www.nature.com/articles/s41592-019-0667-5
